---
title: "Visualization of PCA on clinical parameters"
subtitle: "Baseline and change-score PCA of fasting-related clinical measurements"
abstract: |
  This notebook performs two principal component analyses on clinical metadata
  from the fasting cohort. PCA 1 summarizes between-participant structure in
  complete baseline D-1 measurements across energy metabolism, liver function,
  blood-cell, kidney-function, and coagulation parameters. PCA 2 summarizes
  within-participant clinical changes by analyzing parameter deltas from D-1 to
  D+10 and from D-1 to M+1. For both analyses, the notebook generates PCA score
  plots, scree plots, and grouped loading plots; exports supplementary PCA
  figures as `figures/suppl_pca1.pdf` and `figures/suppl_pca2.pdf`; then
  combines the PCA score plots with the longitudinal weight and selected
  clinical-parameter plots to export the manuscript Figure 2 PDF.
format:
  html:
    theme: sandstone
    embed-resources: true
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, 
                      class.output="bg-light")
```

# Load 
```{r, results=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(tidyr)
library(cowplot)
library(dplyr)
library(readxl)
library(ggpubr)
library(knitr)
library(gridExtra)
library(enrichplot)
library(patchwork)
library(tidytable)
source("utils.R")
```

```{r, results=FALSE, message=FALSE, warning=FALSE}
meta <- read.csv(file="data/meta_3timepoints.csv")
meta <- meta %>% 
  filter(ID != 18) %>%
  mutate(
    ID = factor(ID),
    Timepoint = factor(Timepoint),
    Date_Scan_recode = factor(as.numeric(factor(Date_Scan))),
    sex = factor(sex)
  )
rownames(meta) <- meta$Sample
```

```{r config}
# ============================================================================ #
# Configuration: Parameter groupings for PCA analysis
# ============================================================================ #
energy_vars <- meta %>% 
  select(matches("glucose|insulin|Homa|glykohemoglobin|glyceride|cholester|hdl|ldl|Ketosemia")) %>% 
  colnames()

liver_vars <- meta %>% 
  select(matches("bilirubin|alk|ggt|got|gpt")) %>% 
  colnames()

blood_vars <- c(
  meta %>% select(matches("cytes|_abs|ZYTEN|MCH|MCV|neutrophile|baso|eosinophile|globinge|hematocrit|leukocytes")) %>% colnames(),
  "hemoglobin"
)

kidney_vars <- meta %>% 
  select(matches("uric|urea|gfr|creatinine|crp_hs")) %>% 
  colnames()

coagulation_vars <- c("inr", "ptt", "quick")

main_parameters <- c(energy_vars, liver_vars, blood_vars, kidney_vars, coagulation_vars)
```

# PCA

## PCA 1: All parameters at time point D-1

Analyze complete clinical parameters at baseline (D-1). This PCA describes
cross-sectional differences among participants before the fasting intervention,
using parameter groups for energy metabolism, liver function, blood cells, kidney
function, and coagulation. Columns with missing values are removed before PCA.

```{r}
selected_data <- meta %>% 
  select(ID, Timepoint, all_of(main_parameters)) %>% 
  filter(Timepoint == "D-1") %>%
  select(where(~ all(!is.na(.))))   # Remove columns with NA
  
```

```{r pca1-compute}
# Prepare data matrix and compute PCA (numeric parameters only)
mat <- t(selected_data %>% select(-ID, -Timepoint))
metadata <- selected_data %>% select(ID, Timepoint)
colnames(mat) <- rownames(metadata)

pca_baseline <- pca(mat, metadata = metadata, removeVar = 0)
p_scree1 <- myscreeplot(pca_baseline, components = 1:6)
p_scree1
```

```{r pca1-plot, fig.width=20, fig.height=8}
# PCA biplots (PC1 vs PC2, PC3 vs PC4)
common_theme <- theme(
  legend.position = "none",
  plot.title = element_text(hjust = 0.2,
                            size = 11,
                            margin = margin(b = -30)),
  plot.margin = margin(t = 20, r = 5, b = 0, l = 5),
  axis.text.x = element_blank(), 
  axis.text.y = element_blank(),
  axis.title = element_text(size = 9)
)

p1_12 <- biplot(pca_baseline,
  showLoadings = FALSE,
  colby = "ID",
  lab = NULL,
  pointSize = 2,
  subtitle = NULL
) +
  common_theme + 
  scale_color_manual(values = rep("grey40", 32), drop = FALSE)

p1_34 <- biplot(pca_baseline,
  x = "PC3", y = "PC4",
  showLoadings = FALSE,
  colby = "ID",
  lab = NULL,
  pointSize = 2,
  subtitle = NULL
) +
  scale_color_manual(values = rep("grey40", 32), drop = FALSE) +
  # scale_color_manual(values = patient_colors, drop = FALSE) +
  common_theme

p1_12 + p1_34 + 
  plot_layout(guides = "collect") +
  plot_annotation(title = "PCA Analysis 1: Time Point D-1") &
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, margin = margin(b = -10)),
    plot.margin = margin(t = 15, r = 5, b = -5, l = 10)
  )
```

```{r}
# Scores are samples x PCs
scores_df <- as.data.frame(pca_baseline$rotated) %>%
    tibble::rownames_to_column("sample_ID") %>%
    left_join(
        pca_baseline$metadata %>% tibble::rownames_to_column("sample_ID"),
        by = "sample_ID"
    )

# Check
stopifnot(all(c("PC1", "PC2") %in% colnames(scores_df)))

pf1 <- ggplot(scores_df, aes(x = PC1, y = PC2)) +
  geom_point(size = 2, alpha = 0.9, color ="grey40") +
  labs(
    title = "PCA Analysis 1: Time Point D-1",
    x = paste0("PC1, ", round(pca_baseline$variance[1], 1), "% variation"),
    y = paste0("PC2, ", round(pca_baseline$variance[2], 1), "% variation")
  ) +
  theme_bw() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.2, size = 10, margin = margin(b = 10)),
    plot.margin = margin(t = 20, r = 0, b = 0, l = 5),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title = element_text(size = 9),
    legend.text = element_text(size = 9),
    legend.title = element_text(size = 9)
  )

pf1
```


```{r pca1-loadings, fig.width=16, fig.height=10}
# Loadings with group background shading
df_loadings <- pca_baseline$loadings[, 1:6] %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Factors") %>%
  pivot_longer(cols = -Factors, names_to = "PC", values_to = "Loadings") %>%
  mutate(
    Factors = factor(Factors, levels = main_parameters),
    Group = case_when(
      Factors %in% energy_vars      ~ "Energy metabolism",
      Factors %in% liver_vars       ~ "Liver function",
      Factors %in% blood_vars       ~ "Blood cells",
      Factors %in% kidney_vars      ~ "Kidney function",
      Factors %in% coagulation_vars ~ "Coagulation",
      TRUE                          ~ "Other"
    ),
    Group = factor(
      Group,
      levels = c(
        "Kidney function", "Blood cells", "Liver function",
        "Energy metabolism", "Coagulation", "Other"
      )
    )
  )

# Order factors by group for consistent background blocks
factor_levels <- df_loadings %>%
  distinct(Factors, Group) %>%
  arrange(Group, Factors) %>%
  pull(Factors) %>%
  as.character()

df_loadings <- df_loadings %>%
  mutate(Factors = factor(Factors, levels = factor_levels))

# Background rectangles per group
background_df <- df_loadings %>%
  select(Factors, Group) %>%
  distinct() %>%
  mutate(
    ymin = as.numeric(Factors) - 0.5,
    ymax = as.numeric(Factors) + 0.5,
    xmin = -Inf,
    xmax = Inf
  )

pf1_loadings <- ggplot(df_loadings, aes(x = Loadings, y = Factors, color = Loadings)) +
  geom_rect(
    data = background_df,
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, fill = Group),
    inherit.aes = FALSE,
    alpha = 0.1
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black", linewidth = 0.8) +
  geom_point(size = 2) +
  scale_color_gradientn(
    colors = c("red", "blue", "red"),
    values = c(0, 0.5, 1),
    limits = c(-1, 1)
  ) +
  facet_grid(. ~ PC, scales = "free") +
  theme_bw(base_size = 14) +
  xlim(-1, 1) +
 theme(
    legend.position = "right",
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(size = 8),
    axis.title.x = element_text(size = 12),
    legend.title = element_text(size = 12),
    plot.margin = margin(t = -10, r = 0, b = 10, l = 0)
  ) +
  labs(
    x = "PCA Loadings",
    y = "",
    color = "Loadings",
    fill = "Parameters",
    title = ""
  ) +
  scale_y_discrete(labels = var_to_full)

pf1_loadings
```

```{r}
grDevices::cairo_pdf("figures/suppl_pca1.pdf", width=11, height=12)
plot_grid(p_scree1 + labs(title="") + theme(plot.margin=margin(t=5, r=85, b=15, l=85)),
          pf1_loadings, 
          labels=c("a", "b"), label_size=20, ncol=1, rel_heights=c(1,1.2), align="v", axis="tb")
dev.off()
```

## PCA 2: Changes compared to D-1

Analyze within-participant parameter changes relative to baseline. For each
clinical parameter, the notebook computes two delta values per participant:
`D+10 - D-1` and `M+1 - D-1`. The PCA is then run on these change scores to
summarize coordinated clinical responses during the fasting endpoint and
one-month follow-up.

```{r pca2-prepare}
# Get parameter columns
selected_data <- meta %>% 
  select(ID, Timepoint, all_of(main_parameters))

param_names <- setdiff(names(selected_data), c("ID", "Timepoint"))

# Pivot wider to get one row per patient
data_wide <- selected_data %>%
  pivot_wider(
    names_from = Timepoint,
    values_from = all_of(param_names),
    names_sep = "_"
  )

# Compute deltas for each parameter explicitly
after_diff <- data_wide %>%
  select(ID) %>%
  bind_cols(
    map_dfc(param_names, function(param) {
      baseline_col <- paste0(param, "_D-1")
      after_col <- paste0(param, "_D+10")
      tibble(!!param := data_wide[[after_col]] - data_wide[[baseline_col]])
    })
  ) %>%
  mutate(Time = "ΔD+10")

fup1_diff <- data_wide %>%
  select(ID) %>%
  bind_cols(
    map_dfc(param_names, function(param) {
      baseline_col <- paste0(param, "_D-1")
      fup_col <- paste0(param, "_M+1")
      tibble(!!param := data_wide[[fup_col]] - data_wide[[baseline_col]])
    })
  ) %>%
  mutate(Time = "ΔM+1")

# Combine and prepare for PCA
delta_data <- bind_rows(after_diff, fup1_diff) %>%
  select(where(~ all(!is.na(.))))   # Remove columns with NA
mat <- t(delta_data %>% select(-ID, -Time))
metadata <- delta_data %>% 
  select(ID, Time) %>%
  mutate(Time = factor(Time, levels = c("ΔD+10", "ΔM+1")))
colnames(mat) <- rownames(metadata)

pca_delta <- pca(mat, metadata = metadata, removeVar = 0)
p_scree2 <- myscreeplot(pca_delta, components = 1:6)
p_scree2
```

```{r pca2-plot, fig.width=20, fig.height=10}
# Scores are samples x PCs
scores_df <- as.data.frame(pca_delta$rotated) %>%
    tibble::rownames_to_column("sample_ID") %>%
    left_join(
        pca_delta$metadata %>% tibble::rownames_to_column("sample_ID"),
        by = "sample_ID"
    )

# Check
stopifnot(all(c("PC1", "PC2", "Time") %in% colnames(scores_df)))

pf2 <- ggplot(scores_df, aes(x = PC1, y = PC2, color = Time)) +
  geom_point(size = 2, alpha = 0.9) +
  scale_color_manual(values = c("ΔD+10" = "#4169E1", "ΔM+1" = "#E35B53")) +
  labs(
    title = "PCA Analysis 2: Δ[D+10, D-1] and Δ[M+1, D-1]",
    x = paste0("PC1, ", round(pca_delta$variance[1], 1), "% variation"),
    y = paste0("PC2, ", round(pca_delta$variance[2], 1), "% variation")
  ) + 
  theme_bw() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.2, size = 10, margin = margin(b = 10)),
    plot.margin = margin(t = 20, r = 0, b = 0, l = 5),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title = element_text(size = 9),
    legend.text = element_text(size = 9),
    legend.title = element_text(size = 9)
  )

pf2
```


```{r pca2-loadings, fig.width=16, fig.height=10}
# Prepare loadings data with parameter groupings
df_loadings <- as.data.frame(pca_delta$loadings) %>%
  tibble::rownames_to_column("Factors") %>%
  select(Factors, tidyselect::any_of(paste0("PC", 1:6))) %>%
  pivot_longer(
    cols = starts_with("PC"),
    names_to = "PC",
    values_to = "Loadings"
  ) %>%
  mutate(
    Factors = factor(Factors, levels = main_parameters),
    Group = case_when(
      Factors %in% energy_vars       ~ "Energy metabolism",
      Factors %in% liver_vars        ~ "Liver function",
      Factors %in% blood_vars        ~ "Blood cells",
      Factors %in% kidney_vars       ~ "Kidney function",
      Factors %in% coagulation_vars  ~ "Coagulation",
      TRUE                           ~ "Other"
    ),
    Group = factor(
      Group,
      levels = c("Kidney function", "Blood cells", "Liver function",
                 "Energy metabolism", "Coagulation", "Other")
    )
  ) %>%
  arrange(Group, Factors, PC)

factor_levels <- df_loadings %>%
  distinct(Factors, Group) %>%
  mutate(
    Factors = factor(Factors, levels = main_parameters),
    Group = factor(Group, levels = c(
      "Kidney function", "Blood cells", "Liver function",
      "Energy metabolism", "Coagulation", "Other"
    ))
  ) %>%
  arrange(Group, Factors) %>%
  pull(Factors) %>%
  as.character()

df_loadings <- df_loadings %>%
  mutate(Factors = factor(Factors, levels = factor_levels))

# Background rectangles for groups
background_df <- df_loadings %>%
  select(Factors, Group) %>%
  distinct() %>%
  mutate(
    ymin = as.numeric(Factors) - 0.5,
    ymax = as.numeric(Factors) + 0.5,
    xmin = -Inf,
    xmax = Inf
  )

# Add variance explained to PC labels
pc_labels <- paste0("PC", 1:6, ": ", round(pca_delta$variance[1:6], 1), "%")
df_loadings <- df_loadings %>%
  mutate(PC_labels = pc_labels[as.numeric(sub("PC", "", PC))])

# Plot loadings with group backgrounds
pf3 <- ggplot(df_loadings, aes(x = Loadings, y = Factors, color = Loadings)) +
  geom_rect(
    data = background_df,
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, fill = Group),
    inherit.aes = FALSE, alpha = 0.1
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black", linewidth = 0.5) +
  geom_point(size = 2) +
  scale_color_gradientn(
    colors = c("red", "blue", "red"),
    values = c(0, 0.5, 1),
    limits = c(-1, 1)
  ) +
  facet_grid(. ~ PC_labels, scales = "free") +
  theme_bw(base_size = 14) +
  xlim(-1, 1) +
  theme(
    legend.position = "right",
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(size = 8),
    axis.title.x = element_text(size = 12),
    legend.title = element_text(size = 12),
    plot.margin = margin(t = -10, r = 0, b = 10, l = 0)
  ) +
  labs(x = "PCA Loadings", y = "", color = "Loadings", title = "", fill = "Parameters") +
  scale_y_discrete(labels = var_to_full)

pf3
```

```{r}
grDevices::cairo_pdf("figures/suppl_pca2.pdf", width=11, height=12)
plot_grid(p_scree2 + labs(title="") + theme(plot.margin=margin(t=5, r=85, b=15, l=85)),
          pf3, 
          labels=c("a", "b"), label_size=20, ncol=1, rel_heights=c(1,1.2), align="v", axis="tb")
dev.off()
```

# Figure 2
```{r}
load("figures/p_weight_trends.RData")
load("figures/p_blood_params_selected.RData")
```

```{r}
p_up <- plot_grid(p_weight_trend, pf1, pf2, labels=c("a", "b", "c"), label_size=20, ncol=3, rel_widths=c(1.5,1,1.3), align="h", axis="tb")
p_fig2 <- plot_grid(p_up, p_selected_box, labels=c("", "d"), label_size=20, ncol=1, rel_heights=c(1,1.2), align="v", axis="tb")

grDevices::cairo_pdf("figures/Fig2.pdf", width=12, height=8)
print(p_fig2)
dev.off()
```
