---
title: "Visualization of clinical parameters"
subtitle: "Longitudinal summaries of body weight, BMI, and fasting-related clinical measurements"
abstract: |
  This notebook creates separate descriptive and longitudinal summaries of
  clinical measurements in the fasting cohort. The baseline BMI plot is a
  cross-sectional visualization of D-1 BMI by age, sex, and BMI category. The
  longitudinal weight plot uses a linear mixed model with a spline term for
  time and participant-specific random intercepts to estimate the population
  weight trajectory across fasting and food reintroduction. Selected metabolic, 
  renal, inflammatory, coagulation, vital-sign, and blood-cell parameters are then
  summarized at D-1, D+3, D+10, and M+1; for each parameter, a separate linear
  mixed model compares other three timepoints with baseline and exports adjusted
  p-values, and mean/SD summary tables. A specified subset of these variables 
  are used to create a boxplot, and figure-ready RData objects are exported.
format:
  html:
    theme: sandstone
    embed-resources: true
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r}
library(openxlsx)
library(readr)
library(dplyr)
library(tidyr)
library(stringr)
library(purrr)
library(lme4)
library(lmerTest) 
library(emmeans)
library(rstatix)
library(ggplot2)
library(ggeffects)
library(ggthemes)
library(splines)
library(Polychrome)
library(ggpubr)
library(patchwork)
source("utils.R")

daily_meta <- read_csv("data/meta_daily.csv")
meta3d <- read_csv("data/meta_3timepoints.csv")

labels_daily <- c("D-2","D-1", "D0", paste0("D+", 1:16), "M+1")
labels_4timepoint <- c("D-1", "D+4", "D+10", "M+1")
labels_3timepoint <- c("D-1", "D+10", "M+1")
```

# Weight & BMI

This section contains two different visual summaries:

* The longitudinal weight plot models repeated weight measurements across the
  fasting and food reintroduction period. The fitted model is
  `weight ~ ns(numeric_timepoint, df = 3) + (1 | ID)`, where the natural spline
  captures the nonlinear time trend and `(1 | ID)` accounts for repeated
  measurements from the same participant.
* The baseline BMI plot is descriptive only. It shows each participant's D-1 BMI
  against age, with background bands indicating BMI categories and point shape
  indicating sex. It is not based on the longitudinal LMM.

```{r}
df <- daily_meta %>%
  select(weight, Timepoint, ID, sex, age, BMI) %>%
  filter(Timepoint != "D+17") %>% # because nothing is measured
  filter(ID!=18) %>%
  mutate(ID = factor(ID, levels = unique(ID))) %>% 
  mutate(BMI = as.numeric(BMI)) %>%
  mutate(sex = factor(c("male","female")[sex + 1])) %>%
  mutate(Timepoint = factor(Timepoint, levels = labels_daily)) %>%
  mutate(numeric_timepoint = c(1:19, 25)[match(Timepoint, levels(Timepoint))]) %>%
  mutate(day = numeric_timepoint - 3)
```

```{r}
# Prepare data for modeling
model <- lmer(weight ~ ns(numeric_timepoint, df=3) + (1 | ID), data = df)

# Extract marginal predictions (population-level)
pred <- ggpredict(model, terms = "numeric_timepoint [all]")
pred <- pred %>% filter(x != 1)

# Prepare data frame for plotting
df_plot <- df %>%
  filter(Timepoint != "D-2")

# Build phase shading boundaries
phase_df <- bind_rows(
  tibble(
    phase = "Fasting",
    day_min = 1, day_max = 12
  ),
  tibble(
    phase = "Food reintroduction",
    day_min = 13, day_max = 16
  )
) %>%
  left_join(
    df_plot %>% distinct(day, numeric_timepoint),
    by = c("day_min" = "day")
  ) %>%
  rename(xmin = numeric_timepoint) %>%
  left_join(
    df_plot %>% distinct(day, numeric_timepoint),
    by = c("day_max" = "day")
  ) %>%
  rename(xmax = numeric_timepoint) %>%
  select(phase, xmin, xmax) %>%
  mutate(
    xmin = xmin - 0.5,
    xmax = xmax + 0.5
  )

phase_labels <- phase_df %>%
  mutate(
    x = (xmin + xmax) / 2,
    phase_top = if_else(phase == "Food reintroduction", "Food", phase),
    phase_bottom = if_else(phase == "Food reintroduction", "reintroduction", NA_character_)
  )

# Create plot
p_weight_trend <-
  ggplot(df_plot, aes(x = numeric_timepoint, y = weight)) +
  # Phase shading
  geom_rect(
    data = phase_df,
    aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf, fill = phase),
    inherit.aes = FALSE,
    alpha = 0.2
  ) +
  scale_fill_manual(values = c(Fasting = "grey20", "Food reintroduction" = "grey60"), guide = "none") +
  annotate(
    "text",
    x = phase_labels$x,
    y = Inf,
    label = phase_labels$phase_top,
    vjust = 1.4,
    size = 4,
    color = "black"
  ) +
  annotate(
    "text",
    x = phase_labels$x,
    y = Inf,
    label = phase_labels$phase_bottom,
    vjust = 2.7,
    size = 4,
    color = "black"
  ) +
  # Individual trends (de-emphasized)
  geom_line(aes(group = ID), color = "#6B7C93", alpha = 0.5, linewidth = 0.5) +
  geom_point(aes(group = ID), color = "#6B7C93", alpha = 0.55, size = 0.1) +
  # Mean trend + 95% CI
  geom_ribbon(
    data = pred,
    aes(x = x, ymin = conf.low, ymax = conf.high),
    inherit.aes = FALSE,
    fill = "#0033FF",
    alpha = 0.25
  ) +
  geom_line(
    data = pred,
    aes(x = x, y = predicted),
    inherit.aes = FALSE,
    color = "#0033FF",
    linewidth = 1.2
  ) +
  scale_x_continuous(
    breaks = unique(df_plot$numeric_timepoint),
    labels = labels_daily[2:20]
  ) +
  labs(x = "Time Point", y = "Weight (kg)") +
  theme_minimal() +
  theme(
    legend.position = "none",
    panel.grid.minor.x = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.margin = margin(t = 20, l = 10),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  ) 

data_p_weight_trend <- df_plot
save(p_weight_trend, data_p_weight_trend, file = "figures/p_weight_trends.RData")
```

```{r}
tmp <- meta3d[,c("ID","Timepoint","age","Gender","BMI","BMI_baseline_category")] %>% 
  mutate(Gender=factor(Gender)) %>%
  mutate(ID=factor(ID)) %>%
  filter(Timepoint=="D-1") %>% 
  drop_na(Gender)
tmp$BMI_baseline_category <- factor(tmp$BMI_baseline_category, levels = c("Underweight", "Healthy", "Overweight", "Obesity", "Severe Obesity"))
table(tmp$BMI_baseline_category)

# Define the background shading
bmi_ranges <- data.frame(
  ymin = c(15, 18.5, 25, 30, 40),
  ymax = c(18.5, 25, 30, 40, 45),
  category = c("Underweight", "Healthy", "Overweight", "Obesity", "Severe Obesity")
)
bmi_ranges$category <- factor(bmi_ranges$category, levels = c("Underweight", "Healthy", "Overweight", "Obesity", "Severe Obesity"))


p_bmi_category <- ggplot() +
  geom_rect(data = bmi_ranges, aes(ymin = ymin, ymax = ymax, xmin = -Inf, xmax = Inf, fill = category), alpha = 0.3) +
  scale_fill_manual(values = c("Underweight" = "lightblue",
                               "Healthy" = "lightgreen",
                               "Overweight" = "yellow",
                               "Obesity" = "orange",
                               "Severe Obesity" = "red")) +
  geom_jitter(data=tmp, aes(x = age, y = BMI, color=ID, shape=Gender), size=4, height=0, alpha=0.8) +
  scale_color_manual(values = patient_colors, drop = TRUE) +
  labs(x = "Age", y = "BMI at Baseline", color="Patient ID", fill="BMI Categories", shape="Sex") +
  theme_minimal() + 
 theme(
    legend.position = "right",
    legend.box = "horizontal",     # Arrange legends side by side (columns)
    legend.box.just = "left"       # Optional: aligns legends
  ) +
    guides(
    # color = "none",
    fill = guide_legend(ncol = 1, order = 1),  # Stack fill above shape
    shape = guide_legend(ncol = 1, order = 2)  # Make sure shape is after fill
  ) +
  ylim(15, 45) 

p_bmi_category

save(p_bmi_category, file = "figures/p_bmi_category.RData")
```

```{r}
pdf("figures/suppl_bmi_category.pdf", width=9, height=5)
print(p_bmi_category)
dev.off()
```

## Blood parameters

This section evaluates selected longitudinal clinical parameters at D-1, D+3,
D+10, and M+1. For each parameter, the model is fitted separately as
`Value ~ Timepoint + (1 | ID)`. `Timepoint` estimates fixed differences from the
baseline reference level (`D-1`), and `(1 | ID)` accounts for within-participant
correlation across repeated measurements. Estimated marginal means are used to
test each follow-up timepoint against D-1, with Bonferroni adjustment across the
three baseline contrasts for that parameter.

```{r}
# ============================================================================ #
# COLLEAGUE INPUT: Parameter lists for visualization
# Update these when your colleague provides the final selections
# ============================================================================ #

# List 1: All parameters to display in comprehensive analysis
bp_vars <- daily_meta %>% select(matches("pulse|dbp|sbp")) %>% colnames()
energy_vars <- daily_meta %>% select(matches("glucose|insulin|Homa|glykohemoglobin|glyceride|cholester|hdl|ldl|Ketosemia")) %>% colnames()
liver_vars <- daily_meta %>% select(matches("bilirubin|alk|ggt|got|gpt")) %>% colnames()
blood_vars <- c(daily_meta %>% select(matches("cytes|_abs|ZYTEN|MCH|MCV|neutrophile|baso|eosinophile|globinge|hematocrit|leukocytes")) %>% colnames(), "hemoglobin")
kidney_vars <- daily_meta %>% select(matches("uric|urea|gfr|creatinine|crp_hs")) %>% colnames()
coagulation_vars <- c("inr", "ptt", "quick")

main_parameters <- c(bp_vars, energy_vars, liver_vars, blood_vars, kidney_vars, coagulation_vars)
vital_parameters <- c("BMI", "weight", "sbp", "dbp", "pulse")

custom_labels <- c(
  Ketosemia = "Capillary blood ketone (mmol/L)",
  # Homa = "HOMA-IR",
  # glykohemoglobin = "Glycohemoglobin (mmol/mol)",
  triglyceride = "Triglycerides (mg/dL)",
  ldl_hdl_ratio = "LDL-to-HDL cholesterol Ratio",
  # `gpt/ALT` = "GPT/ALT (U/L)",
  # Bilirubin = "Bilirubin (mg/dL)",
  # alk_phosphatase = "Alkaline Phosphatase (IU/L)",
  # thrombocytes = "Thrombocytes (xE9/L)",
  # NLR = "Neutrophil-to-Lymphocyte Ratio",
  inr = "International normalized ratio (INR)",
  # hdl = "HDL (mg/dL)",
  # ldl = "LDL (mg/dL)",
  # NonHDL.Cholesterin = "Non-HDL Cholesterol (mg/dL)",
  # cholesterol = "Total Cholesterol (mg/dL)",
  # `got/AST` = "GOT/AST (U/L)",
  # ggt = "GGT (U/L)",
  # NEUTROPHILE = "Neutrophils (xE9/L)",
  # LYMPHOZYTEN = "Lymphocytes (xE9/L)",
  # ptt = "PTT (s)",
  # quick = "Quick",
  glucose = "Venous glucose (mg/dL)",
  uric_acid = "Uric acid (mg/dL)",
  leukocytes = "White blood cell count (Tsd./µL)",
  crp_hs = "C-reactive protein (hsCRP) (mg/L)"
)

# List 2: Selected parameters for main figure
selected_parameters <- c("Ketosemia", "glucose", "triglyceride", "ldl_hdl_ratio", 
                         "uric_acid", "leukocytes", "crp_hs", "inr")

# Parameters that should NOT include D-1 timepoint (none now)
parameters_exclude_diet <- character(0)
```

```{r}
# ============================================================================ #
# Data preparation
# ============================================================================ #
df <- daily_meta %>%
  select(Timepoint, ID, all_of(main_parameters), all_of(vital_parameters)) %>%
  mutate(NLR = NEUTROPHILE / LYMPHOZYTEN) %>%
  filter(Timepoint %in% c("D-1","D+3", "D+10", "M+1")) %>%
  mutate(Timepoint = factor(Timepoint, levels = c("D-1","D+3", "D+10", "M+1"))) %>%
  mutate(ID = factor(ID, levels = unique(ID))) 

df_long <- df %>% 
  pivot_longer(
    cols = -c(Timepoint, ID),
    names_to = "Parameter",
    values_to = "Value"
  )
```

```{r}
# ============================================================================ #
# LMM-based p-values for all parameters                   
# ============================================================================ #
ref_tp <- "D-1"
tp_levels <- c("D-1", "D+3", "D+10", "M+1")

df_long2 <- df_long %>%
  mutate(
    ID = factor(ID),
    Timepoint = factor(Timepoint, levels = tp_levels)
  )

fit_one_param <- function(dat, ref = ref_tp) {
  dat <- dat %>% filter(!is.na(Value), !is.na(Timepoint), !is.na(ID))
  if (n_distinct(dat$Timepoint) < 2L) return(NULL)
  m <- lmer(Value ~ Timepoint + (1 | ID), data = dat, REML = TRUE)
  em <- emmeans(m, ~ Timepoint)
  ct <- contrast(em, method = "trt.vs.ctrl", ref = which(levels(dat$Timepoint) == ref))
  out <- as.data.frame(summary(ct, infer = c(TRUE, TRUE), adjust = "bonferroni")) %>%
    rename(
      contrast = contrast,
      estimate = estimate,
      SE = SE,
      df = df,
      t = t.ratio,
      p_adj = p.value,
      conf.low = lower.CL,
      conf.high = upper.CL
    )
  out
}

pval_lmm <- df_long2 %>%
  group_by(Parameter) %>%
  group_modify(~ {
    res <- fit_one_param(.x, ref = ref_tp)
    if (is.null(res)) return(tibble())
    res
  }) %>%
  ungroup()
```

```{r}
# ============================================================================ #
# Prepare data for visualization with p-values          
# ============================================================================ #
# (A) Parse contrasts into group1/group2
pval_lmm_plot <- pval_lmm %>%
  mutate(
    group2 = gsub("[()]", "", str_trim(str_split_fixed(contrast, " - ", 2)[, 1])),
    group1 = gsub("[()]", "", str_trim(str_split_fixed(contrast, " - ", 2)[, 2]))
  )

step_increase <- 0.1
ymax_df <- df_long2 %>%
  group_by(Parameter) %>%
  summarise(ymax = max(Value, na.rm = TRUE), .groups = "drop") %>%
  mutate(step = step_increase * (abs(ymax) + 1))  # scale step to data magnitude

pval_lmm_plot <- pval_lmm_plot %>%
  left_join(ymax_df, by = "Parameter") %>%
  group_by(Parameter) %>%
  arrange(group2, .by_group = TRUE) %>%
  mutate(
    y.position = ymax + row_number() * step
  ) %>%
  ungroup()

pval_lmm_plot <- pval_lmm_plot %>%
  mutate(
    p.adj = p_adj,
    p.adj.signif = case_when(
      is.na(p.adj)        ~ NA_character_,
      p.adj <= 0.0001     ~ "****",
      p.adj <= 0.001      ~ "***",
      p.adj <= 0.01       ~ "**",
      p.adj <= 0.05       ~ "*",
      TRUE                ~ "ns"
    )
  )

# Data for selected parameters visualization
tmp_data <- df_long2 %>%
  filter(Parameter %in% selected_parameters) %>%
  mutate(Parameter = factor(Parameter, levels = selected_parameters))

tmp_pval <- pval_lmm_plot %>%
  filter(Parameter %in% selected_parameters) %>%
  mutate(Parameter = factor(Parameter, levels = selected_parameters))
```

```{r}
# Blood parameters visualization (selected subset)
p_selected_box <- tmp_data %>%
  ggplot(aes(x = Timepoint, y = Value)) +
  geom_point(size=1, color = "grey90") +
  facet_wrap(~ Parameter, scales = "free", ncol = 4, labeller = labeller(Parameter = custom_labels)) +
  geom_boxplot(width = 0.3, outlier.shape = NA, fill = "grey90") +
  # geom_point(size = 1, aes(color = ID)) +
  geom_line(alpha = 0.5, linewidth = 0.1, aes(group = ID), linetype = "dashed") +
  scale_color_manual(values = patient_colors, drop = FALSE) +
  scale_x_discrete(labels = labels_4timepoint) +
  labs(x = "Time Point", y = "", color = "Patient ID") +
  theme_minimal() +
  theme(
    legend.position = "right",
    panel.grid.major = element_blank(),
    panel.grid.minor.x = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  stat_pvalue_manual(
    tmp_pval,
    label = "p.adj.signif",
    tip.length = 0.01,
    size = 3,
    bracket.size = 0.3,
    hide.ns = TRUE
  )

save(p_selected_box, tmp_data, tmp_pval, file = "figures/p_blood_params_selected.RData")
```

```{r}
# Summaries and export using unified LMM results
df_summary <- df_long2 %>%
  group_by(Timepoint, Parameter) %>%
  summarise(
    mean = mean(Value, na.rm = TRUE),
    sd   = sd(Value, na.rm = TRUE),
    .groups = "drop"
  )

df_output <- pval_lmm_plot %>% select(-ymax, -step, -y.position, -p.adj)
```

```{r}
write.xlsx(df_summary %>% filter(Parameter %in% vital_parameters),
           file = "figures/mean_sd_vital_parameters.xlsx")

write.xlsx(df_output %>% filter(Parameter %in% vital_parameters),
           file = "figures/pval_vital_parameters.xlsx")

write.xlsx(df_output, file = "figures/pval_clinical_parameters.xlsx")
write.xlsx(df_summary, file = "figures/mean_sd_clinical_parameters.xlsx")
```
