---
title: "Methylation data exploration"
subtitle: "QC, normalization, and covariate assessment for EPIC beta values"
abstract: |
  This notebook explores the initial Sesame-derived methylation beta matrix and
  sample metadata before downstream differential methylation analysis. It loads
  EPIC probe annotations and the IDAT sample sheet, flags the previously
  identified low-quality sample, removes sex-chromosome probes, and creates two
  BMIQ-normalized autosomal beta matrices: a sample-prioritized matrix that
  keeps all samples and drops probes with missing values, and a probe-prioritized
  matrix that excludes the low-quality sample before dropping probes with missing
  values. The notebook uses ChAMP QC plots, MDS plots, SVD-based covariate
  assessment, and alluvial batch-structure visualizations to evaluate effects of
  sex, scan date, array barcode, array position, plate, and timepoint. It then
  applies ComBat scan-date correction while preserving Timepoint and saves four
  beta-matrix versions for downstream analysis: `cleaned_beta_s.RDS` and
  `cleaned_beta_s_corrected.RDS` for the sample-prioritized matrix without and
  with batch correction, and `cleaned_beta_p.RDS` and
  `cleaned_beta_p_corrected.RDS` for the probe-prioritized matrix without and
  with batch correction.
author: "Siyuan Luo"
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
format:
  html:
    theme: sandstone
    embed-resources: true
    toc: true
editor_options: 
  chunk_output_type: console
---

# Load
```{r, results=FALSE, message=FALSE, warning=FALSE}
proj_dir <- "/home/siluo/public/SiyuanLuo/projects/fasting_clinics/"
setwd(proj_dir)
```

```{r, results=FALSE, message=FALSE, warning=FALSE}
library(readxl)
library(ChAMP)
library(readr)
```

Probe information
```{r, output=FALSE}
df_epic <- read_csv("anno/infinium-methylationepic-v-1-0-b5-manifest-file.csv", skip = 7)
unique(df_epic$CHR_hg38)
df_epic <- df_epic[!is.na(df_epic$CHR_hg38), ]
autosome_prb <-df_epic[!df_epic$CHR_hg38 %in% c("chrX", "chrY"), ]$IlmnID
```

Samplesheet

```{r, warning=FALSE}
samplesheet <- read_excel("IDATs_sorted_Genesis/14.12.2022_sample_sheet_R&D_Buchinger_ALL.xlsx", sheet=1)
samplesheet$Sample <- samplesheet$Basename
samplesheet[] <- lapply(samplesheet, as.factor)
pheno <- samplesheet[,c("Sample","Sample_Name", "Sample_Plate", "ID Patient", "AMP_Plate", "SentrixBarcode_A", "SentrixPosition_A", "Date_Scan", "Gender", "Timepoint")]

betas <- readRDS("betas.RDS")

rownames(pheno) <- pheno$Sample
pheno <- as.data.frame(pheno[colnames(betas),])
names(pheno)[names(pheno) == "ID Patient"] <- "ID_Patient"

```

Add previous outlier detection results
```{r}
pheno$outliers <- FALSE
pheno[pheno$Sample_Name == "4033964733", "outliers"] <- TRUE
```

```{r}
barplot(colSums(is.na(betas)))
```

The outlier `4033964733` has much more `NA` than others.

```{r}
# Exclude probes on X, Y chromosomes:
beta_auto <- betas[intersect(autosome_prb, rownames(betas)),]
cleaned_beta_s <- na.omit(beta_auto) # prioritize samples
```

```{r}
cleaned_beta_p <- na.omit(beta_auto[, !colnames(beta_auto) %in% "206467110003_R08C01"]) # prioritize probes
```


Meta data
```{r}
blmeta <- read_excel("metadata/GENESIS_DATA_19.05.2022_blooddata.xlsx", sheet = 1)
m <- blmeta[,c("ID", "sex", "age")]
m$ID_Patient <- paste0("B__", sapply(m$ID, function(x){sprintf("%03d", x)}))
m <- unique(m)
pheno <- merge(pheno, m, by="ID_Patient")
rownames(pheno) <- pheno$Sample
pheno <- as.data.frame(pheno[colnames(betas),])
table(pheno$Gender, pheno$sex)
```

# Exploration
The procedures include:

- ChAMP's QC check
- ChAMP's SVD to check for covariates 

## QC plots

```{r}
# champ.QC(beta = cleaned_beta_s, 
#        pheno = pheno$Date_Scan,
#        mdsPlot=TRUE,
#        densityPlot=TRUE,
#        dendrogram=TRUE,
#        PDFplot=TRUE,
#        Rplot=TRUE,
#        Feature.sel="SVD",
#        resultsDir="./CHAMP_QCimages/")

# QC.GUI(beta = cleaned_beta_s, 
#        pheno = pheno$Gender,
#        arraytype = "EPIC")
```


```{r, echo=FALSE, include=FALSE, eval=FALSE}
mdsPlot(cleaned_beta, numPositions=1000,
        sampNames = NULL, sampGroups=pheno$Gender)

# Two major splits are gender. 
```


```{r}
#| layout-ncol: 2
mdsPlot(cleaned_beta_s, numPositions=1000,
        sampNames = NULL, sampGroups=pheno$Gender)
mdsPlot(cleaned_beta_s, numPositions=1000,
        sampNames = NULL, sampGroups=pheno$Date_Scan)
mdsPlot(cleaned_beta_s, numPositions=1000,
        sampNames = NULL, sampGroups=pheno$Timepoint)
mdsPlot(cleaned_beta_s, numPositions=1000,
        sampNames = NULL, sampGroups=pheno$SentrixPosition_A)
```


## BMIQ Normlization

```{r}
#| layout-ncol: 2

densityPlot(cleaned_beta_s, sampGroups=pheno$Date_Scan,
            main=paste0("Density plot of beta values before BMIQ normalization"),
            xlab="Beta", legend=FALSE)

cleaned_beta_s <- champ.norm(beta=cleaned_beta_s,
                                  rgSet=NULL,
                                  mset=NULL,
                                  resultsDir="./CHAMP_Normalization/",
                                  method="BMIQ",
                                  plotBMIQ=TRUE,
                                  arraytype="EPIC",
                                  cores=10)

densityPlot(cleaned_beta_s, sampGroups=pheno$Date_Scan,
            main=paste0("Density plot of beta values after BMIQ normalization"),
            xlab="Beta", legend=FALSE)

cleaned_beta_p <- champ.norm(beta=cleaned_beta_p,
                                  rgSet=NULL,
                                  mset=NULL,
                                  resultsDir="./CHAMP_Normalization/",
                                  method="BMIQ",
                                  plotBMIQ=TRUE,
                                  arraytype="EPIC",
                                  cores=10)
```

"Date_Scan" indicates some batch effect.

## Covariate analysis

```{r, echo=FALSE}
library(ggplot2)
champ.SVD <- function(beta=myNorm,
                      rgSet=NULL,
                      pd=myLoad$pd,
                      RGEffect=FALSE,
                      PDFplot=TRUE,
                      Rplot=TRUE,
                      resultsDir="./CHAMP_SVDimages/")
{
    message("[===========================]")
    message("[<<<<< ChAMP.SVD START >>>>>]")
    message("-----------------------------")

    ### Defind some functions to be used in champ.SVD.
   GenPlot <- function(thdens.o,estdens.o,evalues.v){
       minx <- min(min(thdens.o$lambda),min(evalues.v));
       maxx <- max(max(thdens.o$lambda),max(evalues.v));
       miny <- min(min(thdens.o$dens),min(estdens.o$y));
       maxy <- max(max(thdens.o$dens),max(estdens.o$y));
   }
   # Estimate how many latent variable hidden in the data set.
    EstDimRMTv2 <- function(data.m)
    {    
        ### standardise matrix
        M <- data.m;
        for(c in 1:ncol(M)) M[,c] <- (data.m[,c]-mean(data.m[,c]))/sqrt(var(data.m[,c]));
        sigma2 <- var(as.vector(M));
        Q <- nrow(data.m)/ncol(data.m);
        thdens.o <- thdens(Q,sigma2,ncol(data.m));
        C <- 1/nrow(M) * t(M) %*% M;

        eigen.o <- eigen(C,symmetric=TRUE);
        estdens.o <- density(eigen.o$values,from=min(eigen.o$values),to=max(eigen.o$values),cut=0);

        GenPlot(thdens.o,estdens.o,eigen.o$values);
        intdim <- length(which(eigen.o$values > thdens.o$max));
        return(list(cor=C,dim=intdim,estdens=estdens.o,thdens=thdens.o));
    }
    thdens <- function(Q,sigma2,ns)
    {
        lambdaMAX <- sigma2*(1+1/Q + 2*sqrt(1/Q));
        lambdaMIN <- sigma2*(1+1/Q - 2*sqrt(1/Q));
        delta <- lambdaMAX - lambdaMIN;#  print(delta);
        roundN <- 3;
        step <- round(delta/ns,roundN);
        while(step==0){
            roundN <- roundN+1;
            step <- round(delta/ns,roundN);
        }
        lambda.v <- seq(lambdaMIN,lambdaMAX,by=step);
        dens.v <- vector();
        ii <- 1;
        for(i in lambda.v){
            dens.v[ii] <- (Q/(2*pi*sigma2))*sqrt( (lambdaMAX-i)*(i-lambdaMIN) )/i;
            ii <- ii+1;
        }
        return(list(min=lambdaMIN,max=lambdaMAX,step=step,lambda=lambda.v,dens=dens.v));
    }

    # A Function to draw heatmap, based on significance level of correlation between phenotype and SVD componnents.
    drawheatmap <- function(svdPV.m)
    {
        myPalette <- c("darkred","red","orange","pink","white");
        breaks.v <- c(-10000,-10,-5,-2,log10(0.05),0);
        image(x=1:nrow(svdPV.m), y=1:ncol(svdPV.m), z=log10(svdPV.m), col=myPalette, breaks=breaks.v, xlab="", ylab="", axes=FALSE, main= "Singular Value Decomposition Analysis (SVD)");
        axis(1,at=1:nrow(svdPV.m),labels=paste("PC-",1:nrow(svdPV.m),sep=""),las=2);
        suppressWarnings(axis(2,at=1:ncol(svdPV.m),labels=colnames(svdPV.m),las=2));
        legend(x=-(topPCA/2.5),y=3,legend=c(expression("p < 1x"~10^{-10}),expression("p < 1x"~10^{-5}),"p < 0.01", "p < 0.05", "p > 0.05"), fill=c("darkred","red","orange","pink",       "white"),par('usr')[2], par('usr')[4], xpd=NA);

    }
    # Function to organize a list for ggplot the screeplot.
	splot <- function(x=svd.o,y=rmt.o)  
	{
    	scp <- list(u=x$u[1:nrow(x$u),1:y$dim],v=x$v[1:y$dim,1:y$dim],d=x$d[1:y$dim])
    	return(invisible(capture.output(svd.scree(scp))))
  	}
    # Function to draw screeplot. Contributed by Rasmus.
	svd.scree <- function(svd.obj, maintitle="Scree Plot", axis.title.x="Component Index (Singular Vectors)", axis.title.y="Percent Variance Explained") 
    {
        if(is.list(svd.obj) & all(names(svd.obj) %in% c("u","d","v"))) {
            print("Your input data is treated as a SVD output, with u, d, v corresponding to left singular vector, singular values, and right singular vectors, respectively.")
        } else {
            print("Your input data is treated as a vector of singular values. For example, it should be svd.obj$d from a SVD output.")
            svd.obj = list(d=svd.obj)
        }
        print("Scree Plot")
        if(is.null(names(svd.obj$d))) 
        {
            names(svd.obj$d) = paste0("Component",1:ncol(svd.obj$v))
        }
        pve = svd.obj$d^2/sum(svd.obj$d^2) * 100
        pve = data.frame(names=names(svd.obj$d), pve)
        pve$names = factor(pve$names, levels=unique(names(svd.obj$d)))
        g = ggplot2::ggplot(pve, aes(names, pve)) + geom_point() + theme_bw()
        gout = g + ylim(0,NA) + 
            ggtitle(maintitle) + # for the main title
            xlab(axis.title.x) + # for the x axis label
            ylab(axis.title.y) # for the y axis label
        
        return(gout)
    }

	################ Main Program starts here ########################

    if (!file.exists(resultsDir)) dir.create(resultsDir)
    message("champ.SVD Results will be saved in ",resultsDir," .\n")

    if(length(which(is.na(beta)))>0) message(length(which(is.na(beta)))," NA are detected in your beta Data Set, which may cause fail or uncorrect of SVD analysis. You may want to impute NA with champ.impute() function first.")

    if(class(beta) == 'data.frame')
    {
        message("Your beta parameter is data.frame format. ChAMP is now changing it to matrix.")
        beta <- as.matrix(beta)
    }

    message("[SVD analysis will be proceed with ",dim(beta)[1], " probes and ",dim(beta)[2], " samples.]\n")

    message("\n[ champ.SVD() will only check the dimensions between data and pd, instead if checking if Sample_Names are correctly matched (because some user may have no Sample_Names in their pd file),thus please make sure your pd file is in accord with your data sets (beta) and (rgSet).]\n")


	################ Customise Phenotype Data ########################

    if(is.null(pd) | class(pd)=="list") stop("pd parameter in Data Frame or Matrix is necessary And must contain at least tow factors. If your pd is a list, please change its Format.")
    if(class(pd)=="matrix") pd <- as.data.frame(pd)

    PhenoTypes.lv_tmp <- pd[,!colnames(pd) %in% c("Sample_Name","Project","filenames","Basename") & apply(pd,2,function(x) length(unique(x)))!=1]
    PhenoTypes.lv <- PhenoTypes.lv_tmp

    if(!is.null(rownames(pd))) rownames(PhenoTypes.lv) <- rownames(pd)

    if(ncol(PhenoTypes.lv)>=2)
    {
        message("<< Following Factors in your pd(sample_sheet.csv) will be analysised: >>")
        sapply(colnames(PhenoTypes.lv_tmp),function(x) message("<",x,">(",class(PhenoTypes.lv[[x]]),"):",paste(unique(PhenoTypes.lv_tmp[,x]),collapse=", ")))
        message("[champ.SVD have automatically select ALL factors contain at least two different values from your pd(sample_sheet.csv), if you don't want to analysis some of them, please remove them manually from your pd variable then retry champ.SVD().]")
    }else
    {
        stop("You don't have even one factor with at least two value to be analysis. Maybe your factors contains only one value, no variation at all...")
    }

    if(ncol(pd) > ncol(PhenoTypes.lv))
    {
        message("\n<< Following Factors in your pd(sample_sheet.csv) will not be analysis: >>")
        sapply(setdiff(colnames(pd),colnames(PhenoTypes.lv)),function(x) message("<",x,">"))
        message("[Factors are ignored because they only indicate Name or Project, or they contain ONLY ONE value across all Samples.]")
    }
    
    #### PhenoTypes.lv prepare ready.
    if(RGEffect==TRUE & is.null(rgSet)) message("If you want to check Effect of Control Probes, you MUST provide rgSet parameter. Now champ.SVD can only analysis factors in pd.")
	if(!is.null(rgSet) & RGEffect)
	{
        if(rgSet@annotation[1]=="IlluminaHumanMethylation450k") data(ControlProbes450K) else data(ControlProbesEPIC)
        dataC2.m <- as.data.frame(log2(apply(ControlProbes,1,function(x) if(x[3]=="Grn") getGreen(rgSet)[x[2],] else getRed(rgSet)[x[2],])))
        PhenoTypes.lv <- cbind(PhenoTypes.lv,dataC2.m)
        message("\n<< Following rgSet information have been added to PhenoTypes.lv. >>")
        sapply(colnames(dataC2.m),function(x) message("<",x,">"))
        #### RG Information extract!
    }	

    if(nrow(PhenoTypes.lv)==ncol(beta)) message("\n<< PhenoTypes.lv generated successfully. >>") else stop("Dimension of your pd file (and rgSet information) is not equal to your beta matrix.")

	######################## Do SVD #############################

    tmp.m <- beta-rowMeans(beta)
	rmt.o <- EstDimRMTv2(tmp.m);
	svd.o <- svd(tmp.m);
    if(rmt.o$dim > 20) topPCA <- 20  else topPCA <- rmt.o$dim
        
    svdPV.m <- matrix(nrow=topPCA,ncol=ncol(PhenoTypes.lv));
    colnames(svdPV.m) <- colnames(PhenoTypes.lv);

	for(c in 1:topPCA)
        for(f in 1:ncol(PhenoTypes.lv))
            if(class(PhenoTypes.lv[,f])!="numeric")
                svdPV.m[c,f] <- kruskal.test(svd.o$v[,c] ~ as.factor(PhenoTypes.lv[[f]]))$p.value
            else
                svdPV.m[c,f] <- summary(lm(svd.o$v[,c] ~ PhenoTypes.lv[[f]]))$coeff[2,4];

    message("<< Calculate SVD matrix successfully. >>")

	######################## Plot SVD Image #############################

    if(Rplot)
    {
        par(mar=c(5,15,2,1));
        splot(svd.o,rmt.o)
        drawheatmap(svdPV.m)
    }
	       
  #Screeplot end
    if(PDFplot)
    {
        pdf(paste(resultsDir,"SVDsummary.pdf",sep=""),width=10, height=8);
        par(mar=c(5,20,2,1))
        drawheatmap(svdPV.m)
        splot(svd.o,rmt.o)
        dev.off();
    }

    message("<< Plot SVD matrix successfully. >>")

    message("[<<<<<< ChAMP.SVD END >>>>>>]")
    message("[===========================]")
    message("[If the batch effect is not significant, you may want to process champ.DMP() or champ.DMR() or champ.BlockFinder() next, otherwise, you may want to run champ.runCombat() to eliminat batch effect, then rerun champ.SVD() to check corrected result.]\n")
    return(svdPV.m)
}
```

```{r, message=FALSE}
champ.SVD(beta = as.data.frame(cleaned_beta_s),
          pd = pheno,
          RGEffect=FALSE,
          PDFplot=TRUE,
          Rplot=TRUE,
          resultsDir="./CHAMP_SVDimages/")
```


Check the relationship between covariates
```{r}
#| layout-ncol: 2
library(ggalluvial)
ggplot(data = pheno,
       aes(axis1 = Date_Scan, axis2 = SentrixBarcode_A, axis3 = AMP_Plate)) +
  geom_alluvium(aes(fill = Date_Scan), width = 1/12) +
  geom_stratum(width = 1/12, fill = "grey") +
  geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
  scale_x_discrete(limits = c("Date_Scan", "SentrixBarcode_A", "AMP_Plate")) +
  labs(title = "Alluvial Plot of Categorical Relationships") +
  theme_minimal()

ggplot(data = pheno,
       aes(axis1 = Date_Scan, axis2 = Sample_Plate, axis3 = AMP_Plate)) +
  geom_alluvium(aes(fill = Date_Scan), width = 1/12) +
  geom_stratum(width = 1/12, fill = "grey") +
  geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
  scale_x_discrete(limits = c("Date_Scan", "Sample_Plate", "AMP_Plate")) +
  labs(title = "Alluvial Plot of Categorical Relationships") +
  theme_minimal()
```

Check the relationship between SentrixPosition_A, ID_Patient and Date_Scan

```{r}
ggplot(data = pheno,
       aes(axis1 = Date_Scan, axis2 = ID_Patient, axis3 = SentrixPosition_A)) +
  geom_alluvium(aes(fill = Date_Scan), width = 1/12) +
  geom_stratum(width = 1/12, fill = "grey") +
  geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
  scale_x_discrete(limits = c("Date_Scan", "ID_Patient", "SentrixPosition_A")) +
  labs(title = "Alluvial Plot of Categorical Relationships") +
  theme_minimal()
```

Check that in the same Date_Scan, if Sentrix_Barcode_A has technical variation:
```{r}
#| layout-ncol: 2
for (date in unique(pheno$Date_Scan)) {
  subset <- cleaned_beta_s[, pheno$Sample[pheno$Date_Scan == date]]
  densityPlot(subset, 
              sampGroups=factor(pheno$SentrixBarcode_A[pheno$Date_Scan == date]),
              main=paste0("Density plot of beta values of ", date),
              xlab="Beta", legend = FALSE)
}
```

covariate: Date_Scan or SentrixBarcode_A, and SentrixPosition_A

There is 1 sample scanned (and only scanned) at one date.

# Batch correction

```{r, output=FALSE}
cleaned_beta_s_corrected <- champ.runCombat(beta=cleaned_beta_s,
                                            pd=pheno,
                                            variablename="Timepoint",
                                            batchname=c("Date_Scan"))
```

```{r}
#| layout-ncol: 2
densityPlot(cleaned_beta_s, sampGroups=pheno$Date_Scan,
            main=paste0("Density plot of beta values before batch correction"),
            xlab="Beta", legend=FALSE)

densityPlot(cleaned_beta_s_corrected, sampGroups=pheno$Date_Scan,
            main=paste0("Density plot of beta values after batch correction"),
            xlab="Beta", legend=FALSE)
```

```{r, output=FALSE}
saveRDS(cleaned_beta_s, "cleaned_beta_s.RDS")
saveRDS(cleaned_beta_s_corrected, "cleaned_beta_s_corrected.RDS")

rownames(pheno) <- pheno$Sample
cleaned_beta_p_corrected <- champ.runCombat(beta=cleaned_beta_p,
                                            pd=pheno[colnames(cleaned_beta_p),],
                                            variablename="Timepoint",
                                            batchname=c("Date_Scan"))

saveRDS(cleaned_beta_p, "cleaned_beta_p.RDS")
saveRDS(cleaned_beta_p_corrected, "cleaned_beta_p_corrected.RDS")
```

# Meta data

```{r}
head(blmeta)
```

<!-- * Capillary glycemia: 毛细血管血糖 -->
<!-- * Ketosemia: 血酮 -->
<!-- ketosemia vs ketones -->

Cell type information:

- erythrocytes
- hematocrit (the percentage by volume of red cells in blood)
- hemoglobin
- leukocytes
- thrombocytes


# Session information

```{r}
sessionInfo()
```
