---
title: "Differential methylation analysis"
subtitle: "Longitudinal limma models for fasting-associated methylation changes"
abstract: |
  This notebook tests differential DNA methylation across the fasting study
  timepoints D-1, D+10, and M+1 using filtered, uncorrected beta matrices from
  `2_qc_and_prepare_filtered_betas.qmd`. Beta values are transformed to
  M-values for limma modeling, repeated measurements from the same participant
  are handled with `duplicateCorrelation`, and scan-date batch effects are
  controlled by including `Date_Scan_recode` in the design matrix rather than by
  using pre-corrected beta values. The primary model compares timepoints while
  adjusting for age and sex; additional models evaluate adjustment for blood
  cell composition, baseline BMI, and Timepoint x baseline-BMI interactions.
  The notebook saves fitted model objects, compares DMP overlap across models,
  estimates covariate-adjusted delta-beta effect sizes, runs GO and KEGG
  enrichment with `gometh`, performs DMRcate region analysis, and generates
  manuscript-ready Manhattan, volcano, genomic-feature, and CpG-context
  visualizations.
format:
  html:
    theme: sandstone
    embed-resources: true
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
    df-print: paged
editor_options: 
  chunk_output_type: console
---

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 5,
  dpi = 150,
  out.width = "100%",
  class.output = "bg-light"
)
```

```{r, results=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(ggrastr)
library(tidyr)
library(cowplot)
library(dplyr)
library(sesame)
library(sesameData)
library(readxl)
library(limma)
library(UpSetR)
library(minfi)
library(DMRcate)
library(missMethyl)
library(ggpubr)
library(ggrepel)
library(knitr)
library(gridExtra)
library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
library(enrichplot)
source("utils.R")
theme_set(theme_minimal(base_size = 12))
```

```{r output-helpers}
render_table <- function(x, caption = NULL, digits = 3) {
  if (is.null(x) || nrow(x) == 0) {
    return(knitr::kable(
      data.frame(message = "No rows to display."),
      caption = caption,
      format = "html"
    ))
  }

  knitr::kable(x, digits = digits, caption = caption, format = "html")
}

plot_top_cpg_panels <- function(result, pheno, pheno_name, n = 6) {
  cpgs <- rownames(result)[seq_len(min(n, nrow(result)))]
  plot_df <- as.data.frame(betass[cpgs, meta$Sample, drop = FALSE]) %>%
    tibble::rownames_to_column("CpG") %>%
    pivot_longer(-CpG, names_to = "Sample", values_to = "beta") %>%
    mutate(
      Sample = sub("^X", "", Sample),
      pheno = pheno[match(Sample, meta$Sample)]
    )

  ggplot(plot_df, aes(x = pheno, y = beta)) +
    geom_point(aes(color = pheno), alpha = 0.85, show.legend = FALSE) +
    facet_wrap(~CpG, scales = "free_y") +
    labs(x = pheno_name, y = "Beta values")
}

format_decide_tests <- function(fit, caption = NULL) {
  summary(decideTests(fit, adjust.method = "fdr")) %>%
    as.data.frame() %>%
    tibble::rownames_to_column("direction") %>%
    render_table(caption = caption)
}

format_top_table <- function(x, n = 10, caption = NULL) {
  x %>%
    as.data.frame() %>%
    tibble::rownames_to_column("CpG") %>%
    head(n) %>%
    render_table(caption = caption)
}
```

```{r, results=FALSE, message=FALSE, warning=FALSE}
proj_dir <- "~/public/SiyuanLuo/projects/fasting_ms"
setwd(proj_dir)
betass <- readRDS("./data/cleaned_beta_s.RDS")
betasp <- readRDS("./data/cleaned_beta_p.RDS")

meta <- read.csv(file="./data/meta_3timepoints.csv")
meta <- meta %>% filter(meta$Sample %in% colnames(betass))
meta$ID <- factor(meta$ID) 
```

```{r}
# calculate M-values for statistical analysis
# using all samples
M_values <- log2(betass/(1-betass))
mdsPlot(M_values, numPositions=1000,
        sampNames = NULL, sampGroups=meta$Timepoint)
```

Timepoint doesn’t separate the methylation data.

Due to the longitudinal design, use a mixed effect model.

Since we loaded the uncorrected data, we use the Date_Scan as a fixed effect for controlling batch effect.

# Compare between timepoints

::: {.callout-note}
Outlier removal:
* blood cells:
  * patient 28 is an outlier due to high thrombocytes. This is unlikely to affect the DNA composition.
  * patient 29 is an outlier due to low MCH (Mean Corpuscular Hemoglobin). This is unlikely to affect the DNA composition.
* liver function: patient 10, 2, 14
* energy metabolism: patient 10

Let’s remove patient 10 since the BMI is also extreme. 

I tried keeping vs removing 2 and 14 and in the end decide to keep them.
:::

```{r}
meta <- meta %>% filter(as.character(ID) !=10)
rownames(meta) <- meta$Sample
M_values <- M_values[, rownames(meta)]

meta <- meta[!is.na(meta$Sample),]
rownames(meta) <- meta$Sample
meta <- meta[colnames(M_values),]

# Create a model matrix for time points and any other covariates
meta$Timepoint2 <- factor(meta$Timepoint, levels = c("D-1", "D+10", "M+1"))
meta <- meta %>%
  mutate(Timepoint = case_when(
    Timepoint2 == "D-1"  ~ "BEFORE",
    Timepoint2 == "D+10" ~ "AFTER",
    Timepoint2 == "M+1"  ~ "FUP1",
    TRUE ~ Timepoint2
  )) %>%
  mutate(Timepoint = factor(Timepoint, levels = c("BEFORE","AFTER","FUP1"))) %>%
  mutate(ID_Patient = factor(ID_Patient))
meta$Date_Scan_recode <- factor(as.numeric(factor(meta$Date_Scan)))
meta$sex <- factor(meta$sex)

# Sanity check for covariates
stopifnot("LYMPHOZYTEN" %in% colnames(meta))
meta <- meta %>%
  mutate(
    LYMPHOZYTEN = as.numeric(LYMPHOZYTEN)
  )
stopifnot(!anyNA(meta$LYMPHOZYTEN))
```

::: {.panel-tabset}
## Model 1
Model: Timepoint2 + Date_Scan + age + sex + (1 | ID_Patient)

### DMP
```{r}
design <- model.matrix(~ 0 + Timepoint + Date_Scan_recode + age + sex, data = meta)
```

```{r, eval=FALSE}
# Estimate correlations between repeated measures from the same individual
corfit <- duplicateCorrelation(M_values, design, block = meta$ID_Patient)
saveRDS(corfit, file="./data/corfit_design1_rm10.RDS")
```

```{r}
corfit <- readRDS(file="./data/corfit_design1_rm10.RDS")
render_table(
  data.frame(consensus_correlation = corfit$consensus),
  caption = "Estimated consensus correlation between repeated participant measurements"
)
```

```{r}
# fit the linear model
fit <- lmFit(M_values, design, block=meta$ID_Patient, correlation = corfit$consensus)
# Apply the contrasts and compute eBayes statistics
# Define pairwise contrasts
contrasts <- makeContrasts(
  BEFORE_vs_AFTER = TimepointBEFORE - TimepointAFTER,
  BEFORE_vs_FUP1 = TimepointBEFORE - TimepointFUP1,
  AFTER_vs_FUP1 = TimepointAFTER - TimepointFUP1,
  age = age,
  Female_vs_Male = sex1,
  levels = design
)
fit2 <- contrasts.fit(fit, contrasts)
fit2 <- eBayes(fit2)

# Extract results for each contrast
results_BEFORE_vs_AFTER <- topTable(fit2, coef = "BEFORE_vs_AFTER", number = Inf)
results_BEFORE_vs_FUP1 <- topTable(fit2, coef = "BEFORE_vs_FUP1", number = Inf)
results_AFTER_vs_FUP1 <- topTable(fit2, coef = "AFTER_vs_FUP1", number = Inf)
results_age <- topTable(fit2, coef = "age", number = Inf)
results_sex <- topTable(fit2, coef = "Female_vs_Male", number = Inf)

# look at the numbers of DM CpGs at FDR < 0.05
format_decide_tests(fit2, caption = "Model 1 DMP counts at FDR < 0.05")
```

```{r}
saveRDS(fit2, file="./data/fit_model1.RDS")
```

#### Check the effect size
The distribution of log-fold change:

```{r, fig.width=7, fig.height=4}
hist(results_BEFORE_vs_FUP1[results_BEFORE_vs_FUP1$adj.P.Val<0.05, "logFC"], 
     breaks=50,
     main = "Histogram of logFC for significant BEFORE vs FUP1 CpGs")
```

Check the overlap between DMPs between pairs of timepoints:

```{r, fig.width=8, fig.height=5}
dm <- list(BEFORE_vs_FUP1_up = rownames(results_BEFORE_vs_FUP1[results_BEFORE_vs_FUP1$adj.P.Val<0.05 & results_BEFORE_vs_FUP1$logFC>0,]),
           BEFORE_vs_FUP1_down = rownames(results_BEFORE_vs_FUP1[results_BEFORE_vs_FUP1$adj.P.Val<0.05 & results_BEFORE_vs_FUP1$logFC<0,]),
     AFTER_vs_FUP1_down = rownames(results_AFTER_vs_FUP1[results_AFTER_vs_FUP1$adj.P.Val<0.05 & results_AFTER_vs_FUP1$logFC<0,]),
    AFTER_vs_FUP1_up = rownames(results_AFTER_vs_FUP1[results_AFTER_vs_FUP1$adj.P.Val<0.05 & results_AFTER_vs_FUP1$logFC>0,]))

upset(fromList(dm), empty.intersections = TRUE)
```

Plotting the top few differentially methylated CpGs:

##### Age
```{r, fig.width=9, fig.height=6}
plot_top_cpg_panels(results_age, meta$age, "Age", n = 6)
```

##### Sex
```{r, fig.width=9, fig.height=6}
plot_top_cpg_panels(results_sex, meta$Gender, "Gender", n = 6)
```

#### BEFORE vs FUP1

```{r, fig.width=10, fig.height=7}
tmp <- data.frame(betass[rownames(results_BEFORE_vs_FUP1)[1:9], meta$Sample]) 
tmp$cpg <- rownames(tmp)
tmp <- tmp%>% pivot_longer(cols = -cpg, names_to = "Sample", values_to = "beta")
tmp$Sample <- sub("X", "", tmp$Sample)

merge(tmp, meta, by="Sample") %>% 
  filter(Timepoint %in% c("BEFORE", "FUP1")) %>%
  ggplot(aes(x=Timepoint, y=beta, color=Gender, group=ID_Patient)) + 
  geom_point() + 
  geom_line() + 
  facet_wrap(~cpg, scales = "free") + 
  labs(title="BEFORE vs FUP1")
```

### GO testing
#### BEFORE_vs_FUP1
```{r}
DMPs <- results_BEFORE_vs_FUP1
format_top_table(DMPs, n = 10, caption = "Top BEFORE vs FUP1 CpGs from model 1")

# Get the significant CpG sites at less than 5% FDR
sigCpGs <- rownames(DMPs[DMPs$adj.P.Val<0.05,])

# Get all the CpG sites used in the analysis to form the background
all <- rownames(DMPs)
render_table(
  data.frame(
    metric = c("Significant CpGs at FDR < 0.05", "CpGs tested in background"),
    n = c(length(sigCpGs), length(all))
  ),
  caption = "CpG counts used for model 1 GO testing"
)

# Run enrichment - Can take a bit of time...
gst <- gometh(sig.cpg=sigCpGs, all.cpg=all)

# Top 10 GO categories
render_table(topGSA(gst, number=30), caption = "Top model 1 GO enrichment results")
```


### DMR
```{r}
myAnnotation <- cpg.annotate(object = M_values, datatype = "array", what = "M",
                                analysis.type= "differential", design = design,
                                contrasts = TRUE, cont.matrix = contrasts,
                                coef = "BEFORE_vs_FUP1", arraytype = "EPICv1", 
                             fdr = 0.08)
# Run DMRcate
DMRs <- dmrcate(myAnnotation, lambda=1000, C=2)
# convert the regions to annotated genomic ranges
sigRegions <- extractRanges(DMRs)
render_table(
  as.data.frame(sigRegions) %>% head(10),
  caption = "Top BEFORE vs FUP1 DMRcate regions"
)
```

```{r, fig.width=7, fig.height=4}
dmr_regions <- sigRegions[sigRegions$no.cpgs >= 5, ]
dmr_maxdiff <- as.numeric(dmr_regions$maxdiff)
dmr_maxdiff <- dmr_maxdiff[is.finite(dmr_maxdiff)]

if (length(dmr_maxdiff) > 0) {
  hist(dmr_maxdiff, 
       main = "Histogram of BEFORE vs FUP1 DMR maxdiff",
       breaks = 50)
} else {
  message("No BEFORE vs FUP1 DMRs with at least 5 CpGs; skipping maxdiff histogram.")
}

render_table(
  as.data.frame(dmr_regions[as.numeric(dmr_regions$maxdiff) > 0.1, ]),
  caption = "BEFORE vs FUP1 DMRs with maxdiff > 0.1"
)
```

* PEX11B: Peroxisomal Biogenesis Factor 11 Beta
* GNRHR2: Gonadotropin Releasing Hormone Receptor 2 (Pseudogene)
* GPN2: ATP-Binding Domain 1 Family Member B
* GPATCH3: G-Patch Domain Containing 3
* LFNG: LFNG O-Fucosylpeptide 3-Beta-N-Acetylglucosaminyltransferase

```{r}
myAnnotation <- cpg.annotate(object = M_values, datatype = "array", what = "M",
                                analysis.type= "differential", design = design,
                                contrasts = TRUE, cont.matrix = contrasts,
                                coef = "AFTER_vs_FUP1", arraytype = "EPICv1", 
                             fdr = 0.1)
```

## Model 2
Model: Timepoint + Date_Scan + age + sex + ? + (1 | ID_Patient)

### Collect the candidate covariates
Collect the candidate covariates
* BMI_baseline_category/BMI_baseline
* blood cell composition: "LYMPHOZYTEN"

### Check the covariates + timepoint
```{r}
# covariate sets
candidate_covariates <- list(
  "LYMPHOZYTEN",
  "BMI_baseline_category",
  "BMI_baseline"
)
names(candidate_covariates) <- unlist(candidate_covariates)

# Build formulas (no intercept, so timepoint has separate columns)
models <- lapply(candidate_covariates, function(cov_rhs) {
  as.formula(paste("~ 0 + Timepoint + Date_Scan_recode + age + sex +", cov_rhs))
})

# Helper to make filename-safe labels
safe_name <- function(x) gsub("[^A-Za-z0-9_]+", "_", x)

# Ensure factor coding is as expected (important for sex and timepoint column names)
# meta$Timepoint <- factor(meta$Timepoint, levels = c("BEFORE","AFTER","FUP1"))
# meta$sex <- relevel(factor(meta$sex), ref = "0")  # depends on your encoding

model2_decide_summaries <- list()

for (model_name in names(models)) {

  design <- model.matrix(models[[model_name]], data = meta)
  colnames(design) <- make.names(colnames(design))  # robust; avoids weird column names

  # --- correlation fit caching ---
  corfit_file <- file.path("./data", paste0("corfit_", safe_name(model_name), ".RDS"))
  if (!file.exists(corfit_file)) {
    corfit <- duplicateCorrelation(M_values, design, block = meta$ID_Patient)
    saveRDS(corfit, corfit_file)
  }
  corfit <- readRDS(corfit_file)

  # --- fit model ---
  fit <- lmFit(M_values, design, block = meta$ID_Patient, correlation = corfit$consensus)

  # --- define ONLY the timepoint contrasts ---
  # IMPORTANT: use the *actual* coefficient names in design
  # (after make.names). Usually these will be "TimepointBEFORE", etc.
  tp_cols <- grep("^Timepoint", colnames(design), value = TRUE)
  # sanity check
  stopifnot(all(c("TimepointBEFORE", "TimepointAFTER", "TimepointFUP1") %in% tp_cols))

  contrasts <- makeContrasts(
    BEFORE_vs_AFTER = TimepointBEFORE - TimepointAFTER,
    BEFORE_vs_FUP1  = TimepointBEFORE - TimepointFUP1,
    AFTER_vs_FUP1   = TimepointAFTER  - TimepointFUP1,
    levels = design
  )

  fit2 <- contrasts.fit(fit, contrasts)
  fit2 <- eBayes(fit2)

  # --- extract timepoint results ---
  res_BEFORE_vs_AFTER <- topTable(fit2, coef = "BEFORE_vs_AFTER", number = Inf)
  res_BEFORE_vs_FUP1  <- topTable(fit2, coef = "BEFORE_vs_FUP1",  number = Inf)
  res_AFTER_vs_FUP1   <- topTable(fit2, coef = "AFTER_vs_FUP1",   number = Inf)

  # --- extract covariate results from the *original* fit (not contrast fit) ---
  # Covariates are not contrasts; they are coefficients in `fit`.
  fit_cov <- eBayes(fit)

  res_age <- topTable(fit_cov, coef = "age", number = Inf)

  # sex coefficient name depends on coding. If sex is numeric 0/1, model.matrix gives "sex"
  # If sex is factor with 2 levels, it might be "sex1" or "sexFemale".
  sex_coef <- grep("^sex", colnames(design), value = TRUE)
  # prefer the non-timepoint sex coefficient (exclude any weird matches)
  sex_coef <- setdiff(sex_coef, tp_cols)
  # pick one if exactly one
  if (length(sex_coef) == 1) {
    res_sex <- topTable(fit_cov, coef = sex_coef, number = Inf)
  } else {
    res_sex <- NULL  # or stop() if you want strictness
  }

  # covariate(s) - could be 1 term or multiple
  cov_terms <- all.vars(as.formula(paste("~", model_name)))
  cov_coefs <- cov_terms[cov_terms %in% colnames(design)]
  # If BMI_baseline_category expands to multiple dummy columns, capture them:
  cov_coefs <- c(cov_coefs, grep("^BMI_baseline_category", colnames(design), value = TRUE))
  cov_coefs <- unique(cov_coefs)

  res_cov <- lapply(cov_coefs, function(cc) topTable(fit_cov, coef = cc, number = Inf))
  names(res_cov) <- cov_coefs

  # --- summaries ---
  model2_decide_summaries[[model_name]] <- summary(decideTests(fit2, adjust.method = "fdr")) %>%
    as.data.frame() %>%
    tibble::rownames_to_column("direction") %>%
    mutate(model = model_name, .before = direction)

  # --- save ---
  saveRDS(
    list(
      fit2 = fit2,
      fit_cov = fit_cov,
      results = list(
        BEFORE_vs_AFTER = res_BEFORE_vs_AFTER,
        BEFORE_vs_FUP1  = res_BEFORE_vs_FUP1,
        AFTER_vs_FUP1   = res_AFTER_vs_FUP1,
        age = res_age,
        sex = res_sex,
        covariates = res_cov
      ),
      design_colnames = colnames(design),
      model = models[[model_name]]
    ),
    file = file.path("./data", paste0("fit_model2_", safe_name(model_name), ".RDS"))
  )
}

render_table(
  bind_rows(model2_decide_summaries),
  caption = "Model 2 candidate-covariate DMP counts at FDR < 0.05"
)
```

Some observations

- BMI_baseline do have effects on the methylation levels. Since it shouldn’t confound the time effect, it should be added as an interaction term.
- LYMPHOZYTEN has strong effects on the methylation levels, but it is independent of the time effect (from previous correlation analysis), suggesting that it is a strong confounding factor. It should be adjusted for. This can be validated via analyzing the DM CpGs associated with LYMPHOZYTEN.

## Model 3
Model: Timepoint * BMI_baseline + Date_Scan + age + sex + (1 | ID_Patient)

When adding a continuous variable in the interaction term, it is important to model the intercept. For categorical variable only, no interaction term is fine.

```{r}
meta$BMI_baseline_centered <- scale(meta$BMI_baseline, center = TRUE, scale = FALSE)
design <-model.matrix(~ Timepoint*BMI_baseline_centered + Date_Scan_recode + age + sex, data = meta)
corfit_file <- paste0("./data/corfit_design3_interaction.RDS")
  if (!file.exists(corfit_file)) {
    corfit <- duplicateCorrelation(M_values, design, block = meta$ID_Patient)
    saveRDS(corfit, file = corfit_file)
  }
  corfit <- readRDS(corfit_file)
  # fit the linear model
colnames(design) <- make.names(colnames(design))
fit <- lmFit(M_values, design, block=meta$ID_Patient, correlation = corfit$consensus)
# Define pairwise contrasts
contrasts <- makeContrasts(
  age = age,
  Female_vs_Male = sex1,
  BMI_baseline = BMI_baseline_centered,
  BEFORE_vs_AFTER = TimepointAFTER,
  BEFORE_vs_FUP1 = TimepointFUP1,
  AFTER_vs_FUP1 = TimepointAFTER - TimepointFUP1,
  BMI_slope_AFTER_vs_BEFORE = TimepointAFTER.BMI_baseline_centered,  # Does BMI modify T2 vs. T1?
  BMI_slope_FUP1_vs_BEFORE = TimepointFUP1.BMI_baseline_centered,  # Does BMI modify T3 vs. T1?
  BMI_slope_FUP1_vs_AFTER = TimepointFUP1.BMI_baseline_centered - TimepointAFTER.BMI_baseline_centered,  # Does BMI modify T3 vs. T2?
  levels = colnames(design)
)
fit2 <- contrasts.fit(fit, contrasts)
fit2 <- eBayes(fit2)
# look at the numbers of DM CpGs at FDR < 0.05
format_decide_tests(fit2, caption = "Model 3 interaction DMP counts at FDR < 0.05")
saveRDS(fit2, file="./data/fit_model3_interaction.RDS")
```
:::

# BEFORE vs FUP1
## Compare CpGs in model 1 and model 2
```{r}
# model 1
design <- model.matrix(~ 0 + Timepoint + Date_Scan_recode + age + sex, data = meta)
corfit <- readRDS(file="data/corfit_design1_rm10.RDS")
fit <- lmFit(M_values, design, block=meta$ID_Patient, correlation = corfit$consensus)
contrasts <- makeContrasts(
  BEFORE_vs_FUP1 = TimepointBEFORE - TimepointFUP1,
  levels = design
)
fit2 <- contrasts.fit(fit, contrasts)
fit2 <- eBayes(fit2)

# Extract results for each contrast
results_BEFORE_vs_FUP1 <- topTable(fit2, coef = "BEFORE_vs_FUP1", number = Inf)
format_decide_tests(fit2, caption = "Model 1 BEFORE vs FUP1 DMP counts")
```

```{r}
p_th <- 0.05
lfc_th <- 0
```

```{r}
sigCpGsModel1 <- rownames(results_BEFORE_vs_FUP1[results_BEFORE_vs_FUP1$adj.P.Val<p_th,])
dm <- list(model1_up = rownames(results_BEFORE_vs_FUP1[results_BEFORE_vs_FUP1$adj.P.Val<p_th & results_BEFORE_vs_FUP1$logFC> lfc_th,]),
           model1_down = rownames(results_BEFORE_vs_FUP1[results_BEFORE_vs_FUP1$adj.P.Val<p_th & results_BEFORE_vs_FUP1$logFC< -lfc_th,]))
```

```{r}
# model 2
design <- model.matrix(~ 0 + Timepoint + Date_Scan_recode + age + sex + LYMPHOZYTEN, data = meta)
corfit_file <- paste0("data/corfit_", "LYMPHOZYTEN", ".RDS")
if (!file.exists(corfit_file)) {
  corfit <- duplicateCorrelation(M_values, design, block = meta$ID_Patient)
  saveRDS(corfit, file = corfit_file)
}
corfit <- readRDS(corfit_file)
fit <- lmFit(M_values, design, block=meta$ID_Patient, correlation = corfit$consensus)
contrasts <- makeContrasts(
  BEFORE_vs_AFTER = TimepointBEFORE - TimepointAFTER,
  BEFORE_vs_FUP1 = TimepointBEFORE - TimepointFUP1,
  AFTER_vs_FUP1 = TimepointAFTER - TimepointFUP1,
  levels = design
)
fit2 <- contrasts.fit(fit, contrasts)
fit2 <- eBayes(fit2)


# Extract results for each contrast
results_BEFORE_vs_AFTER <- topTable(fit2, coef = "BEFORE_vs_AFTER", number = Inf)
results_BEFORE_vs_FUP1 <- topTable(fit2, coef = "BEFORE_vs_FUP1", number = Inf)
results_AFTER_vs_FUP1 <- topTable(fit2, coef = "AFTER_vs_FUP1", number = Inf)

fit_cov <- eBayes(fit)
results_LYMPHOZYTEN <- topTable(fit_cov, coef = "LYMPHOZYTEN", number = Inf)
format_decide_tests(fit2, caption = "Model 2 lymphocyte-adjusted DMP counts")
```

Calculate the delta Beta values (effect size) adjusted for covariates:

```{r}
# 1. Reconstruct fitted M-values using model coefficients
coef_matrix <- fit$coefficients
design_matrix <- design[colnames(M_values), ]  # Ensure sample order matches
fitted_M <- coef_matrix %*% t(design_matrix)

# 2. Convert to adjusted beta values
adj_beta <- 2^fitted_M / (1 + 2^fitted_M)
rownames(adj_beta) <- rownames(M_values)

# 3. Define reference groups and contrast directions
contrast_specs <- list(
  BEFORE_vs_AFTER = list(ref = "BEFORE", comp = "AFTER"),
  BEFORE_vs_FUP1 = list(ref = "BEFORE", comp = "FUP1"),
  AFTER_vs_FUP1 = list(ref = "AFTER", comp = "FUP1")
)

# 4. Initialize results list
results_list <- list(
  BEFORE_vs_AFTER = results_BEFORE_vs_AFTER,
  BEFORE_vs_FUP1 = results_BEFORE_vs_FUP1,
  AFTER_vs_FUP1 = results_AFTER_vs_FUP1
)

# 5. Calculate delta beta for each contrast
for (contrast in names(results_list)) {
  spec <- contrast_specs[[contrast]]
  
  # Get reference samples
  ref_samples <- which(meta$Timepoint == spec$ref)
  
  # Get comparison samples
  comp_samples <- which(meta$Timepoint == spec$comp)
  
  # Calculate mean adjusted beta values
  ref_beta <- rowMeans(adj_beta[, ref_samples], na.rm = TRUE)
  comp_beta <- rowMeans(adj_beta[, comp_samples], na.rm = TRUE)
  
  # Get current results
  res <- results_list[[contrast]]
  cpg_ids <- rownames(res)
  # Match CpG order
  ref_beta <- ref_beta[cpg_ids]
  comp_beta <- comp_beta[cpg_ids]
  
  # Calculate delta beta (covariate-adjusted)
  delta_beta <- comp_beta - ref_beta
  
  # Add to results (direction matches contrast definition)
  res$delta_beta <- delta_beta
  res$ref_beta <- ref_beta
  results_list[[contrast]] <- res
}

# 6. Update results data frames
results_BEFORE_vs_AFTER <- results_list$BEFORE_vs_AFTER
results_BEFORE_vs_FUP1 <- results_list$BEFORE_vs_FUP1
results_AFTER_vs_FUP1 <- results_list$AFTER_vs_FUP1

# # 7. For continuous variables: approximation
# continuous_vars <- c("LYMPHOZYTEN", "age", "Female_vs_Male")
# for (var in continuous_vars) {
#   res <- get(paste0("results_", var))
#   
#   # Calculate median baseline beta
#   overall_mean_beta <- rowMeans(2^M_values/(1+2^M_values))
#   
#   # Approximation: Δβ ≈ logFC * β(1-β) (derivative of logit transform)
#   res$delta_beta <- res$logFC * overall_mean_beta[rownames(res)] * 
#                     (1 - overall_mean_beta[rownames(res)])
#   
#   assign(paste0("results_", var), res)
# }
```


```{r}
sigCpGsModel2 <- rownames(results_BEFORE_vs_FUP1[results_BEFORE_vs_FUP1$adj.P.Val<p_th,])

dm$model2_up <- rownames(results_BEFORE_vs_FUP1[results_BEFORE_vs_FUP1$adj.P.Val<p_th & results_BEFORE_vs_FUP1$logFC> lfc_th,])
dm$model2_down <- rownames(results_BEFORE_vs_FUP1[results_BEFORE_vs_FUP1$adj.P.Val<p_th & results_BEFORE_vs_FUP1$logFC< -lfc_th,])
dm$LYMPHOZYTEN_up <- rownames(results_LYMPHOZYTEN[results_LYMPHOZYTEN$adj.P.Val < p_th & results_LYMPHOZYTEN$logFC > lfc_th, ])
dm$LYMPHOZYTEN_down <- rownames(results_LYMPHOZYTEN[results_LYMPHOZYTEN$adj.P.Val < p_th & results_LYMPHOZYTEN$logFC < -lfc_th, ])

upset(fromList(dm[c("model1_up", "model1_down","model2_up", "model2_down")]), mainbar.y.label = "CpG intersections")
upset(fromList(dm[c("model1_up", "model2_up", "LYMPHOZYTEN_up")]), mainbar.y.label = "Up CpG intersections")
upset(fromList(dm[c("model1_down", "model2_down", "LYMPHOZYTEN_down")]), mainbar.y.label = "Down CpG intersections")
```

So it is true that almost all DM CpGs in model 2 are also present in model 1.

## GO Testing for model 2

```{r}
delta_beta_th <- 0.02
p_th <- 0.05
lfc_th <- 0
```

```{r}
sigCpGs <- rownames(results_BEFORE_vs_FUP1[results_BEFORE_vs_FUP1$adj.P.Val<p_th & 
                                             abs(results_BEFORE_vs_FUP1$logFC)>lfc_th &
                                             abs(results_BEFORE_vs_FUP1$delta_beta) > delta_beta_th,])
check <- getMappedEntrezIDs(sig.cpg = sigCpGs)
render_table(
  data.frame(
    metric = c("CpGs passing FDR and delta-beta filters", "Mapped Entrez genes"),
    n = c(length(sigCpGs), length(check$sig.eg))
  ),
  caption = "Model 2 GO-testing input size"
)
```

::: {.panel-tabset}
### GO
```{r, fig.width=8, fig.height=9}
gst <- gometh(sig.cpg=sigCpGs, all.cpg=rownames(results_BEFORE_vs_FUP1), 
              collection="GO", array.type = c("EPIC"),
              plot.bias=TRUE, sig.genes = FALSE)

render_table(topGSA(gst, n=20), caption = "Top GO enrichment results for model 2")

gst2compareClusterResult <- function(gst, cluster, n_DE_genes, description="Description"){
  gst$Cluster <- cluster
  df <- data.frame(
  Cluster = gst$Cluster,  
  ID = rownames(gst),  # GO term ID
  Description = gst[,description],  
  GeneRatio = paste0(gst$DE, "/", n_DE_genes),  # Ratio of DE genes
  pvalue = gst$P.DE,
  p.adjust = gst$FDR,
  qvalue = gst$FDR,  # No qvalue in gometh, so reuse FDR
  Count = gst$DE)
  
  df <- df[order(sapply(df$GeneRatio, function(x){eval(parse(text=x))}), decreasing = TRUE),]
# Convert to compareClusterResult object
  compare_cluster_obj <- new(
    "compareClusterResult",
    compareClusterResult = df,
    fun = "gometh"
  )
  return(compare_cluster_obj)
}

compare_cluster_obj <- gst2compareClusterResult(gst, 
                         cluster="AFTER_vs_FUP1", 
                         n_DE_genes=length(getMappedEntrezIDs(sig.cpg = sigCpGs)$sig.eg),
                         description="TERM")

enrichplot::dotplot(compare_cluster_obj, 
                    showCategory = 20, 
                    x = "GeneRatio", 
                    color = "p.adjust", 
                    size="count", 
                    by = "GeneRatio")
```

### KEGG
```{r, fig.width=8, fig.height=9}
gst.kegg <- gometh(sig.cpg=sigCpGs, all.cpg=rownames(results_BEFORE_vs_FUP1), 
                   array.type = c("EPIC"), collection="KEGG",
                   sig.genes = TRUE)

render_table(topGSA(gst.kegg, n=20), caption = "Top KEGG enrichment results for model 2")

compare_cluster_obj <- gst2compareClusterResult(gst.kegg, 
                         cluster="AFTER_vs_FUP1", 
                         n_DE_genes=length(getMappedEntrezIDs(sig.cpg = sigCpGs)$sig.eg),
                         description="Description")

enrichplot::dotplot(compare_cluster_obj, 
                    showCategory = 20, 
                    x = "GeneRatio", 
                    color = "p.adjust", 
                    size="count", 
                    by = "GeneRatio")
```

### KEGG promoter
```{r, fig.width=8, fig.height=9}
gst.kegg.prom <- gometh(sig.cpg=sigCpGs, all.cpg=rownames(results_BEFORE_vs_FUP1), array.type = c("EPIC"),
                        collection="KEGG", genomic.features = c("TSS200",
                                                                "TSS1500",
                                                                "1stExon"))
render_table(topGSA(gst.kegg.prom, n=20), caption = "Top promoter KEGG enrichment results")

compare_cluster_obj <- gst2compareClusterResult(gst.kegg.prom, 
                         cluster="AFTER_vs_FUP1", 
                         n_DE_genes=length(getMappedEntrezIDs(sig.cpg = sigCpGs)$sig.eg),
                         description="Description")

enrichplot::dotplot(compare_cluster_obj, 
                    showCategory = 20, 
                    x = "GeneRatio", 
                    color = "p.adjust", 
                    size="count", 
                    by = "GeneRatio")
```

### KEGG gene body
```{r}
gst.kegg.body <- gometh(sig.cpg=sigCpGs, all.cpg=rownames(results_BEFORE_vs_FUP1), 
                        collection="KEGG", array.type = c("EPIC"), genomic.features = c("Body"))
render_table(topGSA(gst.kegg.body, n=20), caption = "Top gene-body KEGG enrichment results")

gst.kegg.body <- gometh(sig.cpg=sigCpGs, all.cpg=rownames(results_BEFORE_vs_FUP1), 
                        collection="KEGG", array.type = c("EPIC"), genomic.features = c("Body"), sig.genes = TRUE)
render_table(topGSA(gst.kegg.body, n=10), caption = "Top gene-body KEGG enrichment results with significant genes")
```

:::

# Visualization
## M+1 vs D-1
```{r}
res <- results_BEFORE_vs_FUP1

ann <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
res$CpG <- rownames(res)
res$logP <- -log10(res$adj.P.Val)
ann_sub <- ann[rownames(ann) %in% res$CpG, ]
ann_sub$CpG <- rownames(ann_sub)
res$significance <- "Not Sig"
res$significance[res$adj.P.Val < p_th & res$logFC > lfc_th & res$delta_beta < -delta_beta_th] <- "Hypermethylated"
res$significance[res$adj.P.Val < p_th & res$logFC < -lfc_th& res$delta_beta > delta_beta_th] <- "Hypomethylated"
res_annot <- as.data.frame(merge(res, ann_sub, by = "CpG", all.x = TRUE))
res_sig <- res_annot[res_annot$significance != "Not Sig", ]
```

```{r, fig.width=11, fig.height=5}
# Manhattan plot
df <- res_annot %>% select(CpG, chr, pos, P.Value, adj.P.Val, delta_beta) 

# 2. Clean chromosome labels (keep 1..22, X, Y)
df <- df %>% 
  mutate(chr = gsub("chr", "", chr)) %>%
  filter(chr %in% c(1:22, "X", "Y")) %>%
  mutate(chr = factor(chr, levels=c(1:22,"X","Y")))

# 3. Compute cumulative basepair positions for each chromosome
chr_lengths <- df %>%
  group_by(chr) %>%
  summarise(chr_len = max(pos)) %>%
  arrange(chr)

chr_offsets <- cumsum(as.numeric(chr_lengths$chr_len)) - chr_lengths$chr_len
names(chr_offsets) <- chr_lengths$chr

df <- df %>%
  arrange(chr, pos) %>%
  mutate(bp_cum = pos + chr_offsets[as.character(chr)])

# 4. Axis positions for labeling chromosomes
axis_df <- df %>% 
  group_by(chr) %>% 
  summarise(center = (min(bp_cum) + max(bp_cum)) / 2)

# 5. Manhattan plot
p_manhattan <- ggplot(df, aes(x=bp_cum, y=-log10(P.Value), color=as.factor(chr))) +
  geom_point_rast(alpha=0.6, size=0.8) +
  scale_color_manual(values=rep(c("grey30","grey60"), length.out=length(unique(df$chr)))) +
  scale_x_continuous(label=axis_df$chr, breaks=axis_df$center) +
  labs(title = "M+1 vs D-1 differential methylation across the genome",
       x="Chromosome", y="-log10(p)") +
  theme_classic(base_size = 12) +
  theme(
    legend.position="none",
    plot.title = element_text(face = "bold", size = 13)
  ) +
  # geom_hline(yintercept=-log10(0.05), linetype="dashed", color="black") +
  geom_point(data=subset(df, adj.P.Val < 0.05 & delta_beta > 0.02), color="#B2182B", size=0.8) +
  geom_point(data=subset(df, adj.P.Val < 0.05 & delta_beta < -0.02), color="#2166AC", size=0.8)

p_manhattan

save(p_manhattan, file="figures/p_manhattan.RData")

pdf("figures/manhattan.pdf", width=15, height=6)
plot(p_manhattan)
dev.off() 
```

```{r, fig.width=8, fig.height=6}
n_hyper <- sum(res$significance == "Hypermethylated", na.rm = TRUE)
n_hypo <- sum(res$significance == "Hypomethylated", na.rm = TRUE)

# Basic volcano plot
volcano <- ggplot(res, aes(x = delta_beta, y = logP)) +
  rasterise(geom_point(aes(color = significance), alpha = 0.7), dpi = 300) +
  scale_color_manual(values = c("Hypermethylated" = "#B2182B", 
                                "Hypomethylated" = "#2166AC", 
                                "Not Sig" = "grey75")) +
  geom_vline(xintercept = c(-delta_beta_th, delta_beta_th), linetype = "dashed", color = "black") +
  geom_hline(yintercept = -log10(p_th), linetype = "dashed", color = "black") +
  theme_minimal(base_size = 12) +
  labs(title = "M+1 vs D-1 differential methylation",
       x = "Beta value change",
       y = "-log10(FDR)",
       color = "Significance")

p1 <- volcano +
  annotate("text", 
           x = max(res$delta_beta, na.rm = TRUE) * 0.35, 
           y = max(res$logP, na.rm = TRUE) * 0.85, 
           label = paste0("Hypermethylated: ", n_hyper, "\nHypomethylated: ", n_hypo),
           hjust = 0, size = 4, fontface = "italic")

top_hits <- rbind(res_sig[order(res_sig$adj.P.Val), ][1:10, ],res_sig[order(abs(res_sig$delta_beta), decreasing = TRUE), ][1:10, ])
top_hits <- top_hits %>% filter(UCSC_RefGene_Name != "")
top_hits$Gene_Symbol <- sapply(strsplit(top_hits$UCSC_RefGene_Name, ";"), `[`, 1)
p1 <- p1 + geom_text_repel(data = top_hits, aes(x = delta_beta, y = logP, label = Gene_Symbol), 
                           size = 2,
                           inherit.aes = FALSE)
p1
```

```{r, fig.width=7, fig.height=4.5}
plot_df <- as.data.frame(table(res_sig$significance, res_sig$Relation_to_Island))
colnames(plot_df) <- c("Direction", "Region", "Count")
plot_df$Region <- factor(plot_df$Region, 
                         levels = c("Island", "N_Shore", "S_Shore", "N_Shelf", "S_Shelf", "OpenSea"))

bg_df <- as.data.frame(table(res_annot$Relation_to_Island))
colnames(bg_df) <- c("Region", "Count")
bg_df$Region <- factor(bg_df$Region, 
                         levels = c("Island", "N_Shore", "S_Shore", "N_Shelf", "S_Shelf", "OpenSea"))
bg_df$Direction <- "all_CpGs"

p3 <- rbind(plot_df, bg_df) %>% mutate(Direction = factor(Direction, levels = c("all_CpGs", "Hypomethylated","Hypermethylated" ))) %>%
ggplot(aes(x = Direction, y = Count, fill =Region)) +
  geom_bar(position="fill", stat = "identity") +
  theme_minimal() +
  labs(title = "Distribution of Differentially Methylated CpGs",
       x = "Differentially Methylated CpGs", y = "Percentage", fill = "Region") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_flip() +
  scale_fill_brewer(palette="Set2")
p3
```

```{r, fig.width=7, fig.height=4.5}
# Assume res_sig has a column UCSC_RefGene_Group with semicolon-separated entries
res_long <- res_sig %>%
  separate_rows(UCSC_RefGene_Group, sep = ";") %>%
  filter(!is.na(UCSC_RefGene_Group))

# Now you can count and plot like before:
plot_df <- res_long %>%
  group_by(significance, UCSC_RefGene_Group) %>%
  summarise(Count = n(), .groups = "drop") %>%
  group_by(UCSC_RefGene_Group) %>%
  mutate(Ratio = Count / sum(Count))

plot_df$UCSC_RefGene_Group <- factor(plot_df$UCSC_RefGene_Group, 
                                        levels = c("Promoter", "TSS200","TSS1500","5'UTR","1stExon","Body","ExonBnd","3'UTR", ""), labels = c("Promoter", "TSS200","TSS1500","5'UTR","1stExon","Body","ExonBnd","3'UTR", "Others"))

bg_long <- res_annot %>%
  separate_rows(UCSC_RefGene_Group, sep = ";") %>%
  filter(!is.na(UCSC_RefGene_Group))

# Now you can count and plot like before:
bg_df <- bg_long %>%
  group_by(UCSC_RefGene_Group) %>%
  summarise(Count = n(), .groups = "drop") %>%
  group_by(UCSC_RefGene_Group) %>%
  mutate(Ratio = Count / sum(Count))

bg_df$UCSC_RefGene_Group <- factor(bg_df$UCSC_RefGene_Group, 
                                        levels = c("Promoter", "TSS200","TSS1500","5'UTR","1stExon","Body","ExonBnd","3'UTR", ""), labels = c("Promoter", "TSS200","TSS1500","5'UTR","1stExon","Body","ExonBnd","3'UTR", "Others"))

bg_df$significance <- "all_CpGs"

p4 <- rbind(plot_df, bg_df) %>% mutate(significance = factor(significance, levels = c("all_CpGs", "Hypomethylated", "Hypermethylated"))) %>%
ggplot(aes(x = significance, y = Count, fill = UCSC_RefGene_Group)) +
  geom_bar(position="fill", stat = "identity") +
  theme_minimal() +
  labs(title = "Distribution of Differentially Methylated CpGs",
       x = "Differentially Methylated CpGs", y = "Percentage", fill = "Region") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_flip() +
  scale_fill_brewer(palette="Set1")

p4
```

```{r, fig.width=12, fig.height=6}
p5 <- plot_grid(p3, p4, labels=c("b","c"), label_size=20, ncol=1, rel_heights = c(1, 1))
p0 <- plot_grid(p1, p5, labels=c("a",""), label_size=20, ncol=2, rel_widths = c(1.45, 1))

p0

save(p0, p1, p3, p4, p5, res, res_annot, res_sig, file = "figures/Fig3_plot_objects.RData")

grDevices::cairo_pdf("figures/Fig3.pdf", width=13, height=6)
print(p0)
dev.off()
```

## D+10 vs D-1

Use the same visualization workflow as Figure 3 for the `D+10 - D-1` contrast
from model 2. The model contrast is stored as `BEFORE_vs_AFTER`, while
`delta_beta` is calculated as `AFTER - BEFORE`, so the x-axis represents the
adjusted beta-value change from D-1 to D+10. No CpGs pass the combined FDR and
delta-beta threshold for this contrast, so only a volcano plot is generated as
an exploratory overview.

```{r}
res <- results_BEFORE_vs_AFTER

ann <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
res$CpG <- rownames(res)
res$logP <- -log10(res$adj.P.Val)
ann_sub <- ann[rownames(ann) %in% res$CpG, ]
ann_sub$CpG <- rownames(ann_sub)
res$significance <- "Not Sig"
res$significance[res$adj.P.Val < p_th & res$logFC > lfc_th & res$delta_beta < -delta_beta_th] <- "Hypermethylated"
res$significance[res$adj.P.Val < p_th & res$logFC < -lfc_th & res$delta_beta > delta_beta_th] <- "Hypomethylated"
res_annot <- as.data.frame(merge(res, ann_sub, by = "CpG", all.x = TRUE))
res_sig <- res_annot[res_annot$significance != "Not Sig", ]
```

```{r, fig.width=8, fig.height=6}
n_hyper <- sum(res$significance == "Hypermethylated", na.rm = TRUE)
n_hypo <- sum(res$significance == "Hypomethylated", na.rm = TRUE)

volcano <- ggplot(res, aes(x = delta_beta, y = logP)) +
  rasterise(geom_point(aes(color = significance), alpha = 0.7), dpi = 300) +
  scale_color_manual(values = c("Hypermethylated" = "#B2182B", 
                                "Hypomethylated" = "#2166AC", 
                                "Not Sig" = "grey75")) +
  geom_vline(xintercept = c(-delta_beta_th, delta_beta_th), linetype = "dashed", color = "black") +
  geom_hline(yintercept = -log10(p_th), linetype = "dashed", color = "black") +
  theme_minimal(base_size = 12) +
  labs(title = "D+10 vs D-1 Differential Methylation",
       x = "Beta value change",
       y = "-log10(FDR)",
       color = "Significance")

p1_d10 <- volcano +
  annotate("text", 
           x = max(res$delta_beta, na.rm = TRUE) * 0.35, 
           y = max(res$logP, na.rm = TRUE) * 0.85, 
           label = paste0("Hypermethylated: ", n_hyper, "\nHypomethylated: ", n_hypo),
           hjust = 0, size = 4, fontface = "italic")

p1_d10

save(p1_d10, res, res_annot, res_sig, file = "figures/volcano_D10_vs_D1_plot_objects.RData")
```

```{r}
grDevices::cairo_pdf("figures/volcano_D10_vs_D1.pdf", width = 6, height = 5)
print(p1_d10)
dev.off()
```
