---
title: "Differential methylation analysis"
format:
  html:
    theme: sandstone
    embed-resources: true
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, 
                      class.output="bg-light")
```

```{r, results=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(tidyr)
library(cowplot)
library(dplyr)
library(sesame)
library(sesameData)
library(readxl)
library(limma)
library(UpSetR)
library(minfi)
library(DMRcate)
library(missMethyl)
library(ggpubr)
library(knitr)
library(gridExtra)
library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
library(enrichplot)
source("utils.R")
theme_set(theme_minimal())
```

```{r, results=FALSE, message=FALSE, warning=FALSE}
proj_dir <- "~/public/SiyuanLuo/projects/fasting_clinics/publication"
setwd(proj_dir)
betass <- readRDS("./data/cleaned_beta_s.RDS")
betasp <- readRDS("./data/cleaned_beta_p.RDS")

meta <- read.csv(file="./data/meta_3timepoints.csv")
meta <- meta %>% filter(meta$Sample %in% colnames(betass))
meta$ID <- factor(meta$ID) 
```

```{r}
# calculate M-values for statistical analysis
# using all samples
M_values <- log2(betass/(1-betass))
mdsPlot(M_values, numPositions=1000,
        sampNames = NULL, sampGroups=meta$Timepoint)
```

Timepoint doesn’t separate the methylation data.

Due to the longitudinal design, use a mixed effect model.

Since we loaded the uncorrected data, we use the Date_Scan as a fixed effect for controlling batch effect.

# Compare between timepoints

::: {.callout-note}
Outlier removal:
* blood cells:
  * patient 28 is an outlier due to high thrombocytes. This is unlikely to affect the DNA composition.
  * patient 29 is an outlier due to low MCH (Mean Corpuscular Hemoglobin). This is unlikely to affect the DNA composition.
* liver function: patient 10, 2, 14
* energy metabolism: patient 10

Let’s remove patient 10 since the BMI is also extreme. 

I tried keeping vs removing 2 and 14 and in the end decide to keep them.
:::

```{r}
meta <- meta %>% filter(as.character(ID) !=10)
rownames(meta) <- meta$Sample
M_values <- M_values[, rownames(meta)]

meta <- meta[!is.na(meta$Sample),]
rownames(meta) <- meta$Sample
meta <- meta[colnames(M_values),]

# Create a model matrix for time points and any other covariates
meta$Timepoint2 <- factor(meta$Timepoint, levels = c("D-1", "D+10", "M+1"))
meta <- meta %>%
  mutate(Timepoint = case_when(
    Timepoint2 == "D-1"  ~ "BEFORE",
    Timepoint2 == "D+10" ~ "AFTER",
    Timepoint2 == "M+1"  ~ "FUP1",
    TRUE ~ Timepoint2
  )) %>%
  mutate(Timepoint = factor(Timepoint, levels = c("BEFORE","AFTER","FUP1"))) %>%
  mutate(ID_Patient = factor(ID_Patient))
meta$Date_Scan_recode <- factor(as.numeric(factor(meta$Date_Scan)))
meta$sex <- factor(meta$sex)

# Sanity check for covaraites
stopifnot(all(c("NEUTROPHILE","LYMPHOZYTEN") %in% colnames(meta)))
meta <- meta %>%
  mutate(
    NEUTROPHILE = as.numeric(NEUTROPHILE),
    LYMPHOZYTEN = as.numeric(LYMPHOZYTEN)
  )
stopifnot(!anyNA(meta$NEUTROPHILE), !anyNA(meta$LYMPHOZYTEN))
```

::: {.panel-tabset}
## Model 1
Model: Timepoint2 + Date_Scan + age + sex + (1 | ID_Patient)

### DMP
```{r}
design <- model.matrix(~ 0 + Timepoint + Date_Scan_recode + age + sex, data = meta)
```

```{r, eval=FALSE}
# Estimate correlations between repeated measures from the same individual
corfit <- duplicateCorrelation(M_values, design, block = meta$ID_Patient)
saveRDS(corfit, file="./data/corfit_design1_rm10.RDS")
```

```{r}
corfit <- readRDS(file="./data/corfit_design1_rm10.RDS")
corfit$consensus
```

```{r}
# fit the linear model
fit <- lmFit(M_values, design, block=meta$ID_Patient, correlation = corfit$consensus)
# Apply the contrasts and compute eBayes statistics
# Define pairwise contrasts
contrasts <- makeContrasts(
  BEFORE_vs_AFTER = TimepointBEFORE - TimepointAFTER,
  BEFORE_vs_FUP1 = TimepointBEFORE - TimepointFUP1,
  AFTER_vs_FUP1 = TimepointAFTER - TimepointFUP1,
  age = age,
  Female_vs_Male = sex1,
  levels = design
)
fit2 <- contrasts.fit(fit, contrasts)
fit2 <- eBayes(fit2)

# Extract results for each contrast
results_BEFORE_vs_AFTER <- topTable(fit2, coef = "BEFORE_vs_AFTER", number = Inf)
results_BEFORE_vs_FUP1 <- topTable(fit2, coef = "BEFORE_vs_FUP1", number = Inf)
results_AFTER_vs_FUP1 <- topTable(fit2, coef = "AFTER_vs_FUP1", number = Inf)
results_age <- topTable(fit2, coef = "age", number = Inf)
results_sex <- topTable(fit2, coef = "Female_vs_Male", number = Inf)

# look at the numbers of DM CpGs at FDR < 0.05
kable(summary(decideTests(fit2, adjust.method="fdr")))
```

```{r}
saveRDS(fit2, file="./data/fit_model1.RDS")
```

#### Check the effect size
The distribution of log-fold change:

```{r}
hist(results_BEFORE_vs_FUP1[results_BEFORE_vs_FUP1$adj.P.Val<0.05, "logFC"], 
     breaks=50,
     main = "Histogram of logFC for significant BEFORE vs FUP1 CpGs")
```

Check the overlap between DMPs between pairs of timepoints:

```{r}
dm <- list(BEFORE_vs_FUP1_up = rownames(results_BEFORE_vs_FUP1[results_BEFORE_vs_FUP1$adj.P.Val<0.05 & results_BEFORE_vs_FUP1$logFC>0,]),
           BEFORE_vs_FUP1_down = rownames(results_BEFORE_vs_FUP1[results_BEFORE_vs_FUP1$adj.P.Val<0.05 & results_BEFORE_vs_FUP1$logFC<0,]),
     AFTER_vs_FUP1_down = rownames(results_AFTER_vs_FUP1[results_AFTER_vs_FUP1$adj.P.Val<0.05 & results_AFTER_vs_FUP1$logFC<0,]),
    AFTER_vs_FUP1_up = rownames(results_AFTER_vs_FUP1[results_AFTER_vs_FUP1$adj.P.Val<0.05 & results_AFTER_vs_FUP1$logFC>0,]))

upset(fromList(dm), empty.intersections = TRUE)
```

Plotting the top few differentially methylated CpGs:

##### Age
```{r}
sapply(rownames(results_age)[1:6], function(cpg){
  plotCpg(betass[, meta$Sample], cpg=cpg, pheno=meta$age, ylab = "Beta values")
})
```

##### Sex
```{r}
sapply(rownames(results_sex)[1:6], function(cpg){
  plotCpg(betass[,meta$Sample], cpg=cpg, pheno=meta$Gender, ylab = "Beta values")
})
```

#### BEFORE vs FUP1

```{r}
sapply(rownames(results_BEFORE_vs_FUP1)[1:9], function(cpg){
  plotCpg(betass[,meta$Sample], cpg=cpg, pheno=meta$Timepoint, ylab = "Beta values")
})
```

```{r}
tmp <- data.frame(betass[rownames(results_BEFORE_vs_FUP1)[1:9], meta$Sample]) 
tmp$cpg <- rownames(tmp)
tmp <- tmp%>% pivot_longer(cols = -cpg, names_to = "Sample", values_to = "beta")
tmp$Sample <- sub("X", "", tmp$Sample)

merge(tmp, meta, by="Sample") %>% 
  filter(Timepoint %in% c("BEFORE", "FUP1")) %>%
  ggplot(aes(x=Timepoint, y=beta, color=Gender, group=ID_Patient)) + 
  geom_point() + 
  geom_line() + 
  facet_wrap(~cpg, scales = "free") + 
  labs(title="BEFORE vs FUP1")
```

### GO testing
#### BEFORE_vs_FUP1
```{r}
DMPs <- results_BEFORE_vs_FUP1
head(DMPs)

# Get the significant CpG sites at less than 5% FDR
sigCpGs <- rownames(DMPs[DMPs$adj.P.Val<0.05,])
# First 10 significant CpGs
sigCpGs[1:10]

# Total number of significant CpGs at 5% FDR
length(sigCpGs)

# Get all the CpG sites used in the analysis to form the background
all <- rownames(DMPs)
# Total number of CpG sites tested
length(all)

# Run enrichment - Can take a bit of time...
gst <- gometh(sig.cpg=sigCpGs, all.cpg=all)

# Top 10 GO categories
topGSA(gst, number=30)
```


### DMR
```{r}
myAnnotation <- cpg.annotate(object = M_values, datatype = "array", what = "M",
                                analysis.type= "differential", design = design,
                                contrasts = TRUE, cont.matrix = contrasts,
                                coef = "BEFORE_vs_FUP1", arraytype = "EPICv1", 
                             fdr = 0.08)
# Run DMRcate
DMRs <- dmrcate(myAnnotation, lambda=1000, C=2)
# convert the regions to annotated genomic ranges
sigRegions <- extractRanges(DMRs)
head(sigRegions)
```

```{r}
hist(sigRegions[sigRegions$no.cpgs >=5,]$maxdiff, 
     main="Histogram of BEFORE vs FUP1 DMR maxdiff",
     breaks = 50)
sigRegions[sigRegions$no.cpgs >=5 & sigRegions$maxdiff > 0.1,]
```

* PEX11B: Peroxisomal Biogenesis Factor 11 Beta
* GNRHR2: Gonadotropin Releasing Hormone Receptor 2 (Pseudogene)
* GPN2: ATP-Binding Domain 1 Family Member B
* GPATCH3: G-Patch Domain Containing 3
* LFNG: LFNG O-Fucosylpeptide 3-Beta-N-Acetylglucosaminyltransferase

```{r}
myAnnotation <- cpg.annotate(object = M_values, datatype = "array", what = "M",
                                analysis.type= "differential", design = design,
                                contrasts = TRUE, cont.matrix = contrasts,
                                coef = "AFTER_vs_FUP1", arraytype = "EPICv1", 
                             fdr = 0.1)
```

## Model 2
Model: Timepoint + Date_Scan + age + sex + ? + (1 | ID_Patient)

### Collect the candidate covariates
Collect the candidate covariates
* BMI_baseline_category/BMI_baseline
* blood cell composition: "NEUTROPHILE", "LYMPHOZYTEN"

### Check the covariates + timepoint
```{r}
# covariate sets
candidate_covariates <- list(
  "NEUTROPHILE + LYMPHOZYTEN",
  "BMI_baseline_category",
  "BMI_baseline"
)
names(candidate_covariates) <- unlist(candidate_covariates)

# Build formulas (no intercept, so timepoint has separate columns)
models <- lapply(candidate_covariates, function(cov_rhs) {
  as.formula(paste("~ 0 + Timepoint + Date_Scan_recode + age + sex +", cov_rhs))
})

# Helper to make filename-safe labels
safe_name <- function(x) gsub("[^A-Za-z0-9_]+", "_", x)

# Ensure factor coding is as expected (important for sex and timepoint column names)
# meta$Timepoint <- factor(meta$Timepoint, levels = c("BEFORE","AFTER","FUP1"))
# meta$sex <- relevel(factor(meta$sex), ref = "0")  # depends on your encoding

for (model_name in names(models)) {

  design <- model.matrix(models[[model_name]], data = meta)
  colnames(design) <- make.names(colnames(design))  # robust; avoids weird column names

  # --- correlation fit caching ---
  corfit_file <- file.path("./data", paste0("corfit_", safe_name(model_name), ".RDS"))
  if (!file.exists(corfit_file)) {
    corfit <- duplicateCorrelation(M_values, design, block = meta$ID_Patient)
    saveRDS(corfit, corfit_file)
  }
  corfit <- readRDS(corfit_file)

  # --- fit model ---
  fit <- lmFit(M_values, design, block = meta$ID_Patient, correlation = corfit$consensus)

  # --- define ONLY the timepoint contrasts ---
  # IMPORTANT: use the *actual* coefficient names in design
  # (after make.names). Usually these will be "TimepointBEFORE", etc.
  tp_cols <- grep("^Timepoint", colnames(design), value = TRUE)
  # sanity check
  stopifnot(all(c("TimepointBEFORE", "TimepointAFTER", "TimepointFUP1") %in% tp_cols))

  contrasts <- makeContrasts(
    BEFORE_vs_AFTER = TimepointBEFORE - TimepointAFTER,
    BEFORE_vs_FUP1  = TimepointBEFORE - TimepointFUP1,
    AFTER_vs_FUP1   = TimepointAFTER  - TimepointFUP1,
    levels = design
  )

  fit2 <- contrasts.fit(fit, contrasts)
  fit2 <- eBayes(fit2)

  # --- extract timepoint results ---
  res_BEFORE_vs_AFTER <- topTable(fit2, coef = "BEFORE_vs_AFTER", number = Inf)
  res_BEFORE_vs_FUP1  <- topTable(fit2, coef = "BEFORE_vs_FUP1",  number = Inf)
  res_AFTER_vs_FUP1   <- topTable(fit2, coef = "AFTER_vs_FUP1",   number = Inf)

  # --- extract covariate results from the *original* fit (not contrast fit) ---
  # Covariates are not contrasts; they are coefficients in `fit`.
  fit_cov <- eBayes(fit)

  res_age <- topTable(fit_cov, coef = "age", number = Inf)

  # sex coefficient name depends on coding. If sex is numeric 0/1, model.matrix gives "sex"
  # If sex is factor with 2 levels, it might be "sex1" or "sexFemale".
  sex_coef <- grep("^sex", colnames(design), value = TRUE)
  # prefer the non-timepoint sex coefficient (exclude any weird matches)
  sex_coef <- setdiff(sex_coef, tp_cols)
  # pick one if exactly one
  if (length(sex_coef) == 1) {
    res_sex <- topTable(fit_cov, coef = sex_coef, number = Inf)
  } else {
    res_sex <- NULL  # or stop() if you want strictness
  }

  # covariate(s) – could be 1 term or multiple (NEUTROPHILE + LYMPHOZYTEN)
  cov_terms <- all.vars(as.formula(paste("~", model_name)))
  cov_coefs <- cov_terms[cov_terms %in% colnames(design)]
  # If BMI_baseline_category expands to multiple dummy columns, capture them:
  cov_coefs <- c(cov_coefs, grep("^BMI_baseline_category", colnames(design), value = TRUE))
  cov_coefs <- unique(cov_coefs)

  res_cov <- lapply(cov_coefs, function(cc) topTable(fit_cov, coef = cc, number = Inf))
  names(res_cov) <- cov_coefs

  # --- summaries ---
  print(knitr::kable(summary(decideTests(fit2, adjust.method = "fdr"))))
  cat("\n\n<!-- -->\n\n")

  # --- save ---
  saveRDS(
    list(
      fit2 = fit2,
      fit_cov = fit_cov,
      results = list(
        BEFORE_vs_AFTER = res_BEFORE_vs_AFTER,
        BEFORE_vs_FUP1  = res_BEFORE_vs_FUP1,
        AFTER_vs_FUP1   = res_AFTER_vs_FUP1,
        age = res_age,
        sex = res_sex,
        covariates = res_cov
      ),
      design_colnames = colnames(design),
      model = models[[model_name]]
    ),
    file = file.path("./data", paste0("fit_model2_", safe_name(model_name), ".RDS"))
  )
}
```

Some observations

- BMI_baseline do have effects on the methylation levels. Since it shouldn’t confound the time effect, it should be added as an interaction term.
- NEUTROPHILE/LYMPHOZYTEN do have strong effects on the methylation levels, but it is independent of the time effect (from previous correlation analysis), suggesting that it is a strong confounding factor. It should be adjusted for. This can be validated via analyzing the DM CpGs associated to NEUTROPHILE/LYMPHOZYTEN.

## Model 3
Model: Timepoint * BMI_baseline + Date_Scan + age + sex + (1 | ID_Patient)

When adding a continuous variable in the interaction term, it is important to model the intercept. For categorical variable only, no interaction term is fine.

```{r}
meta$BMI_baseline_centered <- scale(meta$BMI_baseline, center = TRUE, scale = FALSE)
design <-model.matrix(~ Timepoint*BMI_baseline_centered + Date_Scan_recode + age + sex, data = meta)
corfit_file <- paste0("./data/corfit_design3_interaction.RDS")
  if (!file.exists(corfit_file)) {
    corfit <- duplicateCorrelation(M_values, design, block = meta$ID_Patient)
    saveRDS(corfit, file = corfit_file)
  }
  corfit <- readRDS(corfit_file)
  # fit the linear model
colnames(design) <- make.names(colnames(design))
fit <- lmFit(M_values, design, block=meta$ID_Patient, correlation = corfit$consensus)
# Define pairwise contrasts
contrasts <- makeContrasts(
  age = age,
  Female_vs_Male = sex1,
  BMI_baseline = BMI_baseline_centered,
  BEFORE_vs_AFTER = TimepointAFTER,
  BEFORE_vs_FUP1 = TimepointFUP1,
  AFTER_vs_FUP1 = TimepointAFTER - TimepointFUP1,
  BMI_slope_AFTER_vs_BEFORE = TimepointAFTER.BMI_baseline_centered,  # Does BMI modify T2 vs. T1?
  BMI_slope_FUP1_vs_BEFORE = TimepointFUP1.BMI_baseline_centered,  # Does BMI modify T3 vs. T1?
  BMI_slope_FUP1_vs_AFTER = TimepointFUP1.BMI_baseline_centered - TimepointAFTER.BMI_baseline_centered,  # Does BMI modify T3 vs. T2?
  levels = colnames(design)
)
fit2 <- contrasts.fit(fit, contrasts)
fit2 <- eBayes(fit2)
# look at the numbers of DM CpGs at FDR < 0.05
kable(summary(decideTests(fit2, adjust.method="fdr")))
saveRDS(fit2, file="./data/fit_model3_interaction.RDS")
```
:::

# BEFORE vs FUP1
## Compare CpGs in model 1 and model 2
```{r}
# model 1
design <- model.matrix(~ 0 + Timepoint + Date_Scan_recode + age + sex, data = meta)
corfit <- readRDS(file="corfit_design1_rm10_2_14.RDS")
fit <- lmFit(M_values, design, block=meta$ID_Patient, correlation = corfit$consensus)
contrasts <- makeContrasts(
  AFTER_vs_FUP1 = TimepointAFTER - TimepointFUP1,
  levels = design
)
fit2 <- contrasts.fit(fit, contrasts)
fit2 <- eBayes(fit2)

# Extract results for each contrast
results_AFTER_vs_FUP1 <- topTable(fit2, coef = "AFTER_vs_FUP1", number = Inf)
summary(decideTests(fit2, adjust.method="fdr"))
```

```{r}
p_th <- 0.05
lfc_th <- 0
sigCpGsModel1 <- rownames(results_AFTER_vs_FUP1[results_AFTER_vs_FUP1$adj.P.Val<p_th,])
dm <- list(model1_up = rownames(results_AFTER_vs_FUP1[results_AFTER_vs_FUP1$adj.P.Val<p_th & results_AFTER_vs_FUP1$logFC> lfc_th,]),
           model1_down = rownames(results_AFTER_vs_FUP1[results_AFTER_vs_FUP1$adj.P.Val<p_th & results_AFTER_vs_FUP1$logFC< -lfc_th,]))
```

```{r}
# model 2
design <- model.matrix(~ 0 + Timepoint + Date_Scan_recode + age + sex + NEUTROPHILE + LYMPHOZYTEN, data = meta)
corfit_file <- paste0("corfit_", "NEUTROPHILE_LYMPHOZYTEN", ".RDS")
if (!file.exists(corfit_file)) {
  corfit <- duplicateCorrelation(M_values, design, block = meta$ID_Patient)
  saveRDS(corfit, file = corfit_file)
}
corfit <- readRDS(corfit_file)
fit <- lmFit(M_values, design, block=meta$ID_Patient, correlation = corfit$consensus)
contrasts <- makeContrasts(
  BEFORE_vs_AFTER = TimepointBEFORE - TimepointAFTER,
  BEFORE_vs_FUP1 = TimepointBEFORE - TimepointFUP1,
  AFTER_vs_FUP1 = TimepointAFTER - TimepointFUP1,
  NEUTROPHILE = NEUTROPHILE,
  LYMPHOZYTEN = LYMPHOZYTEN,
  levels = design
)
fit2 <- contrasts.fit(fit, contrasts)
fit2 <- eBayes(fit2)

# Extract results for each contrast
results_BEFORE_vs_AFTER <- topTable(fit2, coef = "BEFORE_vs_AFTER", number = Inf)
results_BEFORE_vs_FUP1 <- topTable(fit2, coef = "BEFORE_vs_FUP1", number = Inf)
results_AFTER_vs_FUP1 <- topTable(fit2, coef = "AFTER_vs_FUP1", number = Inf)
results_NEUTROPHILE/LYMPHOZYTEN <- topTable(fit2, coef = "NEUTROPHILE/LYMPHOZYTEN", number = Inf)
summary(decideTests(fit2, adjust.method="fdr"))
```

```{r}
sigCpGsModel2 <- rownames(results_AFTER_vs_FUP1[results_AFTER_vs_FUP1$adj.P.Val<p_th,])

dm$model2_up <- rownames(results_AFTER_vs_FUP1[results_AFTER_vs_FUP1$adj.P.Val<p_th & results_AFTER_vs_FUP1$logFC> lfc_th,])
dm$model2_down <- rownames(results_AFTER_vs_FUP1[results_AFTER_vs_FUP1$adj.P.Val<p_th & results_AFTER_vs_FUP1$logFC< -lfc_th,])
dm$NEUTROPHILE/LYMPHOZYTEN_up <- rownames(results_NEUTROPHILE/LYMPHOZYTEN[results_NEUTROPHILE/LYMPHOZYTEN$adj.P.Val<p_th & results_NEUTROPHILE/LYMPHOZYTEN$logFC> lfc_th,])
dm$NEUTROPHILE/LYMPHOZYTEN_down <- rownames(results_NEUTROPHILE/LYMPHOZYTEN[results_NEUTROPHILE/LYMPHOZYTEN$adj.P.Val<p_th & results_NEUTROPHILE/LYMPHOZYTEN$logFC< -lfc_th,])

upset(fromList(dm[c("model1_up", "model1_down","model2_up", "model2_down")]))
upset(fromList(dm[c("model1_up", "model2_up", "NEUTROPHILE/LYMPHOZYTEN_up")]))
upset(fromList(dm[c("model1_down", "model2_down", "NEUTROPHILE/LYMPHOZYTEN_down")]))
```

So it is true that almost all DM CpGs in model 2 are also present in model 1.

## GO Testing for model 2
```{r}
sigCpGsModeldiff <- sigCpGsModel1[!sigCpGsModel1 %in% sigCpGsModel2]
sigCpGs <- sigCpGsModel2
check <- getMappedEntrezIDs(sig.cpg = sigCpGs)
length(check$sig.eg)

p_th <- 0.05
lfc_th <- 0.3
sigCpGs <- rownames(results_AFTER_vs_FUP1[results_AFTER_vs_FUP1$adj.P.Val<p_th & abs(results_AFTER_vs_FUP1$logFC)>lfc_th,])
check <- getMappedEntrezIDs(sig.cpg = sigCpGs)
length(check$sig.eg)

```

::: {.panel-tabset}
### GO
```{r, fig.width=5, fig.height=10}
gst <- gometh(sig.cpg=sigCpGs, all.cpg=rownames(results_AFTER_vs_FUP1), 
              collection="GO", array.type = c("EPIC"),
              plot.bias=TRUE, sig.genes = FALSE)

topGSA(gst, n=20)

gst2compareClusterResult <- function(gst, cluster, n_DE_genes, description="Description"){
  gst$Cluster <- cluster
  df <- data.frame(
  Cluster = gst$Cluster,  
  ID = rownames(gst),  # GO term ID
  Description = gst[,description],  
  GeneRatio = paste0(gst$DE, "/", n_DE_genes),  # Ratio of DE genes
  pvalue = gst$P.DE,
  p.adjust = gst$FDR,
  qvalue = gst$FDR,  # No qvalue in gometh, so reuse FDR
  Count = gst$DE)
  
  df <- df[order(sapply(df$GeneRatio, function(x){eval(parse(text=x))}), decreasing = TRUE),]
# Convert to compareClusterResult object
  compare_cluster_obj <- new(
    "compareClusterResult",
    compareClusterResult = df,
    fun = "gometh"
  )
  return(compare_cluster_obj)
}

compare_cluster_obj <- gst2compareClusterResult(gst, 
                         cluster="AFTER_vs_FUP1", 
                         n_DE_genes=length(getMappedEntrezIDs(sig.cpg = sigCpGs)$sig.eg),
                         description="TERM")

enrichplot::dotplot(compare_cluster_obj, 
                    showCategory = 20, 
                    x = "GeneRatio", 
                    color = "p.adjust", 
                    size="count", 
                    by = "GeneRatio")
```

### KEGG
```{r, fig.width=5, fig.height=10}
gst.kegg <- gometh(sig.cpg=sigCpGs, all.cpg=rownames(results_AFTER_vs_FUP1), 
                   array.type = c("EPIC"), collection="KEGG",
                   sig.genes = TRUE)

topGSA(gst.kegg, n=20)

compare_cluster_obj <- gst2compareClusterResult(gst.kegg, 
                         cluster="AFTER_vs_FUP1", 
                         n_DE_genes=length(getMappedEntrezIDs(sig.cpg = sigCpGs)$sig.eg),
                         description="Description")

enrichplot::dotplot(compare_cluster_obj, 
                    showCategory = 20, 
                    x = "GeneRatio", 
                    color = "p.adjust", 
                    size="count", 
                    by = "GeneRatio")
```

### KEGG promoter
```{r}
gst.kegg.prom <- gometh(sig.cpg=sigCpGs, all.cpg=rownames(results_AFTER_vs_FUP1), array.type = c("EPIC"),
                        collection="KEGG", genomic.features = c("TSS200",
                                                                "TSS1500",
                                                                "1stExon"))
topGSA(gst.kegg.prom, n=20)

compare_cluster_obj <- gst2compareClusterResult(gst.kegg.prom, 
                         cluster="AFTER_vs_FUP1", 
                         n_DE_genes=length(getMappedEntrezIDs(sig.cpg = sigCpGs)$sig.eg),
                         description="Description")

enrichplot::dotplot(compare_cluster_obj, 
                    showCategory = 20, 
                    x = "GeneRatio", 
                    color = "p.adjust", 
                    size="count", 
                    by = "GeneRatio")
```

### KEGG gene body
```{r}
gst.kegg.body <- gometh(sig.cpg=sigCpGs, all.cpg=rownames(results_AFTER_vs_FUP1), 
                        collection="KEGG", array.type = c("EPIC"), genomic.features = c("Body"))
topGSA(gst.kegg.body, n=20)

gst.kegg.body <- gometh(sig.cpg=sigCpGs, all.cpg=rownames(results_AFTER_vs_FUP1), 
                        collection="KEGG", array.type = c("EPIC"), genomic.features = c("Body"), sig.genes = TRUE)
topGSA(gst.kegg.body, n=10)
```

:::

```{r}
sigCpGsModel2 <- rownames(results_BEFORE_vs_FUP1[results_BEFORE_vs_FUP1$adj.P.Val<p_th,])
```

# Visualization
```{r}
for (cov in candidate_covariates) {
  p1 <- meta %>% filter(Timepoint %in% c("BEFORE", "AFTER")) %>%
    ggpaired(x = "Timepoint", y = cov, color = "Timepoint", 
           line.color = "gray", line.size = 0.4, 
           palette = c("#00AFBB", "#E7B800")) +
    labs(x = "Time point", y=cov) +
    ylim(min(meta[,cov]), max(meta[,cov])) +
    stat_compare_means(paired = TRUE, label.y = max(meta[,cov]))
  
  p2 <- meta %>% filter(Timepoint %in% c("BEFORE", "FUP1")) %>%
  ggpaired(x = "Timepoint", y = cov, color = "Timepoint", 
           line.color = "gray", line.size = 0.4, 
           palette = c("#00AFBB", "#FC4E07")) +
    labs(x = "Time point", y=cov) +
    ylim(min(meta[,cov]), max(meta[,cov])) +
    stat_compare_means(paired = TRUE,  label.y = max(meta[,cov]))
  
  p3 <- meta %>% filter(Timepoint %in% c("AFTER", "FUP1")) %>%
  ggpaired(x = "Timepoint", y = cov, color = "Timepoint", 
           line.color = "gray", line.size = 0.4, 
           palette = c("#E7B800", "#FC4E07")) +
    labs(x = "Time point", y=cov) +
   ylim(min(meta[,cov]), max(meta[,cov])) +
    stat_compare_means(paired = TRUE,  label.y = max(meta[,cov]))
  
  grid.arrange(p1, p2, p3, ncol = 3)
}
```

