---
title: "Clock QC and age-acceleration screening"
subtitle: "Clock-vs-age plots and RAA associations with maximum weight loss"
abstract: |
  This notebook performs quality-control visualization and exploratory
  weight-loss association screening for epigenetic clock residual age
  acceleration (RAA). It joins `data/AllClocks_Merged.csv` with three-timepoint
  metadata, estimates RAA for 12 clocks, and checks each clock against
  chronological age using mixed models with participant random intercepts.
  Repeated-measures models then test associations between RAA and maximum
  weight-loss percentage after adjustment for sex, timepoint, lymphocyte
  proportion, and baseline BMI. Greater maximum weight loss is nominally
  associated with lower Horvath 1, Horvath 2, and PhenoAge RAA, with the
  strongest and most stable evidence for Horvath 2 and PhenoAge after baseline
  BMI adjustment. Baseline-only sensitivity analyses show the same negative
  pattern for Horvath 2 and PhenoAge, while weight-loss-by-baseline-BMI
  interactions are not clearly supported. Outputs include clock-vs-age plots,
  repeated-measures and baseline sensitivity tables, interaction tests, and
  RAA-weight-loss plot PDFs.
format:
  html:
    theme: sandstone
    embed-resources: true
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, 
                      class.output="bg-light")
```

# Load 
```{r, results=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(patchwork)
library(tidyr)
library(readr)
library(dplyr)
library(purrr)
library(lme4)
library(lmerTest)
library(ggeffects)
source("utils.R")
```

```{r config}
# ============================================================================ #
# Configuration: Clock visualization
# ============================================================================ #
FILE_OUTPUT <- "data/AllClocks_Merged.csv"
df_age <- read_csv(FILE_OUTPUT)
meta <- read_csv("data/meta_3timepoints.csv")

df_age <- df_age %>%
  left_join(meta %>% select(Sample, ID, Timepoint,
                            LYMPHOZYTEN, max_weight_loss_perc, BMI), 
            by = "Sample") %>%
  mutate(Timepoint = factor(Timepoint, levels = c("D-1", "D+10", "M+1"))) %>%
  group_by(ID) %>%
  mutate(
    baseline_BMI = if (any(Timepoint == "D-1" & !is.na(BMI))) {
      dplyr::first(BMI[Timepoint == "D-1" & !is.na(BMI)])
    } else {
      NA_real_
    }
  ) %>%
  ungroup()

# Define clocks to visualize
clocks_config <- list(
  PCHorvath1 = "PC Horvath 1",
  PCHorvath2 = "PC Horvath 2",
  PCHannum = "PC Hannum",
  PCPhenoAge = "PC PhenoAge",
  PCGrimAge = "PC GrimAge",
  DNAmGrimAge2 = "GrimAge v2",
  PCDNAmTL = "PC DNAmTL",
  epi_age_DunedinPACE = "DunedinPACE",
  SystemsAge = "SystemsAge",
  AdaptAge = "AdaptAge",
  DamAge = "DamAge",
  CausAge = "CausAge"
)

# Age acceleration display names (for RAA plots)
ageaccel_titles <- list(
  PCHorvath1_RAA = "Horvath 1 RAA",
  PCHorvath2_RAA = "Horvath 2 RAA",
  PCHannum_RAA = "Hannum RAA",
  PCPhenoAge_RAA = "PhenoAge RAA",
  PCGrimAge_RAA = "GrimAge RAA",
  DNAmGrimAge2_RAA = "GrimAge v2 RAA",
  PCDNAmTL_RAA = "DNAmTL RAA",
  epi_age_DunedinPACE_RAA = "DunedinPACE RAA",
  SystemsAge_RAA = "SystemsAge RAA",
  AdaptAge_RAA = "AdaptAge RAA",
  DamAge_RAA = "DamAge RAA",
  CausAge_RAA = "CausAge RAA"
)

source("~/public/SiyuanLuo/projects/LongAge/R/CalAgeAccel.R")
ageaccel_clocks <- names(clocks_config)
df_age_accel <- CalResidAgeAccel(df_age, clock_cols=ageaccel_clocks, age_col = "Age")$data_wide
```

# Visualization functions

```{r plot-function}
plot_against_age_lmm <- function(
  df,
  col_name,
  title_name,
  age_col = "Age",
  id_col = "ID",
  time_col = "Timepoint",
  line_mode = c("avg_timepoint", "reference"),
  reference_timepoint = "D-1"
) {
  line_mode <- match.arg(line_mode)

  # Packages needed:
  # lme4, performance, dplyr, ggplot2
  stopifnot(all(c(col_name, age_col, id_col, time_col) %in% colnames(df)))

  # Ensure timepoint is a factor (preserve order if already set)
  if (!is.factor(df[[time_col]])) {
    df[[time_col]] <- factor(df[[time_col]])
  }

  # ---- Fit LMM ----
  fml <- as.formula(
    paste0(col_name, " ~ ", age_col, " + ", time_col, " + (1|", id_col, ")")
  )
  fit <- lme4::lmer(fml, data = df, REML = FALSE)

  # Marginal R2 (fixed effects only)
  r2 <- performance::r2_nakagawa(fit)$R2_marginal
  r2_label <- paste0("R\u00B2 = ", formatC(r2, format = "f", digits = 3))

  # ---- Create prediction line (fixed effects only, consistent with LMM) ----
  age_grid <- seq(
    min(df[[age_col]], na.rm = TRUE),
    max(df[[age_col]], na.rm = TRUE),
    length.out = 100
  )
  tp_levels <- levels(df[[time_col]])

  if (line_mode == "reference") {
    if (!reference_timepoint %in% tp_levels) {
      stop("reference_timepoint not found in levels of time_col.")
    }
    newdat <- data.frame(
      tmp_age = age_grid,
      tmp_tp = factor(reference_timepoint, levels = tp_levels),
      tmp_id = NA
    )
    names(newdat) <- c(age_col, time_col, id_col)

    newdat$yhat <- predict(fit, newdata = newdat, re.form = NA)
    pred_df <- newdat[, c(age_col, "yhat")]
  } else {
    # Average equally across timepoints to get ONE line through all points
    newdat <- expand.grid(
      tmp_age = age_grid,
      tmp_tp = factor(tp_levels, levels = tp_levels)
    )
    newdat$tmp_id <- NA
    names(newdat) <- c(age_col, time_col, id_col)

    newdat$yhat <- predict(fit, newdata = newdat, re.form = NA)

    pred_df <- dplyr::as_tibble(newdat) %>%
      dplyr::group_by(.data[[age_col]]) %>%
      dplyr::summarise(yhat = mean(yhat), .groups = "drop")
  }

  # ---- Plot ----
  p <- ggplot2::ggplot(
    df,
    ggplot2::aes(x = .data[[age_col]], y = .data[[col_name]], color = .data[[time_col]])
  ) +
    ggplot2::geom_point(size = 2, alpha = 0.7) +
    ggplot2::geom_line(
      data = pred_df,
      ggplot2::aes(x = .data[[age_col]], y = yhat),
      inherit.aes = FALSE,
      linetype = "dashed",
      color = "grey30",
      linewidth = 0.8
    ) +
    ggplot2::annotate(
      "text",
      x = Inf, y = Inf,
      label = r2_label,
      hjust = 1.1, vjust = 1.5,
      size = 4.5, color = "black"
    ) +
    ggplot2::labs(
      x = "Chronological Age (years)",
      y = title_name,
      title = title_name,
      color = "Time point"
    ) +
    ggplot2::theme_minimal() +
    ggplot2::theme(
      plot.title = ggplot2::element_text(hjust = 0.5, size = 12, face = "bold"),
      axis.title = ggplot2::element_text(size = 10),
      legend.position = "right"
    )

  return(p)
}
```

```{r}
extract_weight_loss_term <- function(model, outcome, model_name, cov = "max_weight_loss_perc") {
  coef_df <- as.data.frame(summary(model)$coefficients)
  model_n <- nrow(model.frame(model))

  coef_df %>%
    tibble::rownames_to_column("term") %>%
    filter(term == cov) %>%
    transmute(
      outcome = outcome,
      model = model_name,
      beta = Estimate,
      SE = `Std. Error`,
      df = df,
      t_value = `t value`,
      p_value = `Pr(>|t|)`,
      n = model_n
    )
}

extract_model_terms <- function(model,
                                outcome,
                                model_name,
                                terms,
                                response = outcome) {
  coef_df <- as.data.frame(summary(model)$coefficients)
  model_n <- nrow(model.frame(model))
  p_col <- intersect(c("Pr(>|t|)", "Pr(>|z|)"), colnames(coef_df))[1]
  df_col <- intersect(c("df"), colnames(coef_df))[1]
  model_df <- if (!is.na(df_col)) NULL else df.residual(model)

  coef_df %>%
    tibble::rownames_to_column("term") %>%
    filter(term %in% terms) %>%
    transmute(
      outcome = outcome,
      response = response,
      model = model_name,
      term,
      beta = Estimate,
      SE = `Std. Error`,
      df = if (!is.na(df_col)) .data[[df_col]] else model_df,
      t_value = `t value`,
      p_value = .data[[p_col]],
      n = model_n
    )
}

fit_ageaccel_models <- function(df,
                                outcome,
                                cov = "max_weight_loss_perc",
                                bmi_cov = "baseline_BMI") {
  needed <- c(
    outcome, "Female", "Timepoint", "LYMPHOZYTEN",
    cov, bmi_cov, "ID"
  )

  df_sub <- df %>%
    select(all_of(needed)) %>%
    drop_na()

  if (nrow(df_sub) < 5) return(NULL)

  fixed_m3 <- paste(c("Female", "Timepoint", "LYMPHOZYTEN", cov), collapse = " + ")
  fixed_m4 <- paste(c("Female", "Timepoint", "LYMPHOZYTEN", bmi_cov, cov), collapse = " + ")
  formula_m3 <- as.formula(paste(outcome, "~", fixed_m3, "+ (1 | ID)"))
  formula_m4 <- as.formula(paste(outcome, "~", fixed_m4, "+ (1 | ID)"))

  list(
    data = df_sub,
    model3 = lmer(formula_m3, data = df_sub),
    model4 = lmer(formula_m4, data = df_sub)
  )
}

compare_ageaccel_models <- function(df,
                                    outcome,
                                    cov = "max_weight_loss_perc",
                                    bmi_cov = "baseline_BMI") {
  fits <- fit_ageaccel_models(df, outcome, cov = cov, bmi_cov = bmi_cov)
  if (is.null(fits)) return(NULL)

  bind_rows(
    extract_weight_loss_term(fits$model3, outcome, "Model 3: lymphocyte-adjusted", cov = cov),
    extract_weight_loss_term(fits$model4, outcome, "Model 4: + baseline BMI", cov = cov)
  )
}

fit_ageaccel_interaction_model <- function(df,
                                           outcome,
                                           cov = "max_weight_loss_perc",
                                           bmi_cov = "baseline_BMI") {
  needed <- c(
    outcome, "Female", "Timepoint", "LYMPHOZYTEN",
    cov, bmi_cov, "ID"
  )

  df_sub <- df %>%
    select(all_of(needed)) %>%
    drop_na()

  if (nrow(df_sub) < 5) return(NULL)

  formula_m5 <- as.formula(
    paste0(
      outcome, " ~ ", cov, " * ", bmi_cov,
      " + Timepoint + Female + LYMPHOZYTEN + (1 | ID)"
    )
  )

  lmer(formula_m5, data = df_sub)
}

compare_ageaccel_interaction_model <- function(df,
                                               outcome,
                                               cov = "max_weight_loss_perc",
                                               bmi_cov = "baseline_BMI") {
  model <- fit_ageaccel_interaction_model(
    df,
    outcome,
    cov = cov,
    bmi_cov = bmi_cov
  )
  if (is.null(model)) return(NULL)

  extract_model_terms(
    model,
    outcome = outcome,
    model_name = "Model 5: weight loss x baseline BMI",
    terms = c(cov, bmi_cov, paste0(cov, ":", bmi_cov)),
    response = outcome
  )
}

fit_baseline_model6 <- function(df,
                                outcome,
                                cov = "max_weight_loss_perc",
                                bmi_cov = "baseline_BMI") {
  needed <- c(outcome, cov, bmi_cov, "Female", "Timepoint")

  df_sub <- df %>%
    filter(Timepoint == "D-1") %>%
    select(all_of(needed)) %>%
    drop_na()

  if (nrow(df_sub) < 5) return(NULL)

  formula_m6 <- as.formula(
    paste0(outcome, " ~ ", cov, " + ", bmi_cov, " + Female")
  )

  lm(formula_m6, data = df_sub)
}

compare_baseline_model6 <- function(df,
                                    outcome,
                                    cov = "max_weight_loss_perc",
                                    bmi_cov = "baseline_BMI") {
  model <- fit_baseline_model6(df, outcome, cov = cov, bmi_cov = bmi_cov)
  if (is.null(model)) return(NULL)

  extract_model_terms(
    model,
    outcome = outcome,
    model_name = "Model 6: baseline RAA outcome",
    terms = c(cov, bmi_cov),
    response = outcome
  )
}

fit_baseline_model7 <- function(df,
                                outcome,
                                cov = "max_weight_loss_perc",
                                bmi_cov = "baseline_BMI") {
  needed <- c(outcome, cov, bmi_cov, "Female", "Timepoint")

  df_sub <- df %>%
    filter(Timepoint == "D-1") %>%
    select(all_of(needed)) %>%
    drop_na()

  if (nrow(df_sub) < 5) return(NULL)

  formula_m7 <- as.formula(
    paste0(cov, " ~ ", outcome, " + ", bmi_cov, " + Female")
  )

  lm(formula_m7, data = df_sub)
}

compare_baseline_model7 <- function(df,
                                    outcome,
                                    cov = "max_weight_loss_perc",
                                    bmi_cov = "baseline_BMI") {
  model <- fit_baseline_model7(df, outcome, cov = cov, bmi_cov = bmi_cov)
  if (is.null(model)) return(NULL)

  extract_model_terms(
    model,
    outcome = outcome,
    model_name = "Model 7: weight loss outcome",
    terms = c(outcome, bmi_cov),
    response = cov
  )
}

# Helper to fit repeated-measures age-acceleration models and return a plot for one clock
fit_ageaccel <- function(df, outcome, cov="max_weight_loss_perc",
                         xlab="Max Weight Loss Percentage",
                         ylab=NULL, title=NULL,
                         model_version = c("model3", "model4")) {
  model_version <- match.arg(model_version)

  # Use outcome name as default if not provided
  if (is.null(ylab)) ylab <- outcome
  if (is.null(title)) title <- outcome
  
  fits <- fit_ageaccel_models(df, outcome, cov = cov)
  if (is.null(fits)) return(NULL)

  df_sub <- fits$data
  model <- fits[[model_version]]
  preds <- ggpredict(model, terms = c(cov, "Timepoint"))
  coef_est <- summary(model)$coefficients[cov, "Estimate"]
  pval <- summary(model)$coefficients[cov, "Pr(>|t|)"]
  label_text <- paste0(
    "beta = ", signif(coef_est, 3), ", ",
    ifelse(pval < 0.001, "p < 0.001", paste0("p = ", signif(pval, 3)))
  )

  ggplot(df_sub, aes(x = .data[[cov]], y = .data[[outcome]])) +
    geom_point(aes(color = Timepoint), alpha = 0.6) +
    geom_line(aes(group = ID), alpha = 0.3) +
    geom_ribbon(data = preds, aes(x = x, ymin = conf.low, ymax = conf.high, fill = group),
                inherit.aes = FALSE, alpha = 0.2) +
    geom_line(data = preds, aes(x = x, y = predicted, color = group),
              inherit.aes = FALSE, linewidth = 1.2) +
    theme_minimal() +
    labs(
      x = xlab,
      y = ylab,
      title = title
    ) +
    annotate(
      "text", x = Inf, y = Inf, label = label_text,
      hjust = 1.1, vjust = 1.5, size = 4.5, fontface = "italic"
    ) +
    scale_color_brewer(palette = "Dark2", name = "Timepoint") +
    scale_fill_brewer(palette = "Dark2", name = "Timepoint")
}
```

# Epigenetic Age vs Chronological Age

```{r clock-plots, fig.width=16, fig.height=12}
# Generate plots for each clock
plots_list <- map2(
  names(clocks_config),
  unname(clocks_config),
  ~ plot_against_age_lmm(
    df = df_age,
    col_name = .x,
    title_name = .y,
    age_col = "Age"
  )
)

# Combine plots
pf_clocks <- wrap_plots(plots_list, ncol = 4) +
  plot_layout(guides = "collect") +
  # plot_annotation(title = "Epigenetic Clocks vs. Chronological Age") &
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))

pf_clocks
```

```{r save-clocks}
# Save combined plot
ggsave(
  "figures/clocks_vs_age.pdf",
  pf_clocks,
  width = 12,
  height = 8
)
```

# EpiAgeAccel vs Weight Loss

## Models 3 and 4: repeated-measures weight-loss association

Model 3 is a repeated-measures linear mixed model that tests whether each
residual age-acceleration (RAA) measure is associated with maximum weight-loss
percentage after adjustment for sex, timepoint, and lymphocyte proportion, with
a participant-specific random intercept:
`RAA ~ Female + Timepoint + LYMPHOZYTEN + max_weight_loss_perc + (1 | ID)`.

Model 4 uses the same repeated-measures structure as Model 3 and adds baseline
BMI, defined as each participant's D-1 BMI, as an additional covariate:
`RAA ~ Female + Timepoint + LYMPHOZYTEN + baseline_BMI + max_weight_loss_perc + (1 | ID)`.

```{r ageaccel-model3-model4, warning=FALSE, message=FALSE}
ageaccel_model_comparison <- map_dfr(
  names(ageaccel_titles),
  ~ compare_ageaccel_models(df_age_accel, .x)
) %>%
  mutate(
    clock = unname(unlist(ageaccel_titles[outcome])),
    p_value = signif(p_value, 3),
    across(c(beta, SE, df, t_value), ~ round(.x, 3))
  )

ageaccel_model_comparison_wide <- ageaccel_model_comparison %>%
  select(clock, outcome, model, beta, SE, df, p_value, n) %>%
  pivot_wider(
    names_from = model,
    values_from = c(beta, SE, df, p_value, n)
  ) %>%
  transmute(
    clock,
    outcome,
    beta_model3 = `beta_Model 3: lymphocyte-adjusted`,
    p_model3 = `p_value_Model 3: lymphocyte-adjusted`,
    n_model3 = `n_Model 3: lymphocyte-adjusted`,
    beta_model4 = `beta_Model 4: + baseline BMI`,
    p_model4 = `p_value_Model 4: + baseline BMI`,
    n_model4 = `n_Model 4: + baseline BMI`,
    beta_change_after_baseline_BMI =
      `beta_Model 4: + baseline BMI` -
      `beta_Model 3: lymphocyte-adjusted`
  ) %>%
  mutate(beta_change_after_baseline_BMI = round(beta_change_after_baseline_BMI, 3))

knitr::kable(
  ageaccel_model_comparison_wide,
  caption = "Comparison of max-weight-loss coefficient before and after adding baseline BMI",
  digits = 3
)

write_csv(
  ageaccel_model_comparison,
  "figures/AgeAccel_weight_loss_model3_model4_comparison_long.csv"
)

write_csv(
  ageaccel_model_comparison_wide,
  "figures/AgeAccel_weight_loss_model3_model4_comparison_wide.csv"
)
```

## Model 5: weight-loss by baseline BMI interaction

Model 5 is a repeated-measures linear mixed model that evaluates whether baseline BMI
modifies the association between maximum weight-loss percentage and each RAA
measure. It includes the weight-loss-by-baseline-BMI interaction, timepoint, sex,
lymphocyte proportion, and a participant-specific random intercept:
`RAA ~ max_weight_loss_perc * baseline_BMI + Timepoint + Female + LYMPHOZYTEN + (1 | ID)`.

```{r ageaccel-model5-interaction, warning=FALSE, message=FALSE}

ageaccel_model5_interaction <- map_dfr(
  names(ageaccel_titles),
  ~ compare_ageaccel_interaction_model(df_age_accel, .x)
) %>%
  mutate(
    clock = unname(unlist(ageaccel_titles[outcome])),
    p_value = signif(p_value, 3),
    across(c(beta, SE, df, t_value), ~ round(.x, 3))
  ) %>%
  select(clock, outcome, response, model, term, beta, SE, df, t_value, p_value, n)

knitr::kable(
  ageaccel_model5_interaction,
  caption = "Model 5 interaction analysis: RAA ~ weight loss x baseline BMI + timepoint + gender + lymphocytes + (1 | ID)",
  digits = 3
)

write_csv(
  ageaccel_model5_interaction,
  "figures/AgeAccel_weight_loss_model5_interaction.csv"
)
```

## Model 6: baseline-only sensitivity analysis

Model 6 is a baseline-only linear model restricted to D-1 samples. It tests the
cross-sectional association between baseline RAA and maximum weight-loss
percentage while adjusting for baseline BMI and sex:
`baseline RAA ~ max_weight_loss_perc + baseline_BMI + Female`.

```{r ageaccel-model6-baseline-raa, warning=FALSE, message=FALSE}

ageaccel_model6_baseline_RAA <- map_dfr(
  names(ageaccel_titles),
  ~ compare_baseline_model6(df_age_accel, .x)
) %>%
  mutate(
    clock = unname(unlist(ageaccel_titles[outcome])),
    p_value = signif(p_value, 3),
    across(c(beta, SE, df, t_value), ~ round(.x, 3))
  ) %>%
  select(clock, outcome, response, model, term, beta, SE, df, t_value, p_value, n)

knitr::kable(
  ageaccel_model6_baseline_RAA,
  caption = "Model 6 baseline-only sensitivity: baseline RAA ~ weight loss + baseline BMI + gender",
  digits = 3
)

write_csv(
  ageaccel_model6_baseline_RAA,
  "figures/AgeAccel_weight_loss_model6_baseline_RAA.csv"
)
```

## Model 7: baseline RAA predicting weight loss

Model 7 is a baseline-only linear model restricted to D-1 samples with maximum
weight-loss percentage as the outcome. It tests whether baseline RAA predicts
subsequent maximum weight loss after adjustment for baseline BMI and sex:
`max_weight_loss_perc ~ baseline RAA + baseline_BMI + Female`.

```{r ageaccel-model7-weight-loss, warning=FALSE, message=FALSE}

ageaccel_model7_weight_loss <- map_dfr(
  names(ageaccel_titles),
  ~ compare_baseline_model7(df_age_accel, .x)
) %>%
  mutate(
    clock = unname(unlist(ageaccel_titles[outcome])),
    p_value = signif(p_value, 3),
    across(c(beta, SE, df, t_value), ~ round(.x, 3))
  ) %>%
  select(clock, outcome, response, model, term, beta, SE, df, t_value, p_value, n)

knitr::kable(
  ageaccel_model7_weight_loss,
  caption = "Model 7 baseline-only prediction: weight loss ~ baseline RAA + baseline BMI + gender",
  digits = 3
)

write_csv(
  ageaccel_model7_weight_loss,
  "figures/AgeAccel_weight_loss_model7_weight_loss_outcome.csv"
)
```

## Model 3 visualization: RAA versus maximum weight loss

These plots show the observed RAA values against maximum weight-loss percentage
and overlay Model 3 marginal predictions by timepoint.

```{r ageaccel-weights, warning=FALSE, message=FALSE}
# Build plots for each age acceleration outcome
ageaccel_plots <- pmap(
  list(
    outcome = names(ageaccel_titles),
    title = unname(ageaccel_titles),
    ylab = unname(ageaccel_titles)
  ),
  ~ fit_ageaccel(df_age_accel, ..1, ylab = ..3, title = ..2)
)
names(ageaccel_plots) <- names(ageaccel_titles)
ageaccel_plots <- compact(ageaccel_plots)  # drop NULLs

# Combine and save
if (length(ageaccel_plots) > 0) {
  final_ageaccel <- wrap_plots(ageaccel_plots, ncol = 4) +
    plot_layout(guides = "collect") +
    theme(legend.position = "right")
  
  ggsave("figures/AgeAccel.pdf", final_ageaccel, width = 13, height = 8)
  final_ageaccel
}
```

## Model 4 visualization: RAA versus maximum weight loss

These plots use the same visualization as Model 3, but the marginal prediction
lines and annotated coefficient come from Model 4, which additionally adjusts
for baseline BMI.

```{r ageaccel-weights-model4, warning=FALSE, message=FALSE}
# Build Model 4 plots for each age acceleration outcome
ageaccel_model4_plots <- pmap(
  list(
    outcome = names(ageaccel_titles),
    title = unname(ageaccel_titles),
    ylab = unname(ageaccel_titles)
  ),
  ~ fit_ageaccel(
    df_age_accel,
    ..1,
    ylab = ..3,
    title = ..2,
    model_version = "model4"
  )
)
names(ageaccel_model4_plots) <- names(ageaccel_titles)
ageaccel_model4_plots <- compact(ageaccel_model4_plots)  # drop NULLs

# Combine and save
if (length(ageaccel_model4_plots) > 0) {
  final_ageaccel_model4 <- wrap_plots(ageaccel_model4_plots, ncol = 4) +
    plot_layout(guides = "collect") +
    theme(legend.position = "right")

  ggsave(
    "figures/AgeAccel_model4.pdf",
    final_ageaccel_model4,
    width = 13,
    height = 8
  )
  final_ageaccel_model4
}
```

# EpiAgeAccel Subset

```{r ageaccel-subset, warning=FALSE, message=FALSE}
# Select subset of clocks to display
subset_names <- c(
  "PCHorvath1_RAA",
  "PCHorvath2_RAA",
  "PCPhenoAge_RAA",
  "DNAmGrimAge2_RAA",
  "epi_age_DunedinPACE_RAA",
  "CausAge_RAA"
)

# Filter plots by names
ageaccel_subset <- ageaccel_plots[subset_names]
ageaccel_subset <- compact(ageaccel_subset)  # drop any NULLs

# Combine and save subset
if (length(ageaccel_subset) > 0) {
  final_subset <- wrap_plots(ageaccel_subset, ncol = 3) +
    plot_layout(guides = "collect") +
    theme(legend.position = "right")
  
  ggsave("figures/AgeAccel_subset.pdf", final_subset, width = 11, height = 6)
  final_subset
}
```
