---
title: "Integrated visualization of epigenetic clocks"
subtitle: "SRAA trajectories, clock-response PCA, clinical associations, and Figure 4 outputs"
abstract: |
  This notebook performs the integrated analysis of merged epigenetic clock
  outputs from `data/AllClocks_Merged.csv`. It joins clock predictions with
  three-timepoint metadata, calculates residual age acceleration (RAA), scales
  RAA by baseline standard deviation to obtain standardized residual age
  acceleration (SRAA), and computes within-participant SRAA deltas from D-1 to
  D+10 and M+1. Mixed models estimate adjusted clock trajectories while
  controlling for sex and blood-cell composition. The notebook then performs PCA
  on delta-SRAA profiles to summarize coordinated clock responses, tests
  Spearman associations between PCA scores and clinical variables, and examines
  DamAge/AdaptAge delta-SRAA associations with glucose, insulin, and HOMA. It
  exports Figure 4, supplementary Figure 4 panels, PCA loading PDFs, plot/result
  RDS files, and association tables.
format:
  html:
    theme: sandstone
    embed-resources: true
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, 
                      class.output="bg-light")
```

# Load

```{r libraries, message=FALSE, warning=FALSE}
# Visualization
library(ggplot2)
library(cowplot)
library(ggpubr)
library(ggrepel)
library(plotly)

# Data manipulation
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(tibble)

# Statistical modeling
library(emmeans)
library(lme4)
library(lmerTest)

# Output
library(writexl)
```

```{r table-helpers}
render_table <- function(x, caption = NULL, digits = 3, font_size = "11px") {
  if (is.null(x) || nrow(x) == 0) {
    cat("_No rows to display._\n")
    return(invisible(NULL))
  }

  knitr::kable(x, digits = digits, caption = caption, format = "html") %>%
    as.character() %>%
    sprintf(
      "<div style='font-size: %s; overflow-x: auto;'>%s</div>",
      font_size,
      .
    ) %>%
    knitr::asis_output()
}

arrange_adjustment_pairs <- function(df, group_cols, adjustment_col = "model") {
  adjustment_levels <- c(
    "With blood-cell adjustment",
    "Without blood-cell adjustment"
  )

  df %>%
    mutate(.adjustment_order = match(.data[[adjustment_col]], adjustment_levels)) %>%
    arrange(!!!rlang::syms(group_cols), .adjustment_order, .data[[adjustment_col]]) %>%
    select(-.adjustment_order)
}
```

```{r load-data}
devtools::load_all("~/public/SiyuanLuo/projects/LongAge")

df_age <- read_csv("data/AllClocks_Merged.csv")
meta <- read_csv("data/meta_3timepoints.csv")

df_age <- df_age %>%
  left_join(
    meta %>% select(Sample, ID, Timepoint, NEUTROPHILE, LYMPHOZYTEN),
    by = "Sample"
  ) %>%
  mutate(Timepoint = factor(Timepoint, levels = c("D-1", "D+10", "M+1"))) %>%
  rename(DunedinPACE = epi_age_DunedinPACE)
```

```{r config}
# ============================================================================ #
# Configuration: Clocks and clinical variables
# ============================================================================ #

# Main epigenetic clocks
clock_cols <- c(
  "PCHorvath1", "PCHorvath2", "PCHannum", "PCPhenoAge",
  "PCGrimAge", "DNAmGrimAge2", "PCDNAmTL", "DunedinPACE",
  "SystemsAge", "AdaptAge", "DamAge", "CausAge"
)

# Clock metadata for PCA visualization
clock_meta <- tibble::tribble(
  ~clock,                                 ~generation,      ~family,
  "PCHorvath1_RAA_z_base_delta",          "Gen1",           "Chronological",
  "PCHorvath2_RAA_z_base_delta",          "Gen1",           "Chronological",
  "PCHannum_RAA_z_base_delta",            "Gen1",           "Chronological",
  "PCPhenoAge_RAA_z_base_delta",          "Gen2+",          "Risk/Health",
  "PCGrimAge_RAA_z_base_delta",           "Gen2+",          "Risk/Health",
  "DNAmGrimAge2_RAA_z_base_delta",        "Gen2+",          "Risk/Health",
  "DunedinPACE_RAA_z_base_delta", "Gen2+",          "Pace",
  "PCDNAmTL_RAA_z_base_delta",            "Telomere",       "Telomere",
  "SystemsAge_RAA_z_base_delta",          "Gen2+",          "Risk/Health",
  "AdaptAge_RAA_z_base_delta",            "Causal",         "Mechanistic/Other",
  "CausAge_RAA_z_base_delta",             "Causal",         "Mechanistic/Other",
  "DamAge_RAA_z_base_delta",              "Causal",         "Mechanistic/Other"
)
clock_meta$generation <- factor(clock_meta$generation, levels = c("Gen1", "Gen2+", "Telomere", "Causal"))

# Clinical variables for association testing
vars_metabolic <- c(
  "max_weight_loss_perc", "ketones", "Ketosemia", "glucose", "Insulin",
  "Homa", "triglyceride", "hdl", "ldl", "cholesterol"
)

vars_stress <- c("crp_hs", "ESR", "Cortisol", "pulse", "sbp")

vars_hematology <- c(
  "hemoglobin", "hematocrit", "leukocytes",
  "Retikulozyten.ProduktionsIndex", "Erythrozytenverteilungsbreite"
)

vars_subject <- c(
  "tiredness", "hunger", "sleep_disturbance", "anxiety",
  "energy_level_1", "MENTAL_WELL.BEING"
)
```

# Main Analysis Pipeline

```{r main-function}
main <- function(df_age, clock_cols, 
                 remove_prefix = FALSE, 
                 prefix_patterns = NULL,
                 covariates = c("Gender")) {
  clock_cols_RAA <- paste0(clock_cols, "_RAA")
  clock_cols_RAA_z_base <- paste0(clock_cols_RAA, "_z_base")
  clock_cols_RAA_z_base_delta <- paste0(clock_cols_RAA_z_base, "_delta")
  
  res <- NULL
  invisible(capture.output(
    res <- CalResidAgeAccel(df_age, clock_cols=clock_cols, age_col = "Age")
  ))
  data_wide <- res$data_wide
  data_wide <- scale_by_baseline_sd(data_wide, vars=clock_cols_RAA, time_col = "Timepoint", baseline_label = "D-1",suffix = "z_base")
    
  data_wide <- delta_from_subject_baseline(data_wide, vars=clock_cols_RAA_z_base,
                                          id_var = "ID",
                                          time_col = "Timepoint",
                                          baseline_label = "D-1",
                                          suffix = "delta")
  res <- delta_CI_mixed_model(data_wide, 
                              var_cols=clock_cols_RAA_z_base,
                              time_var        = "Timepoint",
                              baseline_label  = "D-1",
                              covariates      = covariates,
                              random_re       = "(1|ID)",
                              ci_level        = 0.95,
                              REML            = TRUE)
  
  tmp <- prep_traj_summary(res$effects,
      baseline_label = "D-1",
      var_label = "clock",
      keep_contrast = TRUE)
  
  df_long <- data_wide %>% pivot_longer(cols = clock_cols_RAA_z_base, names_to = "clock", values_to = "value") %>%
    dplyr::rename(Time = !!rlang::sym("Timepoint"))
  
  name_facets <- sub("_RAA_z_base", "", clock_cols_RAA_z_base)
  
  # Optionally remove prefixes from labels
  if (remove_prefix && !is.null(prefix_patterns)) {
    # prefix_patterns can be a single string or a vector of strings to remove
    for (pattern in prefix_patterns) {
      name_facets <- gsub(paste0("^", pattern), "", name_facets)
    }
  }
  
  names(name_facets) <- clock_cols_RAA_z_base
  
  p1 <- plotEffectTrajectories(
    traj_summary = tmp,
    df_long = df_long,
    baseline_label = "D-1",
    y = "SRAA",
    id_var = "ID",
    value_col = "value",
    x = "",
    facet_ncol = 6,
    facet_labels = name_facets,
    facet_order = names(name_facets),
    title = "Standardized residual age acceleration (baseline re-centered)",
    subtitle = NULL)
    # subtitle = "Black = marginal mean (adjusted for covariates) ± 95% CI; grey = individual trajectories; y-axis on baseline SD units.")
  
  delta_long <- data_wide %>% 
    pivot_longer(cols = clock_cols_RAA_z_base_delta, names_to = "clock", values_to = "delta") 
  
  p2 <- plot_raincloud_delta(delta_long,
                       res$effects,
                       var_label = "clock",
                       time_col = "Timepoint",
                       value_col = "delta",
                       baseline_label = "D-1",
                       y_margin = 0.15,
                       hide_ns = TRUE,
                       x_labels = name_facets,
                       title="Within-subject effect sizes (baseline SD units)",
                       subtitle = NULL)
                       # subtitle="Standardized residual age acceleration (SRAA) change from baseline. \nStars from mixed-model contrasts vs baseline")
  return(list(p1=p1, p2=p2, data_wide=data_wide, effects=res$effects, covariates=covariates))
}
```

```{r effect-size-helpers}
format_time_effects <- function(effects, model_label) {
  effects %>%
    mutate(
      clock = str_remove(variable, "_RAA_z_base$"),
      timepoint = str_match(contrast, "\\(([^)]+)\\)")[, 2],
      model = model_label
    ) %>%
    transmute(
      model,
      clock,
      timepoint,
      effect_size = estimate,
      lower_95 = lower,
      upper_95 = upper,
      SE,
      df,
      p_value = p
    )
}

fit_time_by_sex_models <- function(data_wide,
                                   clock_cols,
                                   sex_var = "Gender",
                                   cell_covariate = "LYMPHOZYTEN",
                                   time_var = "Timepoint",
                                   id_var = "ID",
                                   baseline_label = "D-1",
                                   ci_level = 0.95,
                                   REML = TRUE,
                                   adjust = "none") {
  clock_cols_RAA_z_base <- paste0(clock_cols, "_RAA_z_base")

  fits <- map(clock_cols_RAA_z_base, function(var_name) {
    need_vars <- c(var_name, time_var, sex_var, cell_covariate, id_var)
    dat <- data_wide %>%
      select(all_of(need_vars)) %>%
      drop_na()

    if (nrow(dat) < 5) {
      warning(sprintf("Variable '%s': not enough complete rows. Skipping.", var_name))
      return(NULL)
    }

    dat[[time_var]] <- relevel(factor(dat[[time_var]]), ref = baseline_label)
    dat[[sex_var]] <- factor(dat[[sex_var]])

    form_txt <- sprintf(
      "%s ~ %s * %s + %s + (1 | %s)",
      var_name,
      time_var,
      sex_var,
      cell_covariate,
      id_var
    )
    fit <- lmer(as.formula(form_txt), data = dat, REML = REML)

    emm <- emmeans(fit, as.formula(paste("~", time_var, "|", sex_var)))
    time_contrasts <- contrast(
      emm,
      method = "trt.vs.ctrl",
      ref = baseline_label,
      adjust = adjust
    )

    sex_summary <- time_contrasts %>%
      summary(infer = TRUE, level = ci_level) %>%
      as_tibble()

    if (!"df" %in% colnames(sex_summary)) {
      sex_summary$df <- NA_real_
    }

    sex_contrasts <- sex_summary %>%
      transmute(
        variable = var_name,
        sex = .data[[sex_var]],
        contrast,
        estimate = estimate,
        lower = lower.CL,
        upper = upper.CL,
        SE = SE,
        df = df,
        t = t.ratio,
        p = p.value
      )

    interaction_summary <- contrast(
      time_contrasts,
      method = "pairwise",
      by = "contrast",
      adjust = adjust
    ) %>%
      summary(infer = TRUE, level = ci_level) %>%
      as_tibble()

    if (!"df" %in% colnames(interaction_summary)) {
      interaction_summary$df <- NA_real_
    }

    interaction_terms <- interaction_summary %>%
      transmute(
        variable = var_name,
        time_contrast = contrast,
        sex_difference = contrast1,
        estimate = estimate,
        lower = lower.CL,
        upper = upper.CL,
        SE = SE,
        df = df,
        t = t.ratio,
        p = p.value
      )

    list(sex_contrasts = sex_contrasts, interaction_terms = interaction_terms)
  }) %>%
    compact()

  list(
    sex_contrasts = bind_rows(map(fits, "sex_contrasts")),
    interaction_terms = bind_rows(map(fits, "interaction_terms"))
  )
}

format_time_by_sex_effects <- function(effects, model_label) {
  effects %>%
    mutate(
      clock = str_remove(variable, "_RAA_z_base$"),
      timepoint = str_match(contrast, "\\(([^)]+)\\)")[, 2],
      model = model_label
    ) %>%
    transmute(
      model,
      clock,
      sex,
      timepoint,
      effect_size = estimate,
      lower_95 = lower,
      upper_95 = upper,
      SE,
      df,
      p_value = p
    )
}

extract_blood_cell_effects <- function(data_wide,
                                       clock_cols,
                                       cell_covariates = c("NEUTROPHILE", "LYMPHOZYTEN"),
                                       adjustment_covariates = "Gender",
                                       time_var = "Timepoint",
                                       baseline_label = "D-1",
                                       random_re = "(1|ID)",
                                       ci_level = 0.95,
                                       REML = TRUE) {
  clock_cols_RAA_z_base <- paste0(clock_cols, "_RAA_z_base")
  cell_covariates_z <- paste0(cell_covariates, "_z")

  df_model <- data_wide %>%
    mutate(
      across(
        all_of(cell_covariates),
        ~ as.numeric(scale(.x)),
        .names = "{.col}_z"
      )
    )

  bind_rows(lapply(clock_cols_RAA_z_base, function(var_name) {
    rhs_terms <- c(time_var, adjustment_covariates, cell_covariates_z)
    form_txt <- sprintf("%s ~ %s + %s", var_name, paste(rhs_terms, collapse = " + "), random_re)
    need_vars <- unique(c(var_name, time_var, "ID", adjustment_covariates, cell_covariates_z))

    dat <- df_model %>%
      select(all_of(need_vars)) %>%
      drop_na()

    if (nrow(dat) < 5) return(NULL)

    dat[[time_var]] <- relevel(factor(dat[[time_var]]), ref = baseline_label)
    fit <- lmer(as.formula(form_txt), data = dat, REML = REML)

    coef_df <- as.data.frame(coef(summary(fit))) %>%
      rownames_to_column("term") %>%
      filter(term %in% cell_covariates_z) %>%
      as_tibble()

    names(coef_df) <- make.names(names(coef_df))
    ci_df <- as.data.frame(confint(fit, parm = cell_covariates_z, method = "Wald", level = ci_level)) %>%
      rownames_to_column("term") %>%
      as_tibble()
    names(ci_df) <- c("term", "lower_95", "upper_95")

    coef_df %>%
      left_join(ci_df, by = "term") %>%
      transmute(
        clock = str_remove(var_name, "_RAA_z_base$"),
        term = str_remove(term, "_z$"),
        effect_size = Estimate,
        lower_95,
        upper_95,
        SE = Std..Error,
        df = NA_real_,
        p_value = 2 * pnorm(abs(t.value), lower.tail = FALSE)
      )
  }))
}

format_component_time_effects <- function(adjusted_res,
                                          no_cell_res,
                                          component_set,
                                          prefix_patterns = NULL) {
  bind_rows(
    format_time_effects(adjusted_res$effects, "With blood-cell adjustment"),
    format_time_effects(no_cell_res$effects, "Without blood-cell adjustment")
  ) %>%
    mutate(
      component_set = component_set,
      component = clock
    ) %>%
    {
      out <- .
      if (!is.null(prefix_patterns)) {
        for (pattern in prefix_patterns) {
          out$component <- str_remove(out$component, paste0("^", pattern))
        }
      }
      out
    } %>%
    transmute(
      component_set,
      model,
      component,
      timepoint,
      effect_size,
      lower_95,
      upper_95,
      SE,
      df,
      p_value
    )
}

format_component_blood_cell_effects <- function(adjusted_res,
                                                clock_cols,
                                                component_set,
                                                prefix_patterns = NULL) {
  extract_blood_cell_effects(adjusted_res$data_wide, clock_cols) %>%
    mutate(
      component_set = component_set,
      component = clock
    ) %>%
    {
      out <- .
      if (!is.null(prefix_patterns)) {
        for (pattern in prefix_patterns) {
          out$component <- str_remove(out$component, paste0("^", pattern))
        }
      }
      out
    } %>%
    transmute(
      component_set,
      component,
      term,
      effect_size,
      lower_95,
      upper_95,
      SE,
      df,
      p_value
    )
}
```

## Check collinearity of blood cell covariates

```{r}
# check collinearity for blood cell covariates: lymphocytes and neutrophils
cor(df_age$LYMPHOZYTEN, df_age$NEUTROPHILE)
lm_model <- lm(epi_age_Horvath ~LYMPHOZYTEN + NEUTROPHILE, data = df_age)
car::vif(lm_model)
```

Correlation is -0.93, which is very high. VIFs are around 7.5, which is high. This suggests that there is strong collinearity, and one should include only one of the two covariates in the model.

```{r run-main-analysis, fig.width=10, fig.height=6}
# Run main analysis pipeline
main_res <- main(df_age, clock_cols, covariates = c("Gender", "LYMPHOZYTEN"))
main_res_no_cell <- main(df_age, clock_cols, covariates = "Gender")
data_wide <- main_res$data_wide

# Display plots
main_res$p1 + labs(title = "Standardized residual age acceleration (baseline re-centered, composition adjusted)")
main_res$p2 + labs(title = "Within-subject effect sizes (baseline SD units, composition adjusted)")
main_res_no_cell$p1
main_res_no_cell$p2
```

## Timepoint and Blood-Cell Effect Sizes

Effect sizes are in baseline-SD units of residual age acceleration. Timepoint
effects are mixed-model contrasts versus D-1. Blood-cell effects are fixed-effect
coefficients from the blood-cell-adjusted model after standardizing each blood
cell proportion, so each estimate is the expected SRAA difference per 1 SD
higher neutrophil or lymphocyte proportion.

```{r timepoint-blood-cell-effect-sizes, results='asis'}
timepoint_effect_sizes <- bind_rows(
  format_time_effects(main_res$effects, "With blood-cell adjustment"),
  format_time_effects(main_res_no_cell$effects, "Without blood-cell adjustment")
) %>%
  arrange_adjustment_pairs(group_cols = c("clock", "timepoint")) %>%
  mutate(
    p_value = signif(p_value, 3),
    across(c(effect_size, lower_95, upper_95, SE, df), ~ round(.x, 3))
  )

blood_cell_effect_sizes <- extract_blood_cell_effects(data_wide, clock_cols) %>%
  arrange(clock, term) %>%
  mutate(
    p_value = signif(p_value, 3),
    across(c(effect_size, lower_95, upper_95, SE, df), ~ round(.x, 3))
  )

render_table(
  timepoint_effect_sizes,
  caption = "Timepoint effect sizes with and without blood-cell adjustment"
)

render_table(
  blood_cell_effect_sizes,
  caption = "Standardized blood-cell proportion effect sizes in the adjusted models"
)
```

## Timepoint-by-Sex Interaction Model

This sensitivity model keeps the same SRAA outcome scale, lymphocyte adjustment,
and participant random intercept as the main blood-cell-adjusted model, but
allows the timepoint effect to differ by sex:
`clock_RAA_z_base ~ Timepoint * Gender + LYMPHOZYTEN + (1 | ID)`.

The first table reports sex-specific mixed-model contrasts for D+10 and M+1
versus D-1. The second table reports contrast-of-contrasts tests, which test
whether each timepoint contrast differs between the sex groups.

```{r timepoint-sex-interaction-model, results='asis'}
time_by_sex_res <- fit_time_by_sex_models(
  data_wide,
  clock_cols,
  sex_var = "Gender",
  cell_covariate = "LYMPHOZYTEN",
  time_var = "Timepoint",
  id_var = "ID",
  baseline_label = "D-1"
)

time_by_sex_effect_sizes <- format_time_by_sex_effects(
  time_by_sex_res$sex_contrasts,
  "Timepoint x sex + lymphocyte adjustment"
) %>%
  arrange(clock, sex, timepoint) %>%
  mutate(
    p_value = signif(p_value, 3),
    across(c(effect_size, lower_95, upper_95, SE, df), ~ round(.x, 3))
  )

time_by_sex_interaction_terms <- time_by_sex_res$interaction_terms %>%
  mutate(
    clock = str_remove(variable, "_RAA_z_base$"),
    p_value = signif(p, 3),
    across(c(estimate, lower, upper, SE, df, t), ~ round(.x, 3))
  ) %>%
  select(
    clock,
    time_contrast,
    sex_difference,
    estimate,
    lower_95 = lower,
    upper_95 = upper,
    SE,
    df,
    t,
    p_value
  )

render_table(
  time_by_sex_effect_sizes,
  caption = "Sex-specific timepoint effect sizes from Timepoint x Gender models"
)

render_table(
  time_by_sex_interaction_terms,
  caption = "Timepoint-by-sex contrast-of-contrasts tests"
)

write_csv(
  time_by_sex_effect_sizes,
  "figures/timepoint_by_sex_effect_sizes.csv"
)

write_csv(
  time_by_sex_interaction_terms,
  "figures/timepoint_by_sex_interaction_terms.csv"
)
```

```{r save-main-plots}
# Save individual plots
pdf("figures/spaghetti_with_blood_cell_adjustment.pdf", width = 8, height = 5)
plot(main_res$p1+ labs(title = "Standardized residual age acceleration (baseline re-centered, composition adjusted)"))
dev.off()

pdf("figures/raincloud_with_blood_cell_adjustment.pdf", width = 8, height = 5)
plot(main_res$p2 + labs(title = "Within-subject effect sizes (baseline SD units, composition adjusted)"))
dev.off()

pdf("figures/spaghetti_no_blood_cell_adjustment.pdf", width = 8, height = 5)
plot(main_res_no_cell$p1)
dev.off()

pdf("figures/raincloud_no_blood_cell_adjustment.pdf", width = 8, height = 5)
plot(main_res_no_cell$p2)
dev.off()
```

# Decomposition Analysis

This decomposition estimates the part of each clock signal explained by blood
cell composition and then tests whether the composition-derived and residual
components change over time. Because lymphocytes and neutrophils are highly
collinear in this dataset, the default model uses lymphocytes only.

Step A - predict the composition component:

```r
clock ~ lymphocytes + (1|subject)
```

In the code below, this is fit as:

```r
clock_RAA_z_base ~ LYMPHOZYTEN_z + (1|ID)
```

Step B - compute the two decomposed clock components:

* `composition_component`: predicted value from fixed effects of lymphocytes
* `residual_component`: observed clock value minus the composition component

```r
composition_component = predict(model, re.form = NA)
residual_component = observed_clock - composition_component
```

Step C - test whether each component changes over time, adjusted for gender:

```r
composition_component ~ Timepoint + Gender + (1|ID)
residual_component ~ Timepoint + Gender + (1|ID)
```

Step D - visualize the decomposition:

* Option A: composition component and residual component trajectories over time
* Option B: unadjusted versus lymphocyte-adjusted time effects per clock

Optional interpretation:

* `composition-driven`: the composition component changes over time and explains the clock response
* `intrinsic`: the residual component changes over time after removing composition signal
* `masked`: the adjusted and unadjusted effects differ substantially, suggesting composition masks or inflates the apparent clock response

## Lymphocyte Change From Baseline

This section visualizes the within-subject change in lymphocytes from D-1 and summarizes the mean, standard deviation, and coefficient of variation at each follow-up timepoint.

```{r lymphocyte-change-from-baseline, fig.width=6, fig.height=4, results='asis'}
lymphocyte_delta_data <- data_wide %>%
  select(ID, Timepoint, LYMPHOZYTEN) %>%
  group_by(ID) %>%
  mutate(
    lymphocyte_baseline = LYMPHOZYTEN[Timepoint == "D-1"][1],
    lymphocyte_delta = LYMPHOZYTEN - lymphocyte_baseline
  ) %>%
  ungroup() %>%
  filter(Timepoint != "D-1") %>%
  mutate(Timepoint = factor(Timepoint, levels = c("D+10", "M+1")))

lymphocyte_change_summary <- lymphocyte_delta_data %>%
  group_by(Timepoint) %>%
  summarise(
    n = sum(!is.na(lymphocyte_delta)),
    mean_delta = mean(lymphocyte_delta, na.rm = TRUE),
    sd_delta = sd(lymphocyte_delta, na.rm = TRUE),
    cv_delta = sd_delta / abs(mean_delta),
    cv_delta_percent = 100 * cv_delta,
    .groups = "drop"
  ) %>%
  mutate(
    mean_sd = sprintf("%.3f +/- %.3f", mean_delta, sd_delta),
    across(c(mean_delta, sd_delta, cv_delta, cv_delta_percent), ~ round(.x, 3))
  )

lymphocyte_plot_labels <- lymphocyte_change_summary %>%
  mutate(
    label = sprintf(
      "mean = %.3f\nSD = %.3f\nCV = %.1f%%",
      mean_delta,
      sd_delta,
      cv_delta_percent
    )
  ) %>%
  left_join(
    lymphocyte_delta_data %>%
      group_by(Timepoint) %>%
      summarise(
        y_top = max(lymphocyte_delta, na.rm = TRUE),
        y_bottom = min(lymphocyte_delta, na.rm = TRUE),
        .groups = "drop"
      ),
    by = "Timepoint"
  ) %>%
  mutate(
    y_span = y_top - y_bottom,
    y = y_top - 0.05 * y_span
  )

lymphocyte_y_limits <- range(lymphocyte_delta_data$lymphocyte_delta, na.rm = TRUE)
lymphocyte_y_span <- diff(lymphocyte_y_limits)
if (lymphocyte_y_span == 0) lymphocyte_y_span <- 1

render_table(
  lymphocyte_change_summary,
  caption = "Mean, SD, and coefficient of variation of within-subject lymphocyte changes from D-1"
)

p_lymphocyte_delta <- ggplot(
  lymphocyte_delta_data,
  aes(x = Timepoint, y = lymphocyte_delta)
) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.3) +
  geom_jitter(width = 0.08, height = 0, alpha = 0.6, size = 1.4) +
  stat_summary(
    fun = mean,
    geom = "point",
    size = 2.6,
    color = "red"
  ) +
  stat_summary(
    fun.data = function(x) {
      data.frame(
        y = mean(x, na.rm = TRUE),
        ymin = mean(x, na.rm = TRUE) - sd(x, na.rm = TRUE),
        ymax = mean(x, na.rm = TRUE) + sd(x, na.rm = TRUE)
      )
    },
    geom = "errorbar",
    width = 0.12,
    color = "red"
  ) +
  geom_text(
    data = lymphocyte_plot_labels,
    aes(x = Timepoint, y = y, label = label),
    inherit.aes = FALSE,
    size = 3.2,
    vjust = 1,
    lineheight = 0.9
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0.08, 0.12))
  ) +
  theme_bw(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    plot.margin = margin(t = 8, r = 8, b = 8, l = 8)
  ) +
  labs(
    title = "Within-subject lymphocyte change from baseline",
    x = NULL,
    y = expression(Delta * "lymphocytes from D-1")
  )

p_lymphocyte_delta
```

```{r decomposition-functions}
extract_time_contrasts <- function(fit,
                                   time_var = "Timepoint",
                                   baseline_label = "D-1",
                                   ci_level = 0.95) {
  emm <- emmeans(fit, specs = time_var)
  ref_idx <- which(emm@levels[[time_var]] == baseline_label)

  if (length(ref_idx) != 1) {
    stop("baseline_label must be one of the fitted timepoint levels.")
  }

  as.data.frame(
    summary(
      contrast(emm, method = "trt.vs.ctrl", ref = ref_idx),
      infer = TRUE,
      level = ci_level
    )
  ) %>%
    as_tibble() %>%
    transmute(
      contrast,
      timepoint = str_match(contrast, "^\\(?([^()]+)\\)? - ")[, 2],
      effect_size = estimate,
      lower_95 = lower.CL,
      upper_95 = upper.CL,
      SE,
      df,
      p_value = p.value
    )
}

decompose_clock_composition <- function(data_wide,
                                        clock_cols,
                                        cell_covariates = "LYMPHOZYTEN",
                                        value_suffix = "_RAA_z_base",
                                        adjustment_covariates = "Gender",
                                        id_col = "ID",
                                        time_col = "Timepoint",
                                        baseline_label = "D-1",
                                        ci_level = 0.95,
                                        REML = TRUE) {
  value_cols <- paste0(clock_cols, value_suffix)
  cell_covariates_z <- paste0(cell_covariates, "_z")

  df_model <- data_wide %>%
    mutate(
      across(
        all_of(cell_covariates),
        ~ as.numeric(scale(.x)),
        .names = "{.col}_z"
      )
    )

  decomposed_data <- bind_rows(lapply(seq_along(clock_cols), function(i) {
    clock <- clock_cols[[i]]
    value_col <- value_cols[[i]]
    need_vars <- unique(c(
      id_col,
      time_col,
      value_col,
      cell_covariates_z,
      adjustment_covariates
    ))

    dat <- df_model %>%
      select(all_of(need_vars)) %>%
      drop_na()

    if (nrow(dat) < 5) return(NULL)

    form_txt <- sprintf(
      "%s ~ %s + (1|%s)",
      value_col,
      paste(cell_covariates_z, collapse = " + "),
      id_col
    )
    fit <- lmer(as.formula(form_txt), data = dat, REML = REML)

    dat %>%
      mutate(
        clock = clock,
        observed_component = .data[[value_col]],
        composition_component = as.numeric(predict(fit, re.form = NA)),
        residual_component = observed_component - composition_component
      ) %>%
      select(
        all_of(c(id_col, time_col)),
        all_of(adjustment_covariates),
        clock,
        observed_component,
        composition_component,
        residual_component
      )
  }))

  time_effects <- decomposed_data %>%
    pivot_longer(
      cols = c(composition_component, residual_component),
      names_to = "component",
      values_to = "value"
    ) %>%
    group_by(clock, component) %>%
    group_modify(~ {
      dat <- .x %>%
        mutate("{time_col}" := relevel(factor(.data[[time_col]]), ref = baseline_label))
      rhs_terms <- c(time_col, adjustment_covariates)

      fit <- lmer(
        as.formula(sprintf(
          "value ~ %s + (1|%s)",
          paste(rhs_terms, collapse = " + "),
          id_col
        )),
        data = dat,
        REML = REML
      )

      extract_time_contrasts(
        fit,
        time_var = time_col,
        baseline_label = baseline_label,
        ci_level = ci_level
      )
    }) %>%
    ungroup()

  list(data = decomposed_data, effects = time_effects)
}

decomposition_p_to_stars <- function(p) {
  case_when(
    is.na(p) ~ "",
    p < 0.001 ~ "***",
    p < 0.01 ~ "**",
    p < 0.05 ~ "*",
    TRUE ~ "ns"
  )
}

clean_decomposition_labels <- function(x, prefix_patterns = NULL) {
  out <- x
  if (!is.null(prefix_patterns)) {
    for (pattern in prefix_patterns) {
      out <- str_remove(out, paste0("^", pattern))
    }
  }
  out
}

prepare_decomposition_outputs <- function(decomp_res,
                                          clock_cols,
                                          prefix_patterns = NULL,
                                          title_prefix = "Clock",
                                          y_label = "SRAA component") {
  clock_labels <- setNames(
    clean_decomposition_labels(clock_cols, prefix_patterns),
    clock_cols
  )

  effect_sizes <- decomp_res$effects %>%
    mutate(
      clock = factor(clock, levels = clock_cols),
      clock_label = clock_labels[as.character(clock)],
      component = recode(
        component,
        "composition_component" = "Composition component",
        "residual_component" = "Residual component"
      )
    ) %>%
    arrange(clock, component, timepoint)

  plot_data <- decomp_res$data %>%
    pivot_longer(
      cols = c(composition_component, residual_component),
      names_to = "component",
      values_to = "value"
    ) %>%
    mutate(
      component = recode(
        component,
        "composition_component" = "Composition component",
        "residual_component" = "Residual component"
      ),
      clock = factor(clock, levels = clock_cols),
      clock_label = factor(clock_labels[as.character(clock)], levels = unname(clock_labels)),
      Timepoint = factor(Timepoint, levels = c("D-1", "D+10", "M+1"))
    )

  stars <- effect_sizes %>%
    mutate(stars = decomposition_p_to_stars(p_value)) %>%
    filter(stars != "ns", !is.na(timepoint)) %>%
    left_join(
      plot_data %>%
        group_by(clock, component, Timepoint) %>%
        summarise(y_top = max(value, na.rm = TRUE), .groups = "drop"),
      by = c("clock", "component", "timepoint" = "Timepoint")
    ) %>%
    mutate(
      y = y_top + 0.15,
      Timepoint = factor(timepoint, levels = c("D-1", "D+10", "M+1")),
      clock_label = factor(clock_labels[as.character(clock)], levels = unname(clock_labels))
    )

  p_spaghetti <- ggplot(plot_data, aes(x = Timepoint, y = value, group = ID)) +
    geom_line(color = "grey75", linewidth = 0.3, alpha = 0.6) +
    geom_point(color = "grey45", size = 0.8, alpha = 0.6) +
    stat_summary(
      aes(group = 1),
      fun = mean,
      geom = "line",
      color = "black",
      linewidth = 0.7
    ) +
    stat_summary(
      aes(group = 1),
      fun.data = mean_cl_boot,
      geom = "errorbar",
      width = 0.12,
      color = "black"
    ) +
    geom_text(
      data = stars,
      aes(x = Timepoint, y = y, label = stars),
      inherit.aes = FALSE,
      vjust = 0,
      size = 4,
      color = "blue"
    ) +
    facet_grid(component ~ clock_label, scales = "free_y") +
    theme_minimal(base_size = 11) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(
      title = paste0("Spaghetti plot: ", title_prefix, " lymphocyte-associated and residual components"),
      x = NULL,
      y = y_label
    )

  delta_data <- plot_data %>%
    group_by(ID, clock, clock_label, component) %>%
    mutate(
      baseline_value = value[Timepoint == "D-1"][1],
      delta = value - baseline_value
    ) %>%
    ungroup() %>%
    filter(Timepoint != "D-1")

  delta_stars <- effect_sizes %>%
    mutate(
      stars = decomposition_p_to_stars(p_value),
      Timepoint = factor(timepoint, levels = c("D-1", "D+10", "M+1")),
      clock_label = factor(clock_labels[as.character(clock)], levels = unname(clock_labels))
    ) %>%
    filter(stars != "ns", !is.na(timepoint)) %>%
    left_join(
      delta_data %>%
        group_by(clock, component, Timepoint) %>%
        summarise(y_top = max(delta, na.rm = TRUE), .groups = "drop"),
      by = c("clock", "component", "Timepoint")
    ) %>%
    mutate(y = y_top + 0.15)

  p_raincloud <- ggplot(delta_data, aes(x = clock_label, y = delta)) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_violin(
      width = 0.8,
      fill = "grey88",
      color = "grey55",
      linewidth = 0.3,
      trim = FALSE
    ) +
    geom_jitter(width = 0.12, height = 0, alpha = 0.55, size = 1) +
    stat_summary(
      fun = mean,
      geom = "crossbar",
      width = 0.35,
      fatten = 2,
      color = "red"
    ) +
    geom_text(
      data = delta_stars,
      aes(x = clock_label, y = y, label = stars),
      inherit.aes = FALSE,
      vjust = 0,
      size = 4,
      color = "blue"
    ) +
    facet_grid(component ~ Timepoint, scales = "free_y") +
    theme_bw(base_size = 11) +
    theme(
      panel.grid.minor = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1)
    ) +
    labs(
      title = paste0("Raincloud plot: ", title_prefix, " component changes from baseline"),
      x = NULL,
      y = "Delta SRAA component from D-1"
    )

  list(
    effects = effect_sizes,
    plot_data = plot_data,
    delta_data = delta_data,
    spaghetti = p_spaghetti,
    raincloud = p_raincloud
  )
}
```

```{r decomposition-analysis, fig.width=10, fig.height=6, results='asis'}
decomp_res <- decompose_clock_composition(
  data_wide,
  clock_cols,
  cell_covariates = "LYMPHOZYTEN",
  adjustment_covariates = "Gender"
)

decomposition_effect_sizes <- decomp_res$effects %>%
  mutate(
    component = recode(
      component,
      "composition_component" = "Composition component",
      "residual_component" = "Residual component"
    ),
    p_value = signif(p_value, 3),
    across(c(effect_size, lower_95, upper_95, SE, df), ~ round(.x, 3))
  ) %>%
  arrange(clock, component, timepoint)

render_table(
  decomposition_effect_sizes,
  caption = "Timepoint effects for composition-derived and residual clock components"
)

decomposition_plot_data <- decomp_res$data %>%
  pivot_longer(
    cols = c(composition_component, residual_component),
    names_to = "component",
    values_to = "value"
  ) %>%
  mutate(
    component = recode(
      component,
      "composition_component" = "Composition component",
      "residual_component" = "Residual component"
    )
  )
decomposition_time_levels <- c("D-1", "D+10", "M+1")
decomposition_plot_data <- decomposition_plot_data %>%
  mutate(
    Timepoint = factor(Timepoint, levels = decomposition_time_levels),
    clock = factor(clock, levels = clock_cols)
  )

decomposition_stars <- decomp_res$effects %>%
  mutate(
    component = recode(
      component,
      "composition_component" = "Composition component",
      "residual_component" = "Residual component"
    ),
    stars = decomposition_p_to_stars(p_value)
  ) %>%
  filter(stars != "ns", !is.na(timepoint)) %>%
  left_join(
    decomposition_plot_data %>%
      group_by(clock, component, Timepoint) %>%
      summarise(y_top = max(value, na.rm = TRUE), .groups = "drop"),
    by = c("clock", "component", "timepoint" = "Timepoint")
  ) %>%
  mutate(
    y = y_top + 0.15,
    Timepoint = factor(timepoint, levels = decomposition_time_levels),
    clock = factor(clock, levels = clock_cols)
  )

p_decomposition <- ggplot(
  decomposition_plot_data,
  aes(x = Timepoint, y = value, group = ID)
) +
  geom_line(color = "grey75", linewidth = 0.3, alpha = 0.6) +
  geom_point(color = "grey45", size = 0.8, alpha = 0.6) +
  stat_summary(
    aes(group = 1),
    fun = mean,
    geom = "line",
    color = "black",
    linewidth = 0.7
  ) +
  stat_summary(
    aes(group = 1),
    fun.data = mean_cl_boot,
    geom = "errorbar",
    width = 0.12,
    color = "black"
  ) +
  geom_text(
    data = decomposition_stars,
    aes(x = Timepoint, y = y, label = stars),
    inherit.aes = FALSE,
    vjust = 0,
    size = 4,
    color = "blue"
  ) +
  facet_grid(component ~ clock, scales = "free_y") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "Spaghetti plot: lymphocyte-associated and residual clock components",
    x = NULL,
    y = "SRAA component"
  )

p_decomposition
```

```{r decomposition-raincloud, fig.width=10, fig.height=6}
decomposition_delta <- decomposition_plot_data %>%
  group_by(ID, clock, component) %>%
  mutate(
    baseline_value = value[Timepoint == "D-1"][1],
    delta = value - baseline_value
  ) %>%
  ungroup() %>%
  mutate(clock = factor(clock, levels = clock_cols)) %>%
  filter(Timepoint != "D-1")

decomposition_delta_stars <- decomp_res$effects %>%
  mutate(
    component = recode(
      component,
      "composition_component" = "Composition component",
      "residual_component" = "Residual component"
    ),
    stars = decomposition_p_to_stars(p_value),
    Timepoint = factor(timepoint, levels = decomposition_time_levels)
  ) %>%
  filter(stars != "ns", !is.na(timepoint)) %>%
  left_join(
    decomposition_delta %>%
      group_by(clock, component, Timepoint) %>%
      summarise(y_top = max(delta, na.rm = TRUE), .groups = "drop"),
    by = c("clock", "component", "Timepoint")
  ) %>%
  mutate(
    y = y_top + 0.15,
    clock = factor(clock, levels = clock_cols)
  )

p_decomposition_raincloud <- ggplot(
  decomposition_delta,
  aes(x = clock, y = delta)
) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_violin(
    width = 0.8,
    fill = "grey88",
    color = "grey55",
    linewidth = 0.3,
    trim = FALSE
  ) +
  geom_jitter(width = 0.12, height = 0, alpha = 0.55, size = 1) +
  stat_summary(
    fun = mean,
    geom = "crossbar",
    width = 0.35,
    fatten = 2,
    color = "red"
  ) +
  geom_text(
    data = decomposition_delta_stars,
    aes(x = clock, y = y, label = stars),
    inherit.aes = FALSE,
    vjust = 0,
    size = 4,
    color = "blue"
  ) +
  facet_grid(component ~ Timepoint, scales = "free_y") +
  theme_bw(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  labs(
    title = "Raincloud plot: within-subject component changes from baseline",
    x = NULL,
    y = "Delta SRAA component from D-1"
  )

p_decomposition_raincloud
```

```{r adjusted-vs-unadjusted-helpers}
clean_adjusted_vs_unadjusted_label <- function(x) {
  recode(
    x,
    PCHorvath1 = "Horvath1",
    PCHorvath2 = "Horvath2",
    PCHannum = "Hannum",
    PCPhenoAge = "PhenoAge",
    PCGrimAge = "GrimAge",
    DNAmGrimAge2 = "GrimAge2",
    PCDNAmTL = "DNAmTL",
    .default = x
  )
}

prepare_adjusted_vs_unadjusted <- function(effect_sizes,
                                           label_col = "clock") {
  effect_sizes %>%
    select(
      any_of("component_set"),
      label = all_of(label_col),
      timepoint,
      model,
      effect_size,
      p_value
    ) %>%
    pivot_wider(
      names_from = model,
      values_from = c(effect_size, p_value)
    ) %>%
    mutate(
      label = clean_adjusted_vs_unadjusted_label(label),
      delta_diff =
        `effect_size_With blood-cell adjustment` -
        `effect_size_Without blood-cell adjustment`,
      sig_category = case_when(
        `p_value_Without blood-cell adjustment` < 0.05 &
          `p_value_With blood-cell adjustment` < 0.05 ~ "both",
        `p_value_Without blood-cell adjustment` < 0.05 &
          `p_value_With blood-cell adjustment` >= 0.05 ~ "unadjusted_only",
        `p_value_Without blood-cell adjustment` >= 0.05 &
          `p_value_With blood-cell adjustment` < 0.05 ~ "adjusted_only",
        TRUE ~ "none"
      ),
      sig_category = factor(
        sig_category,
        levels = c("both", "unadjusted_only", "adjusted_only", "none")
      )
    )
}

plot_adjusted_vs_unadjusted <- function(adjusted_vs_unadjusted,
                                        title,
                                        facet_by_component_set = FALSE,
                                        facet_ncol = 2,
                                        label_size = 2.4,
                                        point_size = 2.1) {
  axis_lim <- range(
    c(
      adjusted_vs_unadjusted$`effect_size_Without blood-cell adjustment`,
      adjusted_vs_unadjusted$`effect_size_With blood-cell adjustment`
    ),
    na.rm = TRUE
  )
  axis_pad <- diff(axis_lim) * 0.08
  if (!is.finite(axis_pad) || axis_pad == 0) axis_pad <- 0.1
  axis_lim <- axis_lim + c(-axis_pad, axis_pad)

  p <- ggplot(
    adjusted_vs_unadjusted,
    aes(
      x = `effect_size_Without blood-cell adjustment`,
      y = `effect_size_With blood-cell adjustment`,
      label = label,
      color = delta_diff,
      shape = sig_category
    )
  ) +
    geom_hline(yintercept = 0, linewidth = 0.3, alpha = 0.5) +
    geom_vline(xintercept = 0, linewidth = 0.3, alpha = 0.5) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", alpha = 0.6) +
    geom_point(size = point_size) +
    geom_text_repel(
      size = label_size,
      max.overlaps = Inf,
      box.padding = 0.18,
      point.padding = 0.08,
      min.segment.length = 0,
      segment.alpha = 0.35,
      force = 1.8,
      force_pull = 0.4,
      max.time = 2
    ) +
    scale_color_gradient2(
      low = "blue",
      mid = "grey40",
      high = "red",
      midpoint = 0,
      name = expression(Delta[adj] - Delta[unadj]),
      guide = guide_colorbar(
        title.position = "top",
        barheight = unit(4.2, "cm"),
        barwidth = unit(0.35, "cm")
      )
    ) +
    scale_shape_manual(
      values = c(
        "both" = 16,
        "unadjusted_only" = 17,
        "adjusted_only" = 15,
        "none" = 1
      ),
      labels = c(
        "both" = "Both significant",
        "unadjusted_only" = "Unadjusted only",
        "adjusted_only" = "Adjusted only",
        "none" = "Neither significant"
      ),
      name = "p < 0.05"
    ) +
    coord_equal(xlim = axis_lim, ylim = axis_lim) +
    theme_minimal(base_size = 12) +
    theme(
      legend.position = "right",
      legend.box = "vertical",
      legend.title = element_text(size = 9),
      legend.text = element_text(size = 8),
      panel.grid.minor = element_blank(),
      plot.title = element_text(size = 13),
      plot.subtitle = element_text(size = 9)
    ) +
    labs(
      title = title,
      subtitle = "Red: adjusted > unadjusted; blue: adjusted < unadjusted; grey: near zero difference",
      x = "Unadjusted time effect",
      y = "Lymphocyte-adjusted time effect"
    )

  if (facet_by_component_set && "component_set" %in% colnames(adjusted_vs_unadjusted)) {
    p + facet_grid(component_set ~ timepoint)
  } else {
    p + facet_wrap(~timepoint, ncol = facet_ncol)
  }
}

save_adjusted_vs_unadjusted_pdf <- function(filename,
                                            plot,
                                            width = 8.5,
                                            height = 4.4) {
  grDevices::cairo_pdf(filename, width = width, height = height)
  print(plot)
  dev.off()
}
```

```{r decomposition-adjusted-vs-unadjusted, fig.width=11, fig.height=5.5}
adjusted_vs_unadjusted <- prepare_adjusted_vs_unadjusted(
  timepoint_effect_sizes,
  label_col = "clock"
)

p_adjusted_vs_unadjusted <- plot_adjusted_vs_unadjusted(
  adjusted_vs_unadjusted,
  title = "Clock time effects before and after lymphocyte adjustment",
  facet_ncol = 2,
  label_size = 2.0,
  point_size = 2.1
)

p_adjusted_vs_unadjusted
```

# PCA Analysis on Delta SRAA
## PCA Loadings Visualization

```{r pca-loadings-function}
run_pca_loadings <- function(data_wide, timepoint_value,
                             clock_suffix = "_RAA_z_base_delta",
                             id_col = "ID",
                             time_col = "Timepoint") {

  df_tp <- data_wide %>%
    filter(.data[[time_col]] == timepoint_value)

  delta_mat <- df_tp %>%
    select(ends_with(clock_suffix)) %>%
    as.data.frame()

  rownames(delta_mat) <- df_tp[[id_col]]

  pca <- prcomp(delta_mat, center = TRUE, scale. = TRUE)

  # variance explained
  var_expl <- (pca$sdev^2) / sum(pca$sdev^2)

  load_df <- as.data.frame(pca$rotation) %>%
    rownames_to_column("clock") %>%
    mutate(
      timepoint = timepoint_value,
      var_PC1 = var_expl[1],
      var_PC2 = var_expl[2],
      var_PC3 = var_expl[3]
    )

  list(pca = pca, loadings = load_df, var_expl = var_expl)
}

# Run PCA for both timepoints
res_d10 <- run_pca_loadings(data_wide, timepoint_value = "D+10")
res_m1  <- run_pca_loadings(data_wide, timepoint_value = "M+1")

# Combine loadings with metadata
loadings_all <- bind_rows(res_d10$loadings, res_m1$loadings) %>%
  left_join(clock_meta, by = "clock")
```

```{r pca-loadings-pc12, fig.width=10, fig.height=5}
# Axis limits shared across panels
x_lim <- range(loadings_all$PC1, na.rm = TRUE)
pc12_y_lim <- range(loadings_all$PC2, na.rm = TRUE)
pc12_y_span <- diff(pc12_y_lim)
pc12_plot_y_lim <- c(pc12_y_lim[1], pc12_y_lim[2] + 0.22 * pc12_y_span)

pc12_label_positions <- loadings_all %>%
  distinct(timepoint, var_PC1, var_PC2) %>%
  mutate(
    label_x = x_lim[1] + 0.02 * diff(x_lim),
    label_y = pc12_y_lim[2] + 0.18 * pc12_y_span
  )

p_loadings12 <- ggplot(loadings_all, aes(x = PC1, y = PC2, color = generation)) +
  geom_hline(yintercept = 0, linewidth = 0.3, alpha = 0.5) +
  geom_vline(xintercept = 0, linewidth = 0.3, alpha = 0.5) +
  geom_point(size = 3) +
  geom_text_repel(
    aes(label = gsub("_RAA_z_base_delta", "", clock)),
    size = 3, max.overlaps = Inf, show.legend = FALSE
  ) +
  facet_wrap(~timepoint, nrow = 1) +
  coord_cartesian(xlim = x_lim, ylim = pc12_plot_y_lim) +
  theme_minimal(base_size = 12) +
  labs(
    title = expression("PCA of " * Delta * "SRAA: clock loading structure at D+10 and M+1"),
    x = "PC1 loading",
    y = "PC2 loading",
    color = "Generation"
  ) +
  geom_text(
    data = pc12_label_positions,
    aes(
      x = label_x,
      y = label_y,
      label = paste0(
        "Var(PC1)=", sprintf("%.1f", 100 * var_PC1), "%\n",
        "Var(PC2)=", sprintf("%.1f", 100 * var_PC2), "%"
      )
    ),
    hjust = 0,
    vjust = 1,
    lineheight = 0.95,
    color = "black",
    inherit.aes = FALSE,
    size = 3
  )

p_loadings12
```

```{r pca-loadings-pc13, fig.width=10, fig.height=5}
pc13_y_lim <- range(loadings_all$PC3, na.rm = TRUE)
pc13_y_span <- diff(pc13_y_lim)
pc13_plot_y_lim <- c(pc13_y_lim[1], pc13_y_lim[2] + 0.22 * pc13_y_span)

pc13_label_positions <- loadings_all %>%
  distinct(timepoint, var_PC1, var_PC3) %>%
  mutate(
    label_x = x_lim[1] + 0.02 * diff(x_lim),
    label_y = pc13_y_lim[2] + 0.18 * pc13_y_span
  )

p_loadings13 <- ggplot(loadings_all, aes(x = PC1, y = PC3, color = generation)) +
  geom_hline(yintercept = 0, linewidth = 0.3, alpha = 0.5) +
  geom_vline(xintercept = 0, linewidth = 0.3, alpha = 0.5) +
  geom_point(size = 3) +
  geom_text_repel(
    aes(label = gsub("_RAA_z_base_delta", "", clock)),
    size = 3, max.overlaps = Inf, show.legend = FALSE
  ) +
  facet_wrap(~timepoint, nrow = 1) +
  coord_cartesian(xlim = x_lim, ylim = pc13_plot_y_lim) +
  theme_minimal(base_size = 12) +
  labs(
    title = expression("PCA of " * Delta * "SRAA: clock loading structure at D+10 and M+1"),
    x = "PC1 loading",
    y = "PC3 loading",
    color = "Generation"
  ) +
  geom_text(
    data = pc13_label_positions,
    aes(
      x = label_x,
      y = label_y,
      label = paste0(
        "Var(PC1)=", sprintf("%.1f", 100 * var_PC1), "%\n",
        "Var(PC3)=", sprintf("%.1f", 100 * var_PC3), "%"
      )
    ),
    hjust = 0,
    vjust = 1,
    lineheight = 0.95,
    color = "black",
    inherit.aes = FALSE,
    size = 3
  )

p_loadings13
```

```{r pca-loadings-combined, fig.width=10, fig.height=10}
p_loadings_legend <- get_legend(
  p_loadings12 +
    theme(legend.position = "right")
)

p_loadings_panels <- plot_grid(
  p_loadings12 +
    labs(title = "PC1 vs PC2 clock loadings") +
    theme(legend.position = "none"),
  p_loadings13 +
    labs(title = "PC1 vs PC3 clock loadings") +
    theme(legend.position = "none"),
  labels = c("a", "b"),
  label_size = 20,
  ncol = 1,
  align = "v",
  axis = "lr"
)

p_loadings_pc12_pc13 <- plot_grid(
  p_loadings_panels,
  p_loadings_legend,
  ncol = 2,
  rel_widths = c(1, 0.16),
  align = "h",
  axis = "tb"
)

p_loadings_pc12_pc13
```

```{r pca-3d-plot}
# Prepare 3D loadings data
load_3d <- loadings_all %>%
  select(clock, timepoint, PC1, PC2, PC3, generation) %>%
  filter(timepoint == "D+10")

# Color palette
gen_colors <- c(
  "Gen1" = "#F8766D",
  "Gen2+" = "#B79F00",
  "Telomere" = "#00BFC4",
  "Causal" = "#C77CFF"
)

p_3d <- plot_ly(
  data = load_3d,
  x = ~PC1,
  y = ~PC2,
  z = ~PC3,
  color = ~generation,
  colors = gen_colors,
  type = "scatter3d",
  mode = "markers+text",
  text = ~gsub("_RAA_z_base_delta","",clock),
  textposition = "top center",
  marker = list(size = 6)
) %>%
  layout(
    title = "3D PCA of ΔSRAA clock loadings (PC1–PC3)",
    scene = list(
      xaxis = list(title = "PC1 loading"),
      yaxis = list(title = "PC2 loading"),
      zaxis = list(title = "PC3 loading")
    ),
    legend = list(title = list(text = "Clock generation"))
  )

p_3d
```

## PC Scores and Clinical Associations

```{r pc-scores}
extract_scores <- function(pca_obj, timepoint_value) {
  as.data.frame(pca_obj$x) %>%
    rownames_to_column("ID") %>%
    mutate(Timepoint = timepoint_value) 
}

scores_all <- bind_rows(
  extract_scores(res_d10$pca, "D+10"),
  extract_scores(res_m1$pca,  "M+1")
)

# Join to your clinical data (ensure clinical_df has ID and Timepoint)
clinical_df <- meta %>%
  mutate(
    Timepoint_raw = Timepoint,
    Timepoint = recode(
      Timepoint,
      "BEFORE" = "D-1",
      "AFTER"  = "D+10",
      "FUP1"   = "M+1"
    )
  )

scores_all$ID <- as.integer(scores_all$ID)
df_assoc <- scores_all %>%
  left_join(clinical_df, by = c("ID", "Timepoint"))
```



```{r}
test_spearman_assoc <- function(df, x_col, vars, panel_name, min_n = 5) {

  res <- lapply(vars, function(v) {

    # variable exists?
    if (!v %in% colnames(df)) return(NULL)

    x <- df[[x_col]]
    y <- df[[v]]

    # paired non-missing
    ok <- is.finite(x) & is.finite(y)

    # enough data?
    if (sum(ok) < min_n) return(NULL)

    # non-zero variance?
    if (sd(x[ok]) == 0 || sd(y[ok]) == 0) return(NULL)

    # safe correlation
    ct <- tryCatch(
      cor.test(x[ok], y[ok], method = "spearman", exact = FALSE),
      error = function(e) NULL
    )

    if (is.null(ct)) return(NULL)

    data.frame(
      panel   = panel_name,
      variable = v,
      n        = sum(ok),
      rho      = unname(ct$estimate),
      p_value  = ct$p.value
    )
  }) %>% bind_rows()

  # if nothing passed filters, return NULL explicitly
  if (is.null(res) || nrow(res) == 0) return(NULL)

  res %>% mutate(p_adj = p.adjust(p_value, method = "BH"))
}

test_pc_assoc <- function(df, pc_col, vars, panel_name, min_n = 5) {
  test_spearman_assoc(df, pc_col, vars, panel_name, min_n = min_n)
}

run_assoc_for_timepoint <- function(df_assoc, timepoint_value, pc_col) {
  df_tp <- df_assoc %>% filter(Timepoint == timepoint_value)

  bind_rows(
    test_pc_assoc(df_tp, pc_col, vars_metabolic, "Metabolic"),
    test_pc_assoc(df_tp, pc_col, vars_stress,    "Stress"),
    test_pc_assoc(df_tp, pc_col, vars_hematology,"Hematology"),
    test_pc_assoc(df_tp, pc_col, vars_subject,   "Symptoms")
  ) %>% mutate(Timepoint = timepoint_value, PC = pc_col)
}

assoc_results <- bind_rows(
  run_assoc_for_timepoint(df_assoc, "D+10", "PC1"),
  run_assoc_for_timepoint(df_assoc, "D+10", "PC2"),
  run_assoc_for_timepoint(df_assoc, "D+10", "PC3"),
  run_assoc_for_timepoint(df_assoc, "M+1",  "PC1"),
  run_assoc_for_timepoint(df_assoc, "M+1",  "PC2"),
  run_assoc_for_timepoint(df_assoc, "M+1",  "PC3")
)

render_table(
  assoc_results %>%
    filter(p_adj < 0.05) %>%
    mutate(
      across(c(rho, p_value, p_adj), ~ signif(.x, 3))
    ),
  caption = "PC-clinical associations passing BH-adjusted p < 0.05"
)
```

## Association between DamAge/AdaptAge and glucose, Insulin and Homa

```{r}
# Helper to compute Spearman stats per facet and prepare annotation positions
facet_stats <- function(df, x_col) {
  df %>%
    group_by(variable) %>%
    summarise(
      n = sum(is.finite(.data[[x_col]]) & is.finite(value)),
      rho = suppressWarnings(cor(.data[[x_col]], value, method = "spearman", use = "complete.obs")),
      p_value = suppressWarnings(cor.test(.data[[x_col]], value, method = "spearman", exact = FALSE)$p.value),
      .groups = "drop"
    ) %>%
    mutate(
      label = paste0("n=", n,
                     "\n", "\u03C1=", sprintf("%.2f", rho),
                     "\n", "p=", format.pval(p_value, digits = 2, eps = 1e-3))
    )
}

# Helper for plotting
plot_metabolic_assoc <- function(df_long, x_col, x_label, title_text) {

  stats_df <- facet_stats(df_long, x_col)

  ggplot(df_long, aes(x = .data[[x_col]], y = value)) +
    geom_point(alpha = 0.8, size = 2) +
    geom_smooth(method = "lm", se = TRUE, linewidth = 0.8) +
    facet_wrap(~variable, scales = "free_y") +
    # place annotation in top-right of each facet
    geom_text(
      data = stats_df,
      aes(x = Inf, y = Inf, label = label),
      inherit.aes = FALSE,
      hjust = 1.05, vjust = 1.15,
      size = 3.2
    ) +
    theme_minimal(base_size = 12) +
    labs(
      title = title_text,
      x = x_label,
      y = NULL
    )
}

```

```{r metabolic-associations}
# 1) Join + filter
data_prepared <- clinical_df %>%
  select(Sample, ID, glucose, Insulin, Homa) %>%
  left_join(
    data_wide %>%
      select(Sample, Timepoint, AdaptAge_RAA, DamAge_RAA, CausAge_RAA, AdaptAge_RAA_z_base_delta, DamAge_RAA_z_base_delta, CausAge_RAA_z_base_delta),
    by = "Sample"
  )
```

### D+10
```{r}
# 2a) Delta metabolic variables (D+10 vs D-1)
metabolic_delta <- data_prepared %>%
  filter(Timepoint %in% c("D-1", "D+10")) %>%
  pivot_wider(
    id_cols = ID,
    names_from = Timepoint,
    values_from = c(glucose, Insulin, Homa),
    names_sep = "_"
  ) %>%
  mutate(
    delta_glucose = `glucose_D+10` - `glucose_D-1`,
    delta_Insulin = `Insulin_D+10` - `Insulin_D-1`,
    delta_Homa = `Homa_D+10` - `Homa_D-1`
  )

# 2c) Get D+10 age acceleration deltas
age_accel_delta <- data_prepared %>%
  filter(Timepoint == "D+10") %>%
  select(ID, AdaptAge_RAA_z_base_delta, DamAge_RAA_z_base_delta)

# 3) Combine all into one dataset for testing
data_assoc <- age_accel_delta %>%
  left_join(metabolic_delta, by = "ID") 

data_assoc <- data_assoc %>% filter(ID!=10)
# 2) Long format (metabolic vars)
data_long <- data_assoc %>%
  pivot_longer(
    cols = c(`glucose_D-1`, `glucose_D+10`, 
             `Insulin_D-1`, `Insulin_D+10`, 
             `Homa_D-1`, `Homa_D+10`, 
             `delta_glucose`, `delta_Insulin`, `delta_Homa`),
    names_to = "variable",
    values_to = "value"
  ) %>%
  mutate(variable = factor(variable, levels = c(
    "glucose_D-1", "glucose_D+10", "delta_glucose",
    "Insulin_D-1", "Insulin_D+10", "delta_Insulin",
    "Homa_D-1", "Homa_D+10", "delta_Homa"
  )))
```

```{r}
dam_stats_d10 <- facet_stats(data_long, "DamAge_RAA_z_base_delta")
adapt_stats_d10 <- facet_stats(data_long, "AdaptAge_RAA_z_base_delta")

render_table(dam_stats_d10, caption = "D+10 DamAge metabolic association statistics")
render_table(adapt_stats_d10, caption = "D+10 AdaptAge metabolic association statistics")

# 5) Produce the two figures
p_dam1 <- plot_metabolic_assoc(
  df_long = data_long,
  x_col = "DamAge_RAA_z_base_delta",
  x_label = "ΔSRAA (DamAge)",
  title_text = "D+10: Metabolic markers vs DamAge ΔSRAA"
)

p_adapt1 <- plot_metabolic_assoc(
  df_long = data_long,
  x_col = "AdaptAge_RAA_z_base_delta",
  x_label = "ΔSRAA (AdaptAge)",
  title_text = "D+10: Metabolic markers vs AdaptAge ΔSRAA"
)

p_dam1
p_adapt1
```

### M+1
```{r}
# 2a) Delta metabolic variables (M+1 vs D-1)
metabolic_delta <- data_prepared %>%
  filter(Timepoint %in% c("D-1", "M+1")) %>%
  pivot_wider(
    id_cols = ID,
    names_from = Timepoint,
    values_from = c(glucose, Insulin, Homa),
    names_sep = "_"
  ) %>%
  mutate(
    delta_glucose = `glucose_M+1` - `glucose_D-1`,
    delta_Insulin = `Insulin_M+1` - `Insulin_D-1`,
    delta_Homa = `Homa_M+1` - `Homa_D-1`
  )

# 2c) Get M+1 age acceleration deltas
age_accel_delta <- data_prepared %>%
  filter(Timepoint == "M+1") %>%
  select(ID, AdaptAge_RAA_z_base_delta, DamAge_RAA_z_base_delta)

# 3) Combine all into one dataset for testing
data_assoc <- age_accel_delta %>%
  left_join(metabolic_delta, by = "ID") 

data_assoc <- data_assoc %>% filter(ID!=10)
# 2) Long format (metabolic vars)
data_long <- data_assoc %>%
  pivot_longer(
    cols = c(`glucose_D-1`, `glucose_M+1`, 
             `Insulin_D-1`, `Insulin_M+1`, 
             `Homa_D-1`, `Homa_M+1`, 
             `delta_glucose`, `delta_Insulin`, `delta_Homa`),
    names_to = "variable",
    values_to = "value"
  ) %>%
  mutate(variable = factor(variable, levels = c(
    "glucose_D-1", "glucose_M+1", "delta_glucose",
    "Insulin_D-1", "Insulin_M+1", "delta_Insulin",
    "Homa_D-1", "Homa_M+1", "delta_Homa"
  )))
```

```{r}
dam_stats_m1 <- facet_stats(data_long, "DamAge_RAA_z_base_delta")
adapt_stats_m1 <- facet_stats(data_long, "AdaptAge_RAA_z_base_delta")

render_table(dam_stats_m1, caption = "M+1 DamAge metabolic association statistics")
render_table(adapt_stats_m1, caption = "M+1 AdaptAge metabolic association statistics")

# 5) Produce the two figures
p_dam2 <- plot_metabolic_assoc(
  df_long = data_long,
  x_col = "DamAge_RAA_z_base_delta",
  x_label = "ΔSRAA (DamAge)",
  title_text = "M+1: Metabolic markers vs DamAge ΔSRAA"
)

p_adapt2 <- plot_metabolic_assoc(
  df_long = data_long,
  x_col = "AdaptAge_RAA_z_base_delta",
  x_label = "ΔSRAA (AdaptAge)",
  title_text = "M+1: Metabolic markers vs AdaptAge ΔSRAA"
)

p_dam2
p_adapt2
```

# Figure Saving

```{r save-main-figures}
# Save clinical association plots
grDevices::cairo_pdf("figures/metabolic_assoc_D10_DamAge.pdf", width = 8, height = 7)
print(p_dam1)
dev.off()
grDevices::cairo_pdf("figures/metabolic_assoc_D10_AdaptAge.pdf", width = 8, height = 7)
print(p_adapt1)
dev.off()

grDevices::cairo_pdf("figures/metabolic_assoc_M1_DamAge.pdf", width = 8, height = 7)
print(p_dam2)
dev.off()
grDevices::cairo_pdf("figures/metabolic_assoc_M1_AdaptAge.pdf", width = 8, height = 7)
print(p_adapt2)
dev.off()

# Save PCA loading plots
pdf("figures/clock_PCA_loadings_PC1-2_PC1-3.pdf", width = 10, height = 10)
plot(p_loadings_pc12_pc13)
dev.off()

save_adjusted_vs_unadjusted_pdf(
  "figures/adjusted_vs_unadjusted_clock_time_effects.pdf",
  p_adjusted_vs_unadjusted
)
```

# Clock Component Analysis

## GrimAge Components

```{r grimage-components}
# PCGrimAge components
pc_grimage_cols <- c(
  "PCPACKYRS", "PCADM", "PCB2M", "PCCystatinC",
  "PCGDF15", "PCLeptin", "PCPAI1", "PCTIMP1"
)
res_pc_grimage <- main(df_age, pc_grimage_cols, covariates = c("Gender", "LYMPHOZYTEN"))
res_pc_grimage_no_cell <- main(df_age, pc_grimage_cols, covariates = "Gender")

# GrimAge v1 components
grimage1_cols <- paste0("GrimAge", c(
  "DNAmPACKYRS", "DNAmADM", "DNAmB2M", "DNAmCystatinC",
  "DNAmGDF15", "DNAmLeptin", "DNAmPAI1", "DNAmTIMP1"))
# GrimAge v1 - remove "GrimAge" prefix
res1 <- main(df_age, grimage1_cols, 
             remove_prefix = TRUE, 
             prefix_patterns = "GrimAge", 
             covariates = c("Gender", "LYMPHOZYTEN"))
res1_no_cell <- main(df_age, grimage1_cols, 
                     remove_prefix = TRUE, 
                     prefix_patterns = "GrimAge",
                     covariates = "Gender")

# GrimAge v2 components
grimage2_cols <- paste0("GrimAge2", c(
  "DNAmPACKYRS", "DNAmADM", "DNAmB2M", "DNAmCystatinC",
  "DNAmGDF15", "DNAmLeptin", "DNAmPAI1", "DNAmTIMP1",
  "DNAmlogCRP", "DNAmlogA1C"))
res2 <- main(df_age, grimage2_cols, 
             remove_prefix = TRUE, 
             prefix_patterns = "GrimAge2", 
             covariates = c("Gender", "LYMPHOZYTEN"))
res2_no_cell <- main(df_age, grimage2_cols, 
                     remove_prefix = TRUE, 
                     prefix_patterns = "GrimAge2",
                     covariates = "Gender")
```

### GrimAge Component Effect Sizes

```{r grimage-component-effect-sizes, results='asis'}
grimage_component_time_effects <- bind_rows(
  format_component_time_effects(
    res_pc_grimage,
    res_pc_grimage_no_cell,
    component_set = "PCGrimAge"
  ),
  format_component_time_effects(
    res1,
    res1_no_cell,
    component_set = "GrimAge",
    prefix_patterns = "GrimAge"
  ),
  format_component_time_effects(
    res2,
    res2_no_cell,
    component_set = "GrimAge v2",
    prefix_patterns = "GrimAge2"
  )
) %>%
  arrange_adjustment_pairs(group_cols = c("component_set", "component", "timepoint")) %>%
  mutate(
    p_value = signif(p_value, 3),
    across(c(effect_size, lower_95, upper_95, SE, df), ~ round(.x, 3))
  )

grimage_component_blood_cell_effects <- bind_rows(
  format_component_blood_cell_effects(
    res_pc_grimage,
    pc_grimage_cols,
    component_set = "PCGrimAge"
  ),
  format_component_blood_cell_effects(
    res1,
    grimage1_cols,
    component_set = "GrimAge",
    prefix_patterns = "GrimAge"
  ),
  format_component_blood_cell_effects(
    res2,
    grimage2_cols,
    component_set = "GrimAge v2",
    prefix_patterns = "GrimAge2"
  )
) %>%
  arrange(component_set, component, term) %>%
  mutate(
    p_value = signif(p_value, 3),
    across(c(effect_size, lower_95, upper_95, SE, df), ~ round(.x, 3))
  )

render_table(
  grimage_component_time_effects,
  caption = "GrimAge component timepoint effect sizes with and without blood-cell adjustment"
)

render_table(
  grimage_component_blood_cell_effects,
  caption = "Standardized blood-cell proportion effect sizes for GrimAge component models"
)

adjusted_vs_unadjusted_pc_grimage <- prepare_adjusted_vs_unadjusted(
  grimage_component_time_effects %>% filter(component_set == "PCGrimAge"),
  label_col = "component"
)

adjusted_vs_unadjusted_grimage1 <- prepare_adjusted_vs_unadjusted(
  grimage_component_time_effects %>% filter(component_set == "GrimAge"),
  label_col = "component"
)

adjusted_vs_unadjusted_grimage2 <- prepare_adjusted_vs_unadjusted(
  grimage_component_time_effects %>% filter(component_set == "GrimAge v2"),
  label_col = "component"
)

p_adjusted_vs_unadjusted_pc_grimage <- plot_adjusted_vs_unadjusted(
  adjusted_vs_unadjusted_pc_grimage,
  title = "PCGrimAge component time effects before and after lymphocyte adjustment",
  facet_ncol = 2,
  label_size = 2.3,
  point_size = 2
)

p_adjusted_vs_unadjusted_grimage1 <- plot_adjusted_vs_unadjusted(
  adjusted_vs_unadjusted_grimage1,
  title = "GrimAge component time effects before and after lymphocyte adjustment",
  facet_ncol = 2,
  label_size = 2.3,
  point_size = 2
)

p_adjusted_vs_unadjusted_grimage2 <- plot_adjusted_vs_unadjusted(
  adjusted_vs_unadjusted_grimage2,
  title = "GrimAge v2 component time effects before and after lymphocyte adjustment",
  facet_ncol = 2,
  label_size = 2.2,
  point_size = 2
)
```

### GrimAge Component Trajectory Plots

```{r grimage-component-plots-display, fig.width=10, fig.height=6}
res_pc_grimage$p1 + labs(title = "PCGrimAge SRAA trajectories (composition adjusted)", x = NULL)
res_pc_grimage$p2 + labs(title = "PCGrimAge within-subject effect sizes (composition adjusted)", x = NULL)
res_pc_grimage_no_cell$p1 + labs(title = "PCGrimAge SRAA trajectories", x = NULL)
res_pc_grimage_no_cell$p2 + labs(title = "PCGrimAge within-subject effect sizes", x = NULL)

res1$p1 + labs(title = "GrimAge SRAA trajectories (composition adjusted)", x = NULL)
res1$p2 + labs(title = "GrimAge within-subject effect sizes (composition adjusted)", x = NULL)
res1_no_cell$p1 + labs(title = "GrimAge SRAA trajectories", x = NULL)
res1_no_cell$p2 + labs(title = "GrimAge within-subject effect sizes", x = NULL)

res2$p1 + labs(title = "GrimAge v2 SRAA trajectories (composition adjusted)", x = NULL)
res2$p2 + labs(title = "GrimAge v2 within-subject effect sizes (composition adjusted)", x = NULL)
res2_no_cell$p1 + labs(title = "GrimAge v2 SRAA trajectories", x = NULL)
res2_no_cell$p2 + labs(title = "GrimAge v2 within-subject effect sizes", x = NULL)
```

### GrimAge Component Adjusted-vs-Unadjusted Plots

```{r grimage-component-adjusted-vs-unadjusted, fig.width=11, fig.height=5.5}
p_adjusted_vs_unadjusted_pc_grimage
p_adjusted_vs_unadjusted_grimage1
p_adjusted_vs_unadjusted_grimage2
```

### GrimAge Component Decomposition

```{r grimage-component-decomposition, fig.width=10, fig.height=6, results='asis'}
decomp_pc_grimage <- decompose_clock_composition(
  res_pc_grimage$data_wide,
  pc_grimage_cols,
  cell_covariates = "LYMPHOZYTEN",
  adjustment_covariates = "Gender"
)

decomp_grimage1 <- decompose_clock_composition(
  res1$data_wide,
  grimage1_cols,
  cell_covariates = "LYMPHOZYTEN",
  adjustment_covariates = "Gender"
)

decomp_grimage2 <- decompose_clock_composition(
  res2$data_wide,
  grimage2_cols,
  cell_covariates = "LYMPHOZYTEN",
  adjustment_covariates = "Gender"
)

decomp_pc_grimage_outputs <- prepare_decomposition_outputs(
  decomp_pc_grimage,
  pc_grimage_cols,
  title_prefix = "PCGrimAge"
)

decomp_grimage1_outputs <- prepare_decomposition_outputs(
  decomp_grimage1,
  grimage1_cols,
  prefix_patterns = "GrimAge",
  title_prefix = "GrimAge"
)

decomp_grimage2_outputs <- prepare_decomposition_outputs(
  decomp_grimage2,
  grimage2_cols,
  prefix_patterns = "GrimAge2",
  title_prefix = "GrimAge v2"
)

grimage_component_decomposition_effects <- bind_rows(
  decomp_pc_grimage_outputs$effects %>% mutate(component_set = "PCGrimAge"),
  decomp_grimage1_outputs$effects %>% mutate(component_set = "GrimAge"),
  decomp_grimage2_outputs$effects %>% mutate(component_set = "GrimAge v2")
) %>%
  transmute(
    component_set,
    component_name = clock_label,
    decomposition_component = component,
    timepoint,
    effect_size,
    lower_95,
    upper_95,
    SE,
    df,
    p_value
  ) %>%
  mutate(
    p_value = signif(p_value, 3),
    across(c(effect_size, lower_95, upper_95, SE, df), ~ round(.x, 3))
  )

render_table(
  grimage_component_decomposition_effects,
  caption = "GrimAge component decomposition timepoint effects"
)
```

```{r grimage-component-decomposition-plots, fig.width=10, fig.height=6}
decomp_pc_grimage_outputs$spaghetti
decomp_pc_grimage_outputs$raincloud

decomp_grimage1_outputs$spaghetti
decomp_grimage1_outputs$raincloud

decomp_grimage2_outputs$spaghetti
decomp_grimage2_outputs$raincloud
```

### GrimAge Component Plots

```{r save-component-figures}
grDevices::cairo_pdf("figures/SFig4-7_GrimAge_family_components_blood_adjustment.pdf", width = 8.5, height = 10)
plot_grid(
  res_pc_grimage$p2 + labs(title = "PCGrimAge components (blood-cell adjusted)", x = NULL),
  res1$p2 + labs(title = "GrimAge components (blood-cell adjusted)", x = NULL),
  res2$p2 + labs(title = "GrimAge v2 components (blood-cell adjusted)", x = NULL),
  labels = c("a", "b", "c"),
  label_size = 20,
  ncol = 1,
  rel_heights = c(1, 1, 1)
)
dev.off()

save_adjusted_vs_unadjusted_pdf(
  "figures/adjusted_vs_unadjusted_PCGrimAge_components.pdf",
  p_adjusted_vs_unadjusted_pc_grimage
)

save_adjusted_vs_unadjusted_pdf(
  "figures/adjusted_vs_unadjusted_GrimAge_components.pdf",
  p_adjusted_vs_unadjusted_grimage1
)

save_adjusted_vs_unadjusted_pdf(
  "figures/adjusted_vs_unadjusted_GrimAge2_components.pdf",
  p_adjusted_vs_unadjusted_grimage2
)

# GrimAge component decomposition figures
grDevices::cairo_pdf("figures/decomposition_PCGrimAge_components_spaghetti.pdf", width = 10, height = 5.5)
print(decomp_pc_grimage_outputs$spaghetti)
dev.off()

grDevices::cairo_pdf("figures/decomposition_PCGrimAge_components_raincloud.pdf", width = 10, height = 5.5)
print(decomp_pc_grimage_outputs$raincloud)
dev.off()

grDevices::cairo_pdf("figures/decomposition_GrimAge_components_spaghetti.pdf", width = 10, height = 5.5)
print(decomp_grimage1_outputs$spaghetti)
dev.off()

grDevices::cairo_pdf("figures/decomposition_GrimAge_components_raincloud.pdf", width = 10, height = 5.5)
print(decomp_grimage1_outputs$raincloud)
dev.off()

grDevices::cairo_pdf("figures/decomposition_GrimAge2_components_spaghetti.pdf", width = 11, height = 5.5)
print(decomp_grimage2_outputs$spaghetti)
dev.off()

grDevices::cairo_pdf("figures/decomposition_GrimAge2_components_raincloud.pdf", width = 11, height = 5.5)
print(decomp_grimage2_outputs$raincloud)
dev.off()
```

## SystemsAge Components

```{r systemsage-components}
systemsage_cols <- c(
  "Blood", "Brain", "Inflammation", "Heart", "Hormone",
  "Immune", "Kidney", "Liver", "Metabolic", "Lung", "MusculoSkeletal"
)

ress <- main(df_age, systemsage_cols, covariates = c("Gender", "LYMPHOZYTEN"))
ress_no_cell <- main(df_age, systemsage_cols, covariates = "Gender")
```

### SystemsAge Component Effect Sizes

```{r systemsage-component-effect-sizes, results='asis'}
systemsage_component_time_effects <- format_component_time_effects(
  ress,
  ress_no_cell,
  component_set = "SystemsAge"
) %>%
  arrange_adjustment_pairs(group_cols = c("component_set", "component", "timepoint")) %>%
  mutate(
    p_value = signif(p_value, 3),
    across(c(effect_size, lower_95, upper_95, SE, df), ~ round(.x, 3))
  )

systemsage_component_blood_cell_effects <- format_component_blood_cell_effects(
  ress,
  systemsage_cols,
  component_set = "SystemsAge"
) %>%
  arrange(component_set, component, term) %>%
  mutate(
    p_value = signif(p_value, 3),
    across(c(effect_size, lower_95, upper_95, SE, df), ~ round(.x, 3))
  )

render_table(
  systemsage_component_time_effects,
  caption = "SystemsAge component timepoint effect sizes with and without blood-cell adjustment"
)

render_table(
  systemsage_component_blood_cell_effects,
  caption = "Standardized blood-cell proportion effect sizes for SystemsAge component models"
)

adjusted_vs_unadjusted_systemsage <- prepare_adjusted_vs_unadjusted(
  systemsage_component_time_effects,
  label_col = "component"
)

p_adjusted_vs_unadjusted_systemsage <- plot_adjusted_vs_unadjusted(
  adjusted_vs_unadjusted_systemsage,
  title = "SystemsAge component time effects before and after lymphocyte adjustment",
  facet_ncol = 2,
  label_size = 2.1,
  point_size = 2
)
```

### SystemsAge Component Trajectory Plots

```{r systemsage-component-plots-display, fig.width=10, fig.height=6}
ress$p1 + labs(title = "SystemsAge SRAA trajectories (composition adjusted)", x = NULL)
ress$p2 + labs(title = "SystemsAge within-subject effect sizes (composition adjusted)", x = NULL)
ress_no_cell$p1 + labs(title = "SystemsAge SRAA trajectories", x = NULL)
ress_no_cell$p2 + labs(title = "SystemsAge within-subject effect sizes", x = NULL)
```

### SystemsAge Component Adjusted-vs-Unadjusted Plot

```{r systemsage-component-adjusted-vs-unadjusted, fig.width=11, fig.height=5.5}
p_adjusted_vs_unadjusted_systemsage
```

### SystemsAge Component Decomposition

```{r systemsage-component-decomposition, fig.width=11, fig.height=6, results='asis'}
decomp_systemsage <- decompose_clock_composition(
  ress$data_wide,
  systemsage_cols,
  cell_covariates = "LYMPHOZYTEN",
  adjustment_covariates = "Gender"
)

decomp_systemsage_outputs <- prepare_decomposition_outputs(
  decomp_systemsage,
  systemsage_cols,
  title_prefix = "SystemsAge"
)

systemsage_component_decomposition_effects <- decomp_systemsage_outputs$effects %>%
  transmute(
    component_set = "SystemsAge",
    component_name = clock_label,
    decomposition_component = component,
    timepoint,
    effect_size,
    lower_95,
    upper_95,
    SE,
    df,
    p_value
  ) %>%
  mutate(
    p_value = signif(p_value, 3),
    across(c(effect_size, lower_95, upper_95, SE, df), ~ round(.x, 3))
  )

render_table(
  systemsage_component_decomposition_effects,
  caption = "SystemsAge component decomposition timepoint effects"
)

decomp_systemsage_outputs$spaghetti
decomp_systemsage_outputs$raincloud
```

### SystemsAge Component Plots

```{r save-systemsage-figure}
save_adjusted_vs_unadjusted_pdf(
  "figures/adjusted_vs_unadjusted_SystemsAge_components.pdf",
  p_adjusted_vs_unadjusted_systemsage
)

grDevices::cairo_pdf("figures/decomposition_SystemsAge_components_spaghetti.pdf", width = 11, height = 5.5)
print(decomp_systemsage_outputs$spaghetti)
dev.off()

grDevices::cairo_pdf("figures/decomposition_SystemsAge_components_raincloud.pdf", width = 11, height = 5.5)
print(decomp_systemsage_outputs$raincloud)
dev.off()
```

## Figure 4 and Supplementary Figure 4 Exports

```{r save-reorganized-figure-4}
# Figure 4: unadjusted within-subject raincloud plots
grDevices::cairo_pdf("figures/Fig4.pdf", width = 8.5, height = 11)
print(
  plot_grid(
    main_res_no_cell$p2 +
      labs(title = "Within-subject effect sizes", x = NULL),
    res_pc_grimage_no_cell$p2 +
      labs(title = "PCGrimAge components", x = NULL),
    ress_no_cell$p2 +
      labs(title = "SystemsAge components", x = NULL),
    labels = c("a", "b", "c"),
    label_size = 20,
    ncol = 1,
    rel_heights = c(1, 1, 1)
  )
)
dev.off()

# Figure 4 switched version: adjusted within-subject raincloud plots
grDevices::cairo_pdf("figures/Fig4_blood_cell_adjusted.pdf", width = 8.5, height = 11)
print(
  plot_grid(
    main_res$p2 +
      labs(title = "Within-subject effect sizes (blood-cell adjusted)", x = NULL),
    res_pc_grimage$p2 +
      labs(title = "PCGrimAge components (blood-cell adjusted)", x = NULL),
    ress$p2 +
      labs(title = "SystemsAge components (blood-cell adjusted)", x = NULL),
    labels = c("a", "b", "c"),
    label_size = 20,
    ncol = 1,
    rel_heights = c(1, 1, 1)
  )
)
dev.off()

# Supplementary Figure 4-1: adjusted main clocks and adjusted-vs-unadjusted effects
grDevices::cairo_pdf("figures/SFig4-1_main_clocks_blood_adjustment.pdf", width = 10, height = 9)
print(
  plot_grid(
    main_res$p2 +
      labs(title = "Within-subject effect sizes (blood-cell adjusted)", x = NULL),
    p_adjusted_vs_unadjusted,
    labels = c("a", "b"),
    label_size = 20,
    ncol = 1,
    rel_heights = c(1, 1.15)
  )
)
dev.off()

# Supplementary Figure 4-1 switched version: unadjusted main clocks and adjusted-vs-unadjusted effects
grDevices::cairo_pdf("figures/SFig4-1_main_clocks_no_blood_adjustment.pdf", width = 10, height = 9)
print(
  plot_grid(
    main_res_no_cell$p2 +
      labs(title = "Within-subject effect sizes", x = NULL),
    p_adjusted_vs_unadjusted,
    labels = c("a", "b"),
    label_size = 20,
    ncol = 1,
    rel_heights = c(1, 1.15)
  )
)
dev.off()

# Supplementary Figure 4-2: adjusted PCGrimAge components and adjusted-vs-unadjusted effects
grDevices::cairo_pdf("figures/SFig4-2_PCGrimAge_blood_adjustment.pdf", width = 10, height = 9)
print(
  plot_grid(
    res_pc_grimage$p2 +
      labs(title = "PCGrimAge components (blood-cell adjusted)", x = NULL),
    p_adjusted_vs_unadjusted_pc_grimage,
    labels = c("a", "b"),
    label_size = 20,
    ncol = 1,
    rel_heights = c(1, 1.15)
  )
)
dev.off()

# Supplementary Figure 4-2 switched version: unadjusted PCGrimAge components and adjusted-vs-unadjusted effects
grDevices::cairo_pdf("figures/SFig4-2_PCGrimAge_no_blood_adjustment.pdf", width = 10, height = 9)
print(
  plot_grid(
    res_pc_grimage_no_cell$p2 +
      labs(title = "PCGrimAge components", x = NULL),
    p_adjusted_vs_unadjusted_pc_grimage,
    labels = c("a", "b"),
    label_size = 20,
    ncol = 1,
    rel_heights = c(1, 1.15)
  )
)
dev.off()

# Supplementary Figure 4-3: adjusted SystemsAge components and adjusted-vs-unadjusted effects
grDevices::cairo_pdf("figures/SFig4-3_SystemsAge_blood_adjustment.pdf", width = 10, height = 9)
print(
  plot_grid(
    ress$p2 +
      labs(title = "SystemsAge components (blood-cell adjusted)", x = NULL),
    p_adjusted_vs_unadjusted_systemsage,
    labels = c("a", "b"),
    label_size = 20,
    ncol = 1,
    rel_heights = c(1, 1.15)
  )
)
dev.off()

# Supplementary Figure 4-3 switched version: unadjusted SystemsAge components and adjusted-vs-unadjusted effects
grDevices::cairo_pdf("figures/SFig4-3_SystemsAge_no_blood_adjustment.pdf", width = 10, height = 9)
print(
  plot_grid(
    ress_no_cell$p2 +
      labs(title = "SystemsAge components", x = NULL),
    p_adjusted_vs_unadjusted_systemsage,
    labels = c("a", "b"),
    label_size = 20,
    ncol = 1,
    rel_heights = c(1, 1.15)
  )
)
dev.off()

# Decomposition plots corresponding to Figure 4 panels a-c
decomposition_sfig_main_title <- "Decomposition of clock within-subject changes from baseline"
decomposition_sfig_pc_grimage_title <- "Decomposition of clock within-subject changes from baseline for PCGrimAge components"
decomposition_sfig_systemsage_title <- "Decomposition of clock within-subject changes from baseline for SystemsAge components"

grDevices::cairo_pdf("figures/SFig4-4_panel_a_main_clocks_decomposition.pdf", width = 10, height = 5.5)
print(p_decomposition_raincloud + labs(title = decomposition_sfig_main_title))
dev.off()

grDevices::cairo_pdf("figures/SFig4-5_panel_b_PCGrimAge_decomposition.pdf", width = 10, height = 5.5)
print(decomp_pc_grimage_outputs$raincloud + labs(title = decomposition_sfig_pc_grimage_title))
dev.off()

grDevices::cairo_pdf("figures/SFig4-6_panel_c_SystemsAge_decomposition.pdf", width = 11, height = 5.5)
print(decomp_systemsage_outputs$raincloud + labs(title = decomposition_sfig_systemsage_title))
dev.off()

grDevices::cairo_pdf("figures/p_lymphocyte_delta.pdf", width = 6, height = 4)
print(p_lymphocyte_delta)
dev.off()
```

## Clock and Component Clinical Associations

```{r clock-component-clinical-associations, results='asis'}
run_feature_clinical_assoc <- function(data_wide,
                                       feature_cols,
                                       feature_set,
                                       feature_type,
                                       prefix_patterns = NULL,
                                       timepoints = c("D+10", "M+1")) {
  feature_delta_cols <- paste0(feature_cols, "_RAA_z_base_delta")
  feature_delta_cols <- feature_delta_cols[feature_delta_cols %in% colnames(data_wide)]

  if (length(feature_delta_cols) == 0) {
    return(tibble())
  }

  label_tbl <- tibble(
    feature_col = feature_delta_cols,
    feature = str_remove(feature_delta_cols, "_RAA_z_base_delta$")
  )

  if (!is.null(prefix_patterns)) {
    for (pattern in prefix_patterns) {
      label_tbl$feature <- str_remove(label_tbl$feature, paste0("^", pattern))
    }
  }

  df_feature_assoc <- data_wide %>%
    select(ID, Timepoint, all_of(feature_delta_cols)) %>%
    mutate(
      ID = as.integer(ID),
      Timepoint = as.character(Timepoint)
    ) %>%
    filter(Timepoint %in% timepoints) %>%
    left_join(clinical_df, by = c("ID", "Timepoint"))

  bind_rows(lapply(timepoints, function(timepoint_value) {
    df_tp <- df_feature_assoc %>% filter(Timepoint == timepoint_value)

    bind_rows(lapply(feature_delta_cols, function(feature_col) {
      bind_rows(
        test_spearman_assoc(df_tp, feature_col, vars_metabolic, "Metabolic"),
        test_spearman_assoc(df_tp, feature_col, vars_stress, "Stress"),
        test_spearman_assoc(df_tp, feature_col, vars_hematology, "Hematology"),
        test_spearman_assoc(df_tp, feature_col, vars_subject, "Symptoms")
      ) %>%
        mutate(
          Timepoint = timepoint_value,
          feature_col = feature_col
        )
    }))
  })) %>%
    left_join(label_tbl, by = "feature_col") %>%
    mutate(
      feature_set = feature_set,
      feature_type = feature_type
    ) %>%
    relocate(feature_set, feature_type, feature, feature_col, Timepoint, panel, variable)
}

clock_component_assoc_results <- bind_rows(
  run_feature_clinical_assoc(
    main_res$data_wide,
    clock_cols,
    feature_set = "Main clocks",
    feature_type = "Clock"
  ),
  run_feature_clinical_assoc(
    res_pc_grimage$data_wide,
    pc_grimage_cols,
    feature_set = "PCGrimAge",
    feature_type = "Clock component"
  ),
  run_feature_clinical_assoc(
    res1$data_wide,
    grimage1_cols,
    feature_set = "GrimAge",
    feature_type = "Clock component",
    prefix_patterns = "GrimAge"
  ),
  run_feature_clinical_assoc(
    res2$data_wide,
    grimage2_cols,
    feature_set = "GrimAge v2",
    feature_type = "Clock component",
    prefix_patterns = "GrimAge2"
  ),
  run_feature_clinical_assoc(
    ress$data_wide,
    systemsage_cols,
    feature_set = "SystemsAge",
    feature_type = "Clock component"
  )
) %>%
  mutate(p_adj_all_tests = p.adjust(p_value, method = "BH"))

render_table(
  clock_component_assoc_results %>%
    filter(p_adj_all_tests < 0.05) %>%
    mutate(
      across(c(rho, p_value, p_adj, p_adj_all_tests), ~ signif(.x, 3))
    ) %>%
    arrange(p_adj_all_tests, p_value),
  caption = "Clock and clock-component clinical associations passing global BH-adjusted p < 0.05"
)
```

## Clock and Component Change-Change Clinical Associations

```{r clock-component-clinical-change-associations, results='asis'}
clinical_assoc_vars <- unique(c(
  vars_metabolic,
  vars_stress,
  vars_hematology,
  vars_subject
))

make_clinical_delta_df <- function(clinical_df,
                                   vars,
                                   timepoints = c("D+10", "M+1"),
                                   baseline_label = "D-1") {
  vars_present <- vars[vars %in% colnames(clinical_df)]

  baseline_df <- clinical_df %>%
    mutate(
      ID = as.integer(ID),
      Timepoint = as.character(Timepoint)
    ) %>%
    filter(Timepoint == baseline_label) %>%
    select(ID, all_of(vars_present)) %>%
    rename_with(~ paste0(.x, "_baseline"), all_of(vars_present))

  delta_df <- clinical_df %>%
    mutate(
      ID = as.integer(ID),
      Timepoint = as.character(Timepoint)
    ) %>%
    filter(Timepoint %in% timepoints) %>%
    select(ID, Timepoint, all_of(vars_present)) %>%
    left_join(baseline_df, by = "ID")

  for (v in vars_present) {
    baseline_col <- paste0(v, "_baseline")
    delta_col <- paste0(v, "_delta_from_", str_replace_all(baseline_label, "[^A-Za-z0-9]+", ""))
    delta_df[[delta_col]] <- suppressWarnings(as.numeric(delta_df[[v]])) -
      suppressWarnings(as.numeric(delta_df[[baseline_col]]))
  }

  delta_df
}

clinical_delta_df <- make_clinical_delta_df(clinical_df, clinical_assoc_vars)

clinical_delta_var_map <- tibble(
  variable = clinical_assoc_vars
) %>%
  filter(variable %in% colnames(clinical_df)) %>%
  mutate(
    delta_variable = paste0(variable, "_delta_from_D1")
  )

test_spearman_delta_assoc <- function(df, x_col, vars, panel_name, min_n = 5) {
  delta_vars <- clinical_delta_var_map %>%
    filter(variable %in% vars, delta_variable %in% colnames(df))

  res <- lapply(seq_len(nrow(delta_vars)), function(i) {
    v <- delta_vars$variable[i]
    delta_v <- delta_vars$delta_variable[i]

    x <- suppressWarnings(as.numeric(df[[x_col]]))
    y <- suppressWarnings(as.numeric(df[[delta_v]]))
    ok <- is.finite(x) & is.finite(y)

    if (sum(ok) < min_n) return(NULL)
    if (sd(x[ok]) == 0 || sd(y[ok]) == 0) return(NULL)

    ct <- tryCatch(
      cor.test(x[ok], y[ok], method = "spearman", exact = FALSE),
      error = function(e) NULL
    )

    if (is.null(ct)) return(NULL)

    data.frame(
      panel = panel_name,
      variable = v,
      delta_variable = delta_v,
      n = sum(ok),
      rho = unname(ct$estimate),
      p_value = ct$p.value
    )
  }) %>% bind_rows()

  if (is.null(res) || nrow(res) == 0) return(NULL)

  res %>% mutate(p_adj = p.adjust(p_value, method = "BH"))
}

run_feature_clinical_delta_assoc <- function(data_wide,
                                             feature_cols,
                                             feature_set,
                                             feature_type,
                                             prefix_patterns = NULL,
                                             timepoints = c("D+10", "M+1")) {
  feature_delta_cols <- paste0(feature_cols, "_RAA_z_base_delta")
  feature_delta_cols <- feature_delta_cols[feature_delta_cols %in% colnames(data_wide)]

  if (length(feature_delta_cols) == 0) {
    return(tibble())
  }

  label_tbl <- tibble(
    feature_col = feature_delta_cols,
    feature = str_remove(feature_delta_cols, "_RAA_z_base_delta$")
  )

  if (!is.null(prefix_patterns)) {
    for (pattern in prefix_patterns) {
      label_tbl$feature <- str_remove(label_tbl$feature, paste0("^", pattern))
    }
  }

  df_feature_delta_assoc <- data_wide %>%
    select(ID, Timepoint, all_of(feature_delta_cols)) %>%
    mutate(
      ID = as.integer(ID),
      Timepoint = as.character(Timepoint)
    ) %>%
    filter(Timepoint %in% timepoints) %>%
    left_join(clinical_delta_df, by = c("ID", "Timepoint"))

  bind_rows(lapply(timepoints, function(timepoint_value) {
    df_tp <- df_feature_delta_assoc %>% filter(Timepoint == timepoint_value)

    bind_rows(lapply(feature_delta_cols, function(feature_col) {
      bind_rows(
        test_spearman_delta_assoc(df_tp, feature_col, vars_metabolic, "Metabolic"),
        test_spearman_delta_assoc(df_tp, feature_col, vars_stress, "Stress"),
        test_spearman_delta_assoc(df_tp, feature_col, vars_hematology, "Hematology"),
        test_spearman_delta_assoc(df_tp, feature_col, vars_subject, "Symptoms")
      ) %>%
        mutate(
          Timepoint = timepoint_value,
          feature_col = feature_col
        )
    }))
  })) %>%
    left_join(label_tbl, by = "feature_col") %>%
    mutate(
      feature_set = feature_set,
      feature_type = feature_type,
      clinical_change = paste0(Timepoint, " - D-1")
    ) %>%
    relocate(
      feature_set, feature_type, feature, feature_col,
      Timepoint, clinical_change, panel, variable, delta_variable
    )
}

clock_component_clinical_delta_assoc_results <- bind_rows(
  run_feature_clinical_delta_assoc(
    main_res$data_wide,
    clock_cols,
    feature_set = "Main clocks",
    feature_type = "Clock"
  ),
  run_feature_clinical_delta_assoc(
    res_pc_grimage$data_wide,
    pc_grimage_cols,
    feature_set = "PCGrimAge",
    feature_type = "Clock component"
  ),
  run_feature_clinical_delta_assoc(
    res1$data_wide,
    grimage1_cols,
    feature_set = "GrimAge",
    feature_type = "Clock component",
    prefix_patterns = "GrimAge"
  ),
  run_feature_clinical_delta_assoc(
    res2$data_wide,
    grimage2_cols,
    feature_set = "GrimAge v2",
    feature_type = "Clock component",
    prefix_patterns = "GrimAge2"
  ),
  run_feature_clinical_delta_assoc(
    ress$data_wide,
    systemsage_cols,
    feature_set = "SystemsAge",
    feature_type = "Clock component"
  )
) %>%
  mutate(p_adj_all_tests = p.adjust(p_value, method = "BH"))

render_table(
  clock_component_clinical_delta_assoc_results %>%
    filter(p_adj_all_tests < 0.05) %>%
    mutate(
      across(c(rho, p_value, p_adj, p_adj_all_tests), ~ signif(.x, 3))
    ) %>%
    arrange(p_adj_all_tests, p_value),
  caption = "Clock and clock-component change-change clinical associations passing global BH-adjusted p < 0.05"
)
```

# Save Results

```{r save-all-results}
# Save all plots as RDS
plots <- list(
  p_loadings12 = p_loadings12,
  p_loadings13 = p_loadings13,
  p_spaghetti = main_res$p1,
  p_raincloud = main_res$p2,
  p_raincloud_no_cell_adjustment = main_res_no_cell$p2,
  p_adjusted_vs_unadjusted = p_adjusted_vs_unadjusted,
  main_effects = main_res$effects,
  main_effects_no_cell_adjustment = main_res_no_cell$effects,
  p_dam_D10 = p_dam1,
  p_adapt_D10 = p_adapt1,
  p_dam_M1 = p_dam2,
  p_adapt_M1 = p_adapt2,
  res_GrimAge = res1,
  res_GrimAge_no_cell_adjustment = res1_no_cell,
  res_GrimAge2 = res2,
  res_GrimAge2_no_cell_adjustment = res2_no_cell,
  res_PCGrimAge = res_pc_grimage,
  res_PCGrimAge_no_cell_adjustment = res_pc_grimage_no_cell,
  res_SystemsAge = ress,
  res_SystemsAge_no_cell_adjustment = ress_no_cell,
  grimage_component_time_effects = grimage_component_time_effects,
  grimage_component_blood_cell_effects = grimage_component_blood_cell_effects,
  p_adjusted_vs_unadjusted_PCGrimAge = p_adjusted_vs_unadjusted_pc_grimage,
  p_adjusted_vs_unadjusted_GrimAge = p_adjusted_vs_unadjusted_grimage1,
  p_adjusted_vs_unadjusted_GrimAge2 = p_adjusted_vs_unadjusted_grimage2,
  grimage_component_decomposition_effects = grimage_component_decomposition_effects,
  decomp_PCGrimAge = decomp_pc_grimage_outputs,
  decomp_GrimAge = decomp_grimage1_outputs,
  decomp_GrimAge2 = decomp_grimage2_outputs,
  systemsage_component_time_effects = systemsage_component_time_effects,
  systemsage_component_blood_cell_effects = systemsage_component_blood_cell_effects,
  p_adjusted_vs_unadjusted_SystemsAge = p_adjusted_vs_unadjusted_systemsage,
  systemsage_component_decomposition_effects = systemsage_component_decomposition_effects,
  decomp_SystemsAge = decomp_systemsage_outputs,
  assoc_results = assoc_results,
  clock_component_assoc_results = clock_component_assoc_results,
  clock_component_clinical_delta_assoc_results = clock_component_clinical_delta_assoc_results
)
saveRDS(plots, file = "figures/epiAge_plots.rds")

# Save association results
write.table(
  assoc_results,
  file = "figures/epiAge_PC_association_results.csv",
  sep = ",",
  row.names = FALSE,
  quote = FALSE
)

write_xlsx(
  assoc_results,
  "figures/epiAge_PC_association_results.xlsx",
  col_names = TRUE,
  format_headers = TRUE
)

write.table(
  clock_component_assoc_results,
  file = "figures/clock_component_clinical_association_results.csv",
  sep = ",",
  row.names = FALSE,
  quote = FALSE
)

write_xlsx(
  clock_component_assoc_results,
  "figures/clock_component_clinical_association_results.xlsx",
  col_names = TRUE,
  format_headers = TRUE
)

write.table(
  clock_component_clinical_delta_assoc_results,
  file = "figures/clock_component_clinical_delta_association_results.csv",
  sep = ",",
  row.names = FALSE,
  quote = FALSE
)

write_xlsx(
  clock_component_clinical_delta_assoc_results,
  "figures/clock_component_clinical_delta_association_results.xlsx",
  col_names = TRUE,
  format_headers = TRUE
)
```
